---
title: "14-01, The two by two crosstabulation"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

```{r}
#| label: 14-01-dictionary
#| output: asis
#| message: false
#| warning: false
library(tidyverse)
load("../../../general/chopper.RData")
partition_yaml("titanic.yaml") |>
	cat(sep="\n")
```

## The crosstabulation of two binary variables, 1

```{}
             Good      Bad
			Outcome  Outcome
Placebo        a        b
Treated        c        d
```

-   a, number of placebo patients with good outcome
-   b, number of placebo patients with bad outcome
-   c, number of treated patients with good outcome
-   d, number of treated patients with bad outcome

:::notes
One of the most common tables you will see in Statistics is the 2 by 2 crosstabulation. This table shows the counts associated with the combination of the outcome and the treatment group. Is the risk of a bad outcome different between the two treatment groups?
:::

## The crosstabulation of two binary variables, 2

-   Note: rows could be
    -   Exposed/Unexposed
	-   Female/Male
	-   Old/Young
	-   Many other possibilities
	
:::notes
This table appears in many other contexts. You might want to compare people exposed to an environmental hazard to unexposed people. You might want to compare demographic groups: females to males, old to young, etc. There are many other possibilities.
:::

## Example: Titanic data

-   Crosstabulation

```{r}
#| label: 14-01-crosstabulaton-1
load("../data/titanic.RData")
table1 <- xtabs(~sex+survived, data=ti)
table1
```

:::notes
This is an example of a crosstabulation.The number in the upper left corner, 308, represents the number of female passengers who survived (did not die). This includes Kate Winslet. The number in the lower right corner, 709, represents then number of male passengers who did not survive. This includes, sad to say, Leonardo diCaprio.
:::

## Odds ratio

```{}
             Good      Bad
			Outcome  Outcome  Odds
Placebo        a        b     a/b
Treated        c        d     c/d
```
