---
title: "Analysis of Titanic dataset"
format: 
  html:
    embed-resources: true
---

This program reads data on survival of passengers on the Titanic. Find more
information in the [data dictionary][dd].

[dd]: https://github.com/pmean/data/blob/main/files/titanic.yaml

This code was written by Steve Simon on 2024-11-09 and is placed in the public
domain.

## Load the tidyverse library

```{r}
#| label: setup
#| message: false
#| warning: false
library(epitools)
library(tidyverse)
```

#### Comments on the code

For most of your programs, you should load the [tidyverse library][tid1]. The 
messages and warnings are suppressed.

[tid1]: https://www.tidyverse.org/

In previous programs, I put a label for each chunk inside the curly braces ({}).
It is recommended instead to put the label on a separate line inside the program
chunk. It is a bit more work to provide a unique label for each chunk, but it
helps quite a bit to isolate where to look when your code produces an error.

## Read the data and view a brief summary

```{r}
#| label: read
ti <- read_tsv(
  file="../data/titanic.txt",
  col_names=TRUE,
  col_types="ccncn",
  na="NA")
names(ti) <- tolower(names(ti))
glimpse(ti)
```

#### Comments on the code

Use read_tsv from the [readr package][rea1] to  read this file. Use 
col_names=TRUE because the column names are included as the first row of the 
file. The col_types="ccncn" specifies the first second and fourth columns as
strings and the third and fifth as numeric. There are missing values in this
dataset, designated by the letters "NA".

[rea1]: https://readr.tidyverse.org/

## Replace numeric codes for survived

```{r}
#| label: replace-numbers
ti$survived <-
	factor(
		ti$survived,
		level=1:0,
		labels=c("yes", "no"))
```

## Get counts of sex by survival

```{r}
#| label: counts
table1 <-xtabs(~sex+survived, data=ti)
table1
```

#### Comments on the code

The [table function][tab1] or the [xtabs function][xta1] creates a matrix with 
the number of observations in each combination of sex and survived. These 
values are placed in a single column. The [xtabs function][xta1] or the
[count function][cou1] provide slightly different approaches.

[cou1]: https://dplyr.tidyverse.org/reference/count.html
[tab1]: https://stat.ethz.ch/R-manual/R-devel/library/base/html/table.html
[xta1]: https://stat.ethz.ch/R-manual/R-devel/library/stats/html/xtabs.html

#### Interpretation of the output

There are 154 female passengers who died and 308 who survived. There are 709 
male passengers who died and 142 who survived.

## Chi-squared test, 1

```{r}
#| label: chi-squared-test-1
m1 <- chisq.test(table1, correct=FALSE)
m1
```

#### Comments on the code

The [chisq.test function][chi1] calculates a chi-square test of independence. It
takes input in a variety of forms. In this example, it uses a crosstabulation
computed by the xtabs command as input.

This function also will run a goodness-of-fit test, which is not discussed in 
this lecture.

[chi1]: https://stat.ethz.ch/R-manual/R-devel/library/stats/html/chisq.test.html

#### Interpretation of the output

The chi-squared statistic is much larger than the degrees of freedom and the
p-value is small. You should reject the null hypothesis and conclude that sex
and survival are related (not independent)

## Chi-squared test, 2

```{r}
#| label: chi-squared-test-2
m1$observed
```

## Chi-squared test, 3

```{r}
#| label: chi-squared-test-3
m1$expected
```

## Fisher's Exact test

```{r}
#| label: fishers-exact
fisher.test(table1)
```

#### Comments on the code

The [fisher.test function][fis1] calculates the Fisher's exact test, which is
helpful for small sample sizes. The 1,313 passengers on the Titanic do not
constitute a small sample size by any means. This test is just shown as an 
example of how to calculate this test. 

[fis1]: https://stat.ethz.ch/R-manual/R-devel/library/stats/html/fisher.test.html

#### Interpretation of the output

The p-value is small. You should reject the null hypothesis and conclude that 
sex and survival are related (not independent). The estimated odds ratio is 
9.97. The confidence interval for the odds ratio excludes the value of 1 
leading to the same conclusion. In fact, even after allowing for sampling error
that odds of survival are at least 7.6 times greater for women than for me.

## Odds ratio calculation

```{r}
#| label: odds-ratio
oddsratio(table1)
```

#### Comments on the code

The oddsratio function and riskratio function (see below) are part of the 
[epitools library][epi1]. It produces an odds ratio and confidence interval 
and p-values associated with the Fisher's Exact test and the Chi-squared test of
independence.

[epi1]: https://cran.r-project.org/web/packages/epitools/epitools.pdf

#### Interpretation of the output

We are 95% confident that the odds ratio of survival for women versus men is at
least 7.7 and possibly as large as 13, after accounting for sampling error. 
This interval excludes the value of 1, so you can conclude that the risk of 
death is much higher for men than for women. Equivalently you could conclude 
that the odds of survival are much higher for women than for men.

## Risk ratio calculation, 1

```{r}
#| label: risk-ratio-1
table1 |>
  proportions("sex")
```

#### Interpretation of the output

Before calculating the risk ratio, let's look at the row percentages one more 
time. The probability of survival is around 2/3 for women and about 1/6 for men.
This means that the risk ratio from a survival perspective is around 4 (2/3 
divided by 1/6). The probability of death is 1/3 for females and about 5/6 for
males. The risk ratio from a mortality perspective is 0.4 (1/3 divided by 5/6).

## Risk ratio calculation, 2

```{r}
#| label: risk-ratio-2
riskratio(table1)
```

#### Interpretation of the output

The risk ratio is comparing the probability of death between men and women. Men 
have 2.5 times higher probability of death compared to women. The confidence 
interval excludes the value of 1, indicating a statistically significant 
increase.

## Risk ratio calculation, 3

```{r}
#| label: risk-ratio-3
riskratio(table1, rev="columns")
```

#### Interpretation of the output

The risk ratio is comparing the probability of survival between men and women. 
Men has one-fourth the probability of survival compared to women. The confidence 
interval excludes the value of 1, indicating that men have a statistically 
significantly lower probability of survival compared to women. 

## Save data for later use

```{r save}
save(ti, file="../data/titanic.RData")
```

#### Comments on the code

It is usually a good idea to [save][sav1] your data in an RData file to make it
easier to retrieve this data later (with the [load function][loa1]).

[sav1]: https://stat.ethz.ch/R-manual/R-devel/library/base/html/save.html
[loa1]: https://stat.ethz.ch/R-manual/R-devel/library/base/html/save.html
