---
title: "13-02, Test of equality of two proportions"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

## The binomial distribution

-   You have n trials of an "experiment"
-   Two outcomes, "success" or "failure"
-   $\pi$ is probability of success
-   Each trial is independent

## Example, creating a family of three children, 1

-   You have 3 trials (pregnancies)
-   Two outcomes, girl (success) or boy (failure)
-   $\pi$ = 0.5 is probability of success
-   Each child is independent

## Example, creating a family of three children, 2

-   Eight possible outcomes
    -   X=0, BBB
    -   X=1, BBG, BGB, GBB
    -   X=2, BGG, GBG, GGB
    -   X=3, GGG
    
## Example, creating a family of three children, 3

```{r}
#| label: 13-02-setup
#| message: false
#| warning: false
library(tidyverse)
```

```{r}
#| label: 13-02-binomial-graph-1
x <- 0:3
y <- dbinom(0:3, 3, 0.5)
data.frame(x, y) |>
  ggplot() +
    aes(x, y) +
    geom_col(
      color="black",
      fill="white") +
  scale_y_continuous(breaks=(1:3)/8) +
  xlab("Binomial(n=3, pi=0.5)") +
  ylab("Probability") +
  ggtitle("Graph drawn by Steve Simon on 2024-11-12")
```

## Example, Binomial(n=10, pi=0.2)

```{r}
#| label: 13-02-binomial-graph-2
x <- 0:10
y <- dbinom(0:10, 10, 0.2)
data.frame(x, y) |>
  ggplot() +
    aes(x, y) +
    geom_col(
      color="black",
      fill="white") +
  xlab("Binomial(n=10, pi=0.2)") +
  ylab("Probability") +
  scale_x_continuous(
    breaks=0:10, 
    minor_breaks=NULL) +
  ggtitle("Graph drawn by Steve Simon on 2024-11-12")
```


## Two proportions model, 1

-   Scenario 1
    -   $X_1$ is a random binomial($n_1$, $pi_1$)
    -   $X_2$ is a random binomial($n_2$, $pi_2$)
    -   $X_1$ and $X_2$ are independent

## Two proportions model, 2

-   Scenario 2
    -   Sample $X_{11}, X_{12}, ..., X_{1n_1}$
    -   Sample $X_{21}, X_{22}, ..., X_{2n_2}$
    -   Only possible values are 0, 1
    -   $P[X_{1i}=1]=\pi_1$, $P[X_{2i}=1]=\pi_2$

## Two proportions model, 3

-   $H_0:\ \pi_1 - \pi_2 = 0$
-   $H_1:\ \pi_1 - \pi_2 \ne 0$

## Two proportions model, 4

-   $T = \frac{p_1-p_2}{se}$
    -   $se=\sqrt{\bar{p}(1-\bar{p}(\frac{1}{n_1}+\frac{1}{n_2})}$
    -   $\bar{p}=\frac{n_1p_1+n_2p_2}{n_1+n_2}
-   Accept $H_0$ if T is close to zero.
    -   $if z(\alpha/2) < T < z(1-\alpha/2)$

## Two proportions model, 5

-   p-value = $2P[Z >|T|]
-   Accept $H_0$ if p-value > $\alpha$

:::notes
You can also compute a p-value which is the probability of observing the test statistic, T, or a value more extreme.
:::

## Some variations of the test statistic, 1

-   Yates continuity correction
    -   $T = \frac{p_1-p_2 + c}{se}$
    -   Formula for c is messy
    -   Net effect is to pull T closer to zero
    -   Better to meet the normal approximation

## Some variations of the test statistic, 2

-   Chi-squared test
    -   $T = \Big(\frac{p_1-p_2}{se}\Big)^2$
    -   Accept $H_0$ if T < $\chi^2(1-\alpha, df=1)$
    -   p-value = $P[\chi^2(df=1) > T]$
    -   Does not allow for easy test of one-sided hypothesis
-   Chi-squared test with Yates continuity correction
    -   $T = \Big(\frac{p_1-p_2 + c}{se}\Big)^2$

## One-sided test, 1

-   $H_0:\ \pi_1 - \pi_2 = 0$
-   $H_1:\ \pi_1 - \pi_2 > 0$
    -   Accept $H_0$ if T < $z(1-\alpha)$
    -   p-value = $P[Z > T]$
    -   Accept $H_0$ if p-value > $\alpha$

## One-sided test, 2

-   $H_0:\ \pi_1 - \pi_2 = 0$
-   $H_1:\ \pi_1 - \pi_2 < 0$
    -   Accept $H_0$ if T > $z(\alpha)$
    -   p-value = $P[Z < T]$
    -   Accept $H_0$ if p-value > $\alpha$

## Two sample test of proportions with Titanic data, 1

```{r}
#| label: 13-02-titanic-1
load("../data/titanic.RData")
table1 <- xtabs(~sex+survived, data=ti)
table1
```

## Two sample test of proportions with Titanic data, 2

```{r}
#| label: 13-02-titanic-2
prop.table(table1, margin=1)
```

## Two sample test of proportions with Titanic data, 3

```{r}
#| label: 13-02-titanic-3
prop.test(
  table1, 
  correct=FALSE)
```
  
## Don't forget to do a sample size calculaton!!!
