---
title: "13-01, The two by two crosstabulation"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

```{r}
#| label: 13-01-dictionary
#| output: asis
#| message: false
#| warning: false

library(epitools)
library(tidyverse)
```

## The crosstabulation of two binary variables

```{}
            Variable2 
             0   1
Variable1 0 n00 n01
          1 n10 n11
```

:::notes
One of the most common tables you will see in Statistics is the 2 by 2 crosstabulation. This table shows the counts associated with the combination of the levels of the two binary variables. There are only four numbers in this table, but there are numerous statistics that you can use to summarize what's going on in this table.
:::

## Example: Titanic data

-   Crosstabulation

```{r}
#| label: 13-01-crosstabulaton-1
load("../data/titanic.RData")
table1 <- xtabs(~sex+survived, data=ti)
table1
```

:::notes
This is an example of a crosstabulation.The number in the upper left corner, 308, represents the number of female passengers who survived (did not die). This includes Kate Winslet. The number in the lower right corner, 709, represents then number of male passengers who did not survive. This includes, sad to say, Leonardo diCaprio.
:::

## Multiple ways to display, 1

:::: {.columns}
::: {.column width="50%"}
-   Crosstabulation

```{r}
#| label: 13-01-crosstabulation-2
table1
```

-   Swap rows

```{r}
#| label: 13-01-crosstabulation-3
table1[2:1, ]
```
:::
::: {.column width="50%"}
-   Swap columns

```{r}
#| label: 13-01-crosstabulation-4
table1[ , 2:1]
```

-   Swab both

```{r}
#| label: 13-01-crosstabulation-5
table1[2:1, 2:1]
```
:::
::::

:::notes
The cross tabulation changes when you swap the rows, the columns, or both.
:::

## Multiple ways to display, 2

:::: {.columns}
::: {.column width="50%"}
-   Transposed

```{r}
#| label: 13-01-crosstabulation-6
table2 <- t(table1)
table2
```

-   Transposed, swap rows

```{r}
#| label: 13-01-crosstabulation-7
table2[2:1, ]
```

:::
::: {.column width="50%"}
-   Transposed, swap columns

```{r}
#| label: 13-01-crosstabulation-8
table2[ , 2:1]
```

-   Transposed, swap both

```{r}
#| label: 13-01-crosstabulation-9
table2[2:1, 2:1]
```
:::
::::

:::notes
You can get four more tables by transposing the matrix. What was the rows becomes the columns and what was the columns becomes the rows.
:::

## Row and column percents

:::: {.columns}
::: {.column width="50%"}
-   Crosstabulation with row totals

```{r}
#| label: 13-01-add-row-margins
addmargins(table1, margin=2)
```

-  Row percents

```{r}
#| label: 13-01-row-percents
table1 |>
  prop.table(margin=1) |>
  addmargins(margin=2)
```

:::
::: {.column width="50%"}

-   Cross tabulation with column totals

```{r}
#| label: 13-01-add-column-margins
addmargins(table1, margin=1)
```

-  Column percents

```{r}
#| label: 13-01-column-percents
table1 |>
  prop.table(margin=2) |>
  addmargins(margin=1)
```
:::
::::

:::notes
The column percents are computed by dividing by the column totals. They add up to 100% within each column. The row percents are computed by dividing by the row totals. They add up to 100% within each row. The cell percents are computed by dividing by the overall total. They only add up to 100% when you sum across both the rows and the columns.
:::

## Cell percents

-   Cell totals

```{r}
#| label: 13-01-add-both-margins
table1 |>
  addmargins()
```

-   Cell percents

```{r}
#| label: 13-01-cell-percents
table1 |>
  prop.table() |>
  addmargins()
```


## My recommendation

-   Outcome variable is the columns
-   Intervention/exposure variable is the rows
-   Calculate row percentages

```{r}
#| label: 13-01-recommendation
table1 |>
  prop.table(margin=1)
```

:::notes
I have found that nine times out of ten, you want row percentages with the exposure/intervention variable as the rows and the outcome variable as the columns. This doesn't always work, but it is usually what I try first. It shows how much the chances of a good outcomes (or sometimes the chances of a bad outcome) change when you switch levels of the exposure/intervention. 

The orientation is also important. You want the percentages that are most interesting as close as possible. This is the proximity principle. The values within a column are nested above/below the other. The values within a row are farther apart.
:::

## Always round, 1

```{}
  0.6666667
- 0.1668625
  ---------
  ?????????
```

:::notes
Often in a cross tabulation, you will do some mental math. You might, for example, want to subtract survival probabilities.
::::

## Always round, 2

```{}
  0.67
- 0.17
  ----
  ????
```

:::notes
Notice how much easier the subtraction becomes when you round to two significant figures.
::::