---
title: "13-06, Chi-squared test of independence"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

```{r}
#| label: 13-06-setup
#| message: false
#| warning: false

library(tidyverse)
```

## Crosstabulation with row and column totals

-   Counts

```{r}
#| label: 13-06-crosstabulaton-1

load("../data/titanic.RData")
table1 <- xtabs(~sex+survived, data=ti)
table2 <- addmargins(table1)
table2
```

-   Cell percents

```{r}
#| label: 13-06-crosstabulaton-2

table3 <- table2/1313
table3

```

## Conditional probability

-   $P[A|B] = \frac{P[A \cap B]}{P[B]}$
    -   Read $P[A|B]$ as probability of A given B
    -   Read $P[A \cap B]$ as probability of A and B
    -   Note: $P[A|B] \ne P[B|A]$

## What does independence mean?

-   $P[A|B] = P[A]$
-   Equivalent definition of independence
    -   $P[A \cap B] = P[A] \times P[B]$

## Positive association

-   $P[A|B] > P[A]$
-   $P[A \cap B] > P[A] \times P[B]$
    -   Change direction for negative association

## Expected counts

```{}
             Good      Bad
            Outcome  Outcome     Total
Placebo        ?        ?       (a+b)/n
Treated        ?        ?       (c+d)/n
Total       (a+c)/n  (b+d)/n       1

where n=a+b+c+d

```

-   $E_{11}=n\times\frac{a+b}{n}\times\frac{a+c}{n}$
    -   $E_{12},\ E_{21},\ E_{22}$ are defined similarly

## Expected counts for Titanic data

```{}
        survived
sex            yes        no       Sum
  female                     0.3518660
  male                       0.6481340
  Sum    0.3427266 0.6572734 1.0000000
```

-   $E_{11}$ = 1313 $\times$ 0.3518660 $\times$ 0.3427266 = `r 1313*0.3518660*0.3427266`
-   $E_{12}$ = `r 1313*0.3518660*0.6572734`  
-   $E_{21}$ = `r 1313*0.6481340*0.3427266`
-   $E_{22}$ = `r 1313*0.6481340*0.6572734`  

##  Expected counts for Titanic data

-   Observed counts

```{r}
#| label: 13-06-observed

m1 <- chisq.test(table1, correct=FALSE)
m1$observed
```

-   Expected counts

```{r}
#| label: 13-06-expected

m1$expected
```

## Test statistic

-   $H_0:\ Independence$
-   $H_1:\ Dependence$
    -   $T=\Sigma_i \Sigma_j \frac{(O_{ij}-E_{ij})^2}{E_{ij}}$
    -   p-value = $P[T > \chi^2(df=1)]$
        -   Accept $H_0$ if T < $\chi^2(1-\alpha, df=1)$
        -   Accept $H_0$ if p-value > $\alpha$
    
## Example with Titanic data

-   $H_0:$ Mortality is independent of sex
-   $H_1:$ Mortality is related to sex
    
```{r}
#| label: 13-06-chi-squared
#| echo: true

m1 <- chisq.test(table1, correct=FALSE)
m1
```

:::notes
Since the test statistic is a lot larger than the degrees of freedom and since the p-value is small, reject the null hypothesis and conclude that there is a relationship between sex and survival.
:::

## Chi-squared test is an approximation

-   Reasonable if all expected counts > 5
-   Use Fisher's Exact test otherwise

## Fisher's exact test for the Titanic data

```{r}
#| label: 13-06-fisher
#| echo: true

m2 <- fisher.test(table1)
m2
```

:::notes
Although the expected counts are much larger than 5, here is the code for running Fisher's Exact test.
:::

