---
title: "Analysis of Titanic dataset"
format: 
  html:
    embed-resources: true
---

This program reads data on survival of passengers on the Titanic. Find more information in the [data dictionary][ref-simon-nodate].

[ref-simon-nodate]: https://github.com/pmean/data/blob/main/files/titanic.yaml

This code was written by Steve Simon on 2024-11-09 and is placed in the public domain.

## Load the tidyverse  and epitools library

```{r}
#| label: setup
#| message: false
#| warning: false

library(epitools)
library(tidyverse)
```

#### Comments on the code

For most of your programs, you should load the tidyverse library. The [epitools][ref-aragon-2020] library helps with calculations involving the odds ratio and relative risk.

[ref-aragon-2020]: https://CRAN.R-project.org/package=epitools 

## Read the data and view a brief summary

```{r}
#| label: read
ti <- read_tsv(
  file="../data/titanic.txt",
  col_names=TRUE,
  col_types="ccncn",
  na="NA")
names(ti) <- tolower(names(ti))
glimpse(ti)
```

#### Comments on the code

Use read_tsv from the [readr package][web02] to  read this file. Use col_names=TRUE because the column names are included as the first row of the file. The col_types="ccncn" specifies the first second and fourth columns as strings and the third and fifth as numeric. There are missing values in this dataset, designated by the letters "NA".

[web02]: https://readr.tidyverse.org/

# --- Part 1

## Replace numeric codes for survived

```{r}
#| label: replace-numbers
ti$survived <-
	factor(
		ti$survived,
		level=1:0,
		labels=c("yes", "no"))
```

#### Comments on the code

The [factor function][web03] places the levels of a categorical variable in a specific order and (optionally) attaches labels to each level. In this code, the number codes are reordered so that 1 appears first followed by 0. The labels "yes" and "no" are attached to these two codes.

[web03]: https://stat.ethz.ch/R-manual/R-devel/library/base/html/factor.html

## Get counts of sex by survival

```{r}
#| label: counts
table1 <-xtabs(~sex+survived, data=ti)
table1
```

#### Comments on the code

The [table function][web04] or the [xtabs function][web05] creates a matrix with the number of observations in each combination of sex and survived. The [count function][web06] provides an alternative where these values are placed in a single column. 

[web04]: https://stat.ethz.ch/R-manual/R-devel/library/base/html/table.html
[web05]: https://stat.ethz.ch/R-manual/R-devel/library/stats/html/xtabs.html
[web06]: https://dplyr.tidyverse.org/reference/count.html

#### Interpretation of the output

There are 154 female passengers who died and 308 who survived. There are 709 male passengers who died and 142 who survived.

## Get proportions for died/survived by sex

```{r}
#| label: proportions
table1 |>
  proportions("sex")
```

#### Comments on the code

The [proportions function][pro1] calculate proportions for a matrix, either by row (margin=1), by column (margin=2) or across all cells (margin=NULL). This code shows an alternative where you specify which variable to summarize across. The line proportions("sex") tells R to compute totals for each level of sex and divide by those totals. In this case, since sex represents the rows of table1, R produces row percentages. This syntax is easier to remember than margin=1 and margin=2.

[pro1]: https://stat.ethz.ch/R-manual/R-devel/library/base/html/proportions.html

#### Interpretation of the output

The proportion of women who died is 33%. The proportion of men who died is much higher at 83%.

## Bar chart

```{r}
#| label: bar
#| fig-height: 1.5
#| fig-width: 8

ti |>
  ggplot() +
  aes(y = sex, fill = survived) +
  geom_bar(position = "fill") +
  labs(
    x="Sex",
    y="Proportion",
    caption="Steve Simon, 2024-11-11, CC0")
```

#### Comments on the code

The [geom_bar][bar] function creates a bar chart. Setting one of the variables to y in the aes function will produce a horizontal bar chart. By default, geom_bar presents the counts for the cross-classification of "fill" and "y". The [position="fill"][pos] argument stacks these counts and standardizes them. This is effectively computing the conditional probability of the "fill" variable given the "y" variable.

[bar]: https://ggplot2.tidyverse.org/reference/geom_bar.html
[pos]: https://ggplot2.tidyverse.org/reference/position_stack.html

#### Interpretation of the output

I am not a big fan of bar charts, but they are quite popular. This bar chart shows a large disparity in the proportion who survived in each sex (the upper portion of the bar in peach). Equivalently, you can say that there is a large disparity in the proportion who died in each sex (the lower portion of the bar in green-blue). 

## Hypothesis test for two proportions

```{r}
#| label: equality-of-proportions

prop.test(table1, correct=FALSE)
```

#### Comments on the code

The [prop.test function][pro2] does not work on raw data. You need to get summary counts, either from the table function or the xtabs function. The option correct=FALSE informs R to not use the [Yates continuity correction][yat1].

[pro2]: https://search.r-project.org/R/refmans/stats/html/prop.test.html
[yat1]: https://www.statskingdom.com/121proportion_normal2.html

#### Interpretation of the output. 

The Chi-squared statistic is much larger than the degrees of freedom and the p-value is small. So you would reject the null hypothesis and conclude that there is a statistically significant difference in the mortality rates between men and women on the Titanic. The 95% confidence interval for the difference in proportions is -0.55 to -0.45. This interval excludes the value of zero and indicates that the mortality rate is at least 45% lower and possibly as much as
55% lower for men.

# --- Part 2

## Power calculation

```{r}
#| label: power
power.prop.test(p1=0.17, p2=0.67, power=0.9)
```

#### Comments on the code

The [power.prop.test][pow1] function will compute a sample size if you specify the two group probabilities, the significance level, and the power. It will also do other calculations, such as the minimum difference in probabilities that you could detect for a specific sample size and a specific power.

[pow1]: https://stat.ethz.ch/R-manual/R-devel/library/stats/html/power.prop.test.html

#### Interpretation of the output

Clearly, a sample size of 1,313 passengers provided more than enough power and precision, especially considering that the difference in survival rates was so large. Out of curiosity, what sample size would still be adequate for testing a shift of 50% from a survival rate of 17% to a survival rate of 67%? If you wanted power to be at least 90%, then a sample size of only 19 passengers of each sex would be needed.

# --- Part 3

## Chi-squared test

```{r}
#| label: chi-square-test
chisq.test(table1, correct=FALSE)
```

#### Comments on the code

The [chisq.test function][chi1] calculates a chi-square test of independence. It takes input in a variety of forms. In this example, it uses a crosstabulation computed by the xtabs command as input.

[chi1]: https://stat.ethz.ch/R-manual/R-devel/library/stats/html/chisq.test.html

#### Interpretation of the output

The chi-squared statistic is much larger than the degrees of freedom and the p-value is small. You should reject the null hypothesis and conclude that sex and survival are related (not independent)

## Fisher's Exact test

```{r}
#| label: fishers-exact
fisher.test(table1)
```

#### Comments on the code

The [fisher.test function][fis1] calculates the Fisher's exact test, which is helpful for small sample sizes. The 1,313 passengers on the Titanic do not constitute a small sample size by any means. This test is just shown as an example of how to calculate this test. 

[fis1]: https://stat.ethz.ch/R-manual/R-devel/library/stats/html/fisher.test.html

#### Interpretation of the output

The p-value is small. You should reject the null hypothesis and conclude that sex and survival are related (not independent). The estimated odds ratio is 9.97. The confidence interval for the odds ratio excludes the value of 1, leading to the same conclusion. In fact, even after allowing for sampling error, the odds of survival are at least 7.6 times greater for women than for men.

# --- Part 4

## Odds ratio calculation

```{r}
#| label: odds-ratio
oddsratio(table1)
```

#### Comments on the code

The oddsratio function and riskratio function (see below) are part of the epitools library epi1. It produces an odds ratio and confidence interval and p-values associated with the Fisher's Exact test and the Chi-squared test of independence.

#### Interpretation of the output

We are 95% confident that the odds ratio of survival for women versus men is at least 7.7 and possibly as large as 13, after accounting for sampling error.  This interval excludes the value of 1, so you can conclude that the risk of death is much higher for men than for women. Equivalently you could conclude that the odds of survival are much higher for women than for men.

## Risk ratio calculation, 1

```{r}
#| label: risk-ratio-1
table1 |>
  proportions("sex")
```

#### Interpretation of the output

Before calculating the risk ratio, let's look at the row percentages one more time. The probability of survival is around 2/3 for women and about 1/6 for men. This means that the risk ratio from a survival perspective is around 4 (2/3 divided by 1/6). The probability of death is 1/3 for females and about 5/6 for males. The risk ratio from a mortality perspective is 0.4 (1/3 divided by 5/6)

## Risk ratio calculation, 2

```{r}
#| label: risk-ratio-2
riskratio(table1)
```

#### Interpretation of the output

The risk ratio is comparing the probability of death between men and women. Men have 2.5 times higher probability of death compared to women. The confidence interval excludes the value of 1, indicating a statistically significant increase.

## Risk ratio calculation, 3

```{r}
#| label: risk-ratio-3
riskratio(table1, rev="columns")
```

#### Interpretation of the output

The risk ratio is comparing the probability of survival between men and women. Men has one-fourth the probability of survival compared to women. The confidence interval excludes the value of 1, indicating that men have a statistically significantly lower probability of survival compared to women. 

## Save data for later use

```{r save}
save(ti, file="../data/titanic.RData")
```

#### Comments on the code

It is usually a good idea to [save][save] your data in an RData file to make it easier to retrieve this data later (with the [load function][load]).

[save]: https://stat.ethz.ch/R-manual/R-devel/library/base/html/save.html
[load]: https://stat.ethz.ch/R-manual/R-devel/library/base/html/save.html
