---
title: "Analysis of Titanic dataset"
format: 
  html:
    embed-resources: true
---

This program reads data on survival of passengers on the Titanic. Find more
information in the [data dictionary][dd].

[dd]: https://github.com/pmean/data/blob/main/files/titanic.yaml

This code was written by Steve Simon on 2024-11-09 and is placed in the public
domain.

## Load the tidyverse library

```{r}
#| label: setup
#| message: false
#| warning: false
library(broom)
library(epitools)
library(tidyverse)
```

#### Comments on the code

For most of your programs, you should load the [tidyverse library][tid1]. The messages and warnings are suppressed.

[tid1]: https://www.tidyverse.org/

In previous programs, I put a label for each chunk inside the curly braces ({}).
It is recommended instead to put the label on a separate line inside the program
chunk. It is a bit more work to provide a unique label for each chunk, but it
helps quite a bit to isolate where to look when your code produces an error.

## Read the data and view a brief summary

```{r}
#| label: read
ti <- read_tsv(
  file="../data/titanic.txt",
  col_names=TRUE,
  col_types="ccncn",
  na="NA")
names(ti) <- tolower(names(ti))
glimpse(ti)
```

#### Comments on the code

Use read_tsv from the [readr package][rea1] to  read this file. Use 
col_names=TRUE because the column names are included as the first row of the 
file. The col_types="ccncn" specifies the first second and fourth columns as
strings and the third and fifth as numeric. There are missing values in this
dataset, designated by the letters "NA".

[rea1]: https://readr.tidyverse.org/

## Replace numeric codes for survived

```{r}
#| label: replace-numbers
ti$survived <-
	factor(
		ti$survived,
		level=1:0,
		labels=c("yes", "no"))
```

## Get counts of sex by survival

```{r}
#| label: counts
table1 <-xtabs(~sex+survived, data=ti)
table1
```

#### Comments on the code

The [table function][tab1] or the [xtabs function][xta1] creates a matrix with the number of observations in each combination of sex and survived. These values are placed in a single column. The [xtabs function][xta1] or the [count function][cou1] provide slightly different approaches.

[cou1]: https://dplyr.tidyverse.org/reference/count.html
[tab1]: https://stat.ethz.ch/R-manual/R-devel/library/base/html/table.html
[xta1]: https://stat.ethz.ch/R-manual/R-devel/library/stats/html/xtabs.html

#### Interpretation of the output

There are 154 female passengers who died and 308 who survived. There are 709 male passengers who died and 142 who survived.

## Get proportions for died/survived by sex

```{r}
#| label: proportions
table1 |>
  proportions("sex")
```

#### Comments on the code

The [proportions function][pro1] calculate proportions for a matrix, either by row (margin=1), by column (margin=2) or across all cells (margin=NULL).

[pro1]: https://stat.ethz.ch/R-manual/R-devel/library/base/html/proportions.html

#### Interpretation of the output

The proportion of women who died is 33%. The proportion of men who died is much higher at 83%.

## Bar chart

```{r}
#| label: bar
ti |>
  ggplot() +
    aes(x = sex, fill = survived) +
    geom_bar(position = "fill") +
	  xlab("Sex") +
	  ylab("Proportion") +
	  ggtitle("Graph drawn by Steve Simon on 2024-11-11")
```

## Hypothesis test for two proportions

```{r}
#| label: equality-of-proportions
prop.test(table1, correct=FALSE)
```

#### Comments on the code

The [prop.test function][pro2] does not work on raw data. You need to get summary counts, either from the table function or the xtabs function.

[pro2]: https://search.r-project.org/R/refmans/stats/html/prop.test.html

#### Interpretation of the output. 

The Chi-squared statistic is much larger than the degrees of freedom and the
p-value is small. So you would reject the null hypothesis and conclude that 
there is a statistically significant difference in the mortality rates between
men and women on the Titanic. The 95% confidence interval for the difference
in proportions is -0.55 to -0.45. This interval excludes the value of zero and
indicates that the mortality rate is at least 45% lower and possibly as much as
55% lower for men.

## Power calculation

```{r}
#| label: power
power.prop.test(p1=0.33, p2=0.83, power=0.9)
```

## Chi-squared test

```{r}
#| label: chi-square-test
chisq.test(table1)
```

## Fisher's Exact test

```{r}
#| label: fishers-exact
fisher.test(table1)
```

## Odds ratio calculation

```{r}
#| label: odds-ratio
oddsratio(table1)
```

## Risk ratio calculation

```{r}
#| label: risk-ratio
riskratio(table1)
```

## Save data for later use

```{r save}
save(ti, file="../data/titanic.RData")
```

#### Comments on the code

It is usually a good idea to save your data in an RData file to make it easier to retrieve this data later (with the load function).

#### Interpretation of the output

