---
title: "04-05, R code for bar plots"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

```{r data0b}
library(glue)
library(readr)
library(stringr)
library(yaml)
f1 <- "https://raw.githubusercontent.com/pmean"
f2 <- "data/main/files/gardasil.yaml"
dd <- read_lines(glue("{f1}/{f2}"))
# if I am not online use the line below
# dd <- read_lines("../../../../data/files/gardasil.yaml")
h <- str_which(str_trim(dd), "^$")
chunk_beg <- c(1, h+1)
chunk_end <- c(h, length(dd))
```

## Data dictionary for gardasil, 1

```{r data1b}
cat(dd[chunk_beg[1]:chunk_end[1]], sep="\n")
```

```{r data2b}
cat(dd[chunk_beg[2]:chunk_end[2]], sep="\n")
```

```{r data3b}
cat(dd[chunk_beg[3]:chunk_end[3]], sep="\n")
```

```{r data4b}
cat(dd[chunk_beg[4]:chunk_end[4]], sep="\n")
```

:::notes
Here is a dataset you will need for your programming assignment. It is a study of pulmonary function in children.
:::

## Data dictionary for gardasil, 2

```{r data5b}
cat(dd[chunk_beg[5]:chunk_end[5]], sep="\n")
```

```{r data6b}
cat(dd[chunk_beg[6]:chunk_end[6]], sep="\n")
```

```{r data7b}
cat(dd[chunk_beg[7]:chunk_end[7]], sep="\n")
```

```{r data8b}
cat(dd[chunk_beg[8]:chunk_end[8]], sep="\n")
```

```{r data9b}
cat(dd[chunk_beg[9]:chunk_end[9]], sep="\n")
```

:::notes

:::

## Data dictionary for gardasil, 3

```{r data10b}
cat(dd[chunk_beg[10]:chunk_end[10]], sep="\n")
```

```{r data11b}
cat(dd[chunk_beg[11]:chunk_end[11]], sep="\n")
```

:::notes
T.
:::

## Data dictionary for gardasil, 4

```{r data12b}
cat(dd[chunk_beg[12]:chunk_end[12]], sep="\n")
```

```{r data13b}
cat(dd[chunk_beg[13]:chunk_end[13]], sep="\n")
```

```{r data14b}
cat(dd[chunk_beg[14]:chunk_end[14]], sep="\n")
```

:::notes
T.
:::

## Data dictionary for gardasil, 5

```{r data15b}
cat(dd[chunk_beg[15]:chunk_end[15]], sep="\n")
```

```{r data16b}
cat(dd[chunk_beg[16]:chunk_end[16]], sep="\n")
```

```{r data17b}
cat(dd[chunk_beg[17]:chunk_end[17]], sep="\n")
```

```{r data18b}
cat(dd[chunk_beg[18]:chunk_end[18]], sep="\n")
```

## Data dictionary for gardasil, 6

```{r data19b}
cat(dd[chunk_beg[19]:chunk_end[19]], sep="\n")
```

```{r data20b}
cat(dd[chunk_beg[20]:chunk_end[20]], sep="\n")
```

```{r data21b}
cat(dd[chunk_beg[21]:chunk_end[21]], sep="\n")
```

```{r data22b}
cat(dd[chunk_beg[22]:chunk_end[22]], sep="\n")
```

## simon-5501-04-gardasil.qmd, 1

```{r write0b}
fn <- "simon-5501-04-gardasil.qmd"
template <- read_lines(file=fn)
h <- str_which(template, "##")
chunk_beg <- c(1, h)
chunk_end <- c(h-1, length(template))
```

```{r write1b}
cat(template[chunk_beg[1]:chunk_end[1]], sep="\n")
```

:::notes
The first few lines are the documentation header.
:::

## simon-5501-04-gardasil.qmd, 2

```{r write2b}
cat(template[chunk_beg[2]:chunk_end[2]], sep="\n")
```

:::notes
H.
:::

## simon-5501-04-gardasil.qmd, 3

```{r write3b}
cat(template[chunk_beg[3]:chunk_end[3]], sep="\n")
```

:::notes
L.
:::

## simon-5501-04-gardasil.qmd, 4

```{r write4b}
cat(template[chunk_beg[4]:chunk_end[4]], sep="\n")
```

:::notes
Use the read_tsv function when your data uses tab delimiters.
:::

## simon-5501-04-gardasil.qmd, 5

```{r write5b}
cat(template[chunk_beg[5]:chunk_end[5]], sep="\n")
```
:::notes
T. 
:::

## simon-5501-04-gardasil.qmd, 6

```{r write6b}
cat(template[chunk_beg[6]:chunk_end[6]], sep="\n")
```

:::notes
T.
:::

## simon-5501-04-gardasil.qmd, 7

```{r write7b}
cat(template[chunk_beg[7]:chunk_end[7]], sep="\n")
```

:::notes
T.
:::

## simon-5501-04-gardasil.qmd, 8

```{r write8b}
cat(template[chunk_beg[8]:chunk_end[8]], sep="\n")
```

:::notes
T.
:::

## simon-5501-04-gardasil.qmd, 9

```{r write9b}
cat(template[chunk_beg[9]:chunk_end[9]], sep="\n")
```

:::notes
T.
:::
