---
title: "05-02, R code linear regression"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

```{r data0}
library(glue)
library(readr)
library(stringr)
library(yaml)
f1 <- "https://raw.githubusercontent.com/pmean"
f2 <- "data/main/files/breast-feeding-preterm.yaml"
dd <- read_lines(glue("{f1}/{f2}"))
# if I am not online use the line below
# dd <- read_lines("../../../../data/files/breast-feeding-preterm.yaml")
h <- str_which(str_trim(dd), "^$")
chunk_beg <- c(1, h+1)
chunk_end <- c(h, length(dd))
```

## Data dictionary for breast feeding study, 1

```{r data1}
cat(dd[chunk_beg[1]:chunk_end[2]], sep="\n")
```

:::notes
Here is a dataset you will need for your programming assignment. It is a study of pulmonary function in children.
:::

## Data dictionary for breast feeding study, 2

```{r data2}
cat(dd[chunk_beg[3]:chunk_end[3]], sep="\n")
```

:::notes
Here is information about the copyright.
:::

## Data dictionary for breast feeding study, 3

```{r data3}
cat(dd[chunk_beg[4]:chunk_end[7]], sep="\n")
```

:::notes
This is a comma delimited file with a missing value code of -1.
:::

## Data dictionary for breast feeding study, 4

```{r data4}
cat(dd[chunk_beg[8]:chunk_end[9]], sep="\n")
```

:::notes
Here are the first few variables for the data.
:::

## Data dictionary for breast feeding study, 5

```{r data5}
cat(dd[chunk_beg[10]:chunk_end[11]], sep="\n")
```

:::notes
Here are some additional variables.
:::

## Data dictionary for breast feeding study, 6

```{r data6}
cat(dd[chunk_beg[12]:chunk_end[13]], sep="\n")
```

:::notes
Here are some additional variables. We won't show the information for the remaining variables in the data.
:::

## simon-5501-05-bf.qmd, 1

```{r write0}
fn <- "simon-5501-05-bf.qmd"
template <- read_lines(file=fn)
h <- str_which(template, "##")
chunk_beg <- c(1, h)
chunk_end <- c(h-1, length(template))
```

```{r write1}
cat(template[chunk_beg[1]:chunk_end[1]], sep="\n")
```

:::notes
The first few lines are the documentation header.
:::

## simon-5501-05-bf.qmd, 2

```{r write2}
cat(template[chunk_beg[2]:chunk_end[2]], sep="\n")
```

:::notes
H.
:::

## simon-5501-05-bf.qmd, 3

```{r write3}
cat(template[chunk_beg[3]:chunk_end[3]], sep="\n")
```

:::notes
L.
:::

## simon-5501-05-bf.qmd, 4

```{r write4}
cat(template[chunk_beg[4]:chunk_end[4]], sep="\n")
```

:::notes
Use the read_tsv function when your data uses tab delimiters.
:::

## simon-5501-05-bf.qmd, 5

```{r write5}
cat(template[chunk_beg[5]:chunk_end[5]], sep="\n")
```
:::notes
T. 
:::

## simon-5501-05-bf.qmd, 6

```{r write6}
cat(template[chunk_beg[6]:chunk_end[6]], sep="\n")
```

:::notes
T.
:::

## simon-5501-05-bf.qmd, 7

```{r write7}
cat(template[chunk_beg[7]:chunk_end[7]], sep="\n")
```

:::notes
T.
:::

## simon-5501-05-bf.qmd, 8

```{r write8}
cat(template[chunk_beg[8]:chunk_end[8]], sep="\n")
```

:::notes
T.
:::

