---
title: "simon-5501-05-demo"
format: 
  html:
    embed-resources: true
---

# --- Start of part 1

## Documentation

This program reads data and fits various linear regression models on a breast feeding study in pre-term infants.  Find more information in the [data dictionary][dd]. This program was written by Steve Simon on 2024-09-11 and is placed in the public domain.

[dd]: https://raw.githubusercontent.com/pmean/datasets/master/breast-feeding-preterm.yaml
If you have more than one dataset, possibly because you changed something, then it is a good idea to list the datasets and explain what changes were made.

-   bf: read from breast-feeding-preterm.csv
-   bf1: converted -1 to missing for age_stop

```{r}
#| label: setup
#| message: false
#| warning: false

library(broom)
library(tidyverse)
```

### Comments on the code

The [broom library][ref-broom] provides simple format for output from various statistical functions in R. We will use it for the lm function, but it does a nice job with quite a few other functions as well.

[ref-broom]: https://broom.tidymodels.org/

```{r}
#| label: read

bf <- read_csv(
  file="../data/breast-feeding-preterm.csv",
  col_types="cnnnnnnnncnnnnnnnnnnnnnnnnnn",
  col_names=TRUE)
glimpse(bf)
```

### Comments on the code

Use the read_csv function to read the data. With a large number of variables, you may choose to leave the col_types out. R will usually figure out which variables are numeric and which are strings.

## Data management

```{r}
#| label: convert-missing

bf |>
  mutate(age_stop=na_if(age_stop, -1)) -> bf1
```

### Comments on the code

The [na_if function][ref-na-if] sets a value to missing (NA) if the first argument equals the second argument. If they are not equal, it keeps the first argument.

There are several other variables that use -1 for missing. You should extend this section if you plan on using any of those variables.

[ref-na-if]: https://dplyr.tidyverse.org/reference/na_if.html

## Basic descriptive statistics

```{r}
#| label: mom-age

bf1 |>
	summarize(
		mean_mom_age=mean(mom_age, na.rm=TRUE),
		sd_mom_age=sd(mom_age, na.rm=TRUE),
		min_mom_age=min(mom_age, na.rm=TRUE),
		max_mom_age=max(mom_age, na.rm=TRUE),
		n_missing=sum(is.na(mom_age))) |>
	data.frame()
```

### Interpretation of the output

This is a reasonable distribution of ages. If you saw mothers much younger than 16 years or much older than 44 years, that might raise some concerns about the data.

```{r}
#| label: dc-age

bf1 |>
	summarize(
		mean_age_stop=mean(age_stop, na.rm=TRUE),
		sd_age_stop=sd(age_stop, na.rm=TRUE),
		min_age_stop=min(age_stop, na.rm=TRUE),
		max_age_stop=max(age_stop, na.rm=TRUE),
		n_missing=sum(is.na(age_stop))) |>
	data.frame()
```

### Interpretation of the output

The maximum value, 34 weeks, was a bit of concern for me, because the study was a six month study, which would imply the largest value would be 24 or 26. But I was told that breast feeding duration included time in the hospital, which could easily be as long as 8 or 10 weeks for a pre-term infant.

```{r}
#| label: plot
#| message: false

bf1 |>
  filter(!is.na(age_stop)) |>
	ggplot(aes(mom_age, age_stop)) +
	  geom_point() +
    geom_smooth(method="lm", se=FALSE) +
	  labs(
	  	x="Mother's age (years)",
	  	y="When breast feeding ended (weeks)",
	  	caption="Simon, 2024-09-17")
```

### Interpretation of the output

There is a weak relationship between mother's age and age when she stopped breast feeding.

## Data analysis

```{r}
#| label: linear-regression

m1 <- lm(
  formula=age_stop~mom_age, 
  data=bf1,
  na.action=na.omit)
m1
```

### Comments on the code

The [lm function][ref-lm], part of base R, computes a linear regression model. The first argument is a formula. A [formula][ref-formula] is an R object that has a tilde (~) with variables either on the left, on the right, or on both sides of the tilde. This has various interpretations in R, but the simplest interpretation is the one that is used with lm. For the lm function, the left side of the tilde represents a continuous outcome. You can only have one outcome.

The na.action argument tells R how to handle missing values.

The right side of the tilde lists an independent variable. The independent variable can be either continuous or categorical. There is a way to specify multiple independent variables, interactions, and other complex analyses. In this module, just list a single independent variable.

The data argument tells R what data frame or tibble has the variables that you specified in your formula.

The results of the lm function are stored in a [list][ref-list]. This list is very complex if you try to peek at all the details. Just printing the list gives an abbreviated summary of the linear regression. If you are interested in exploring more about how the list that lm creates, read my web page, [Creatures in the R zoo][ref-simon-2024].

[ref-lm]: https://stat.ethz.ch/R-manual/R-devel/library/stats/html/lm.html
[ref-formula]: https://stat.ethz.ch/R-manual/R-devel/library/stats/html/formula.html
[ref-list]: https://stat.ethz.ch/R-manual/R-devel/library/base/html/list.html
[ref-simon-2024]: http://pmean.com/posts/r-creatures/

### Interpretation of the output

The estimated average duration of breast feeding increases by 0.39 weeks for each increase of one year in the mother's age. The estimated average duration of breast feeding is 5.9 weeks for a mother of age zero. This is an extrapolation well beyond the range of the data. 

# --- Start of part 2

## Additional data analysis

```{r}
#| label: anova

anova(m1)
```

### Interpretation of the output

The F-ratio is large and the p-value is small, so you would reject the null hypothesis and conclude that there is a linear relationship between mother's age and duration of breast feeding.

```{r}
#| label: r-squared

glance(m1)$r.squared
```

### Comments on the code

The glance function is part of the broom library. It provides a variety of single number summaries of the statistical model. The only summary we are interested in for now is r.squared.

### Interpretation of the output

Although there is a statistically significant relationship between mother's age and duration of breast feeding, as shown above, this relationship is very weak.

# --- Start of part 3

## Additional data analysis

```{r}
#| label: ci-age-stop

confint(m1)
```

### Interpretation of the output

The confidence interval includes only positive values, so we are 95% confident that the duration of breast feeding increases as the mother's age increases. The slope could be as small as  0.067 weeks per year of mother's age or as large as 0.71 weeks per year of mother's age. This is a very wide interval indicating a large degree of uncertainty about the true value of the slope parameter.

```{r}
#| label: t-test-for-age-stop

tidy(m1)
```

### Comments on the code

The tidy function is part of the broom library and chooses a simple and flexible display of the intercept, slope, their standard efforts, t-statistics, and p-values.

### Interpretation of the output

The T statistic is testing the slope parameter is large and the p-value is small, both indicating that you should reject the null hypothesis and conclude that there is a positive relationship between mother's age and duration of breast feeding.
