---
title: "simon-5501-05-image-maker"
format: 
  html:
    embed-resources: true
editor: source
---

# --- Preliminaries

## Documentation

This program is for internal use, but you are welcome to look things over.

This program Create and store datasets and images for module05 slides. It was written by Steve Simon on 2026-09-15 and is placed in the public domain.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(broom)
library(tidyverse)
```

# --- Data entry

## Read breast-feeding-preterm data

```{r}
#| label: read-bf
#| message: false
#| warning: false

bf <- read_csv(
  file="../data/breast-feeding-preterm.csv",
  col_names=TRUE)
```

## Convert -1 to NA

```{r}
#| label: missing

bf |>
  mutate(age_stop=na_if(age_stop, -1)) -> bf1

```

## Create simple dataset

```{r read-simple}
fn <- "../data/hypothetical-correlation.csv"
read_csv(
  file=fn,
  col_types="nn") -> simple
```

# --- Create graphic images for _05-01

## Create draw_line function

```{r draw-line}
draw_line <- function(a, b, plot_number, d=simple, pts=simple$y) {
  # sse <- sum(d$res^2)
  d %>%
    mutate(y=pts) |>
    mutate(pred=a+b*x) |>
    mutate(res=pts-pred) -> d
  d |>
    ggplot() +
      aes(x=x, y=pts, xend=x, yend=pred) +
      expand_limits(x=c(0, 20), y=c(-5, 40)) +
      geom_point() +
      geom_segment() +
      annotate(
        "label",
        x=d$x,
        y=-3,
        label=d$res) +
      annotate(
        "segment",
        x=0,
        y=a,
        xend=20,
        yend=a+b*20) +
      labs(
        x="x (hypothetical data)",
        y="y (hypothetical data)",
        caption= "Simon 2026-09-22") -> least_squares_plot
  
  ggsave(paste0("../images/plot", plot_number, ".png"))
  
  least_squares_plot
}
```

## Create regression model and coefficients

```{r}
#| label: m1

m1 <- lm(age_stop~mom_age, data=bf1)
b <- coef(m1)
```

## Draw plots

```{r}
#| label: plot01
#| message: false

bf1 |>
  ggplot() +
  aes(mom_age, age_stop) +
  geom_point() +
	labs(
	 	x="Mother's age (years)",
	 	y="When breast feeding ended (weeks)",
	 	caption="Simon, 2024-09-17") -> plot01
plot01  
ggsave("../images/plot01.png")
```

```{r}
#| label: plot02
#| message: false

plot01 +
  geom_smooth(
    method="lm",
    se=FALSE) -> plot02
plot02
ggsave("../images/plot02.png")
```

```{r}
#| label: plot03
#| message: false

plot02 +
  expand_limits(x=0) +
  geom_segment(
    x=0, 
    y=b[1],
    xend=16,
    yend=b[1]+16*b[2],
    linetype="dotted") -> plot03
plot03
ggsave("../images/plot03.png")
```

# --- Draw plots for _05-03

## Bad fit, a=20, b=1

```{r}
draw_line(20, 1, "04")
```

## Better fit, a=10, b=1

```{r}
draw_line(10, 1, "05")
```

## Best fit, a=30, b=-1

```{r}
draw_line(30, -1, "06")
```

## --- Draw plots for _05-04

## Sum of squares regression

```{r}
yhat <- 30-simple$x
draw_line(20, 0, "07", pts=yhat)
```

## Sum of squares error

```{r}
draw_line(30, -1, "08")
```

## Sum of squares total / corrected total

```{r}
draw_line(20, 0, "09")
```

## Sum of squares total (uncorrected)

```{r}
draw_line(0, 0, "10")
```

## --- Florida 2000 election results

```{r}
#| label: read-florida

candidate_names <- c(
  "County",
  "Bush_count",
  "Bush_percent",
  "Gore_count",
  "Gore_percent",
  "Nader_count",
  "Nader_percent",
  "Buchanan_count",
  "Buchanan_percent",
  "Other_count",
  "Other_percent",
  "Margin_count",
  "Margin_percent",
  "Total_count"
)

fl <- read_tsv(
  "../data/florida-votes-2000.txt",
  col_names = candidate_names,
  col_types="cnnnnnnnnnnnn"
)

glimpse(fl)
```

```{r}
fl |>
  mutate(Bush_count_thousands = Bush_count/1000) |>
  filter(County != "Totals") -> fl1

with(fl1, cor(Buchanan_count, Bush_count))
with(fl1, cor(Buchanan_count, Total_count))
```

```{r}
m2 <- lm(Buchanan_count ~ Bush_count_thousands, data=fl1)
m2
augment(m2) |>
  filter(fl1$County=="Palm Beach")
```

```{r}
fl1 |>
  ggplot() +
  aes(x=Bush_count_thousands, y=Buchanan_count) +
  geom_point(shape=1) +
  labs(
    x="Votes for Bush (thousands)",
    y="Votes for Buchanan",
    caption="Simon 2026-09-23") -> plot11
plot11
ggsave("../images/plot11.png")
```

## Save everything

```{r save}
save(
  bf,
  bf1,
  m1, 
  b, 
  simple,
  fl,
  fl1,
  m2,
  file="../data/module05.RData")
```