---
title: "Analysis of breast feeding study"
author: "Steve Simon"
format: 
  html:
    embed-resources: true
date: 2024-09-11
---

This program reads data and fits various linear regression models on a breast feeding study in pre-term infants.  Find more information in the [data dictionary][dd].

[dd]: https://raw.githubusercontent.com/pmean/datasets/master/breast-feeding-preterm.yaml

This code is placed in the public domain.

## Load the tidyverse library

For most of your programs, you should load the tidyverse library. The messages and warnings are suppressed.

```{r setup}
#| message: false
#| warning: false
library(tidyverse)
```

## Read the data and view a brief summary

Use the read_csv function to read the data. With a large number of variables, you may choose to leave the col_types out.R will usually figure out which variables are numeric and which are strings.

Replace all the numeric codes of -1 with the missing value code (NA).

```{r read}
bf <- read_csv(
  file="../data/breast-feeding-preterm.csv",
  col_names=TRUE)
bf[bf==-1] <- NA
glimpse(bf)
```



## Calculate statistics for gest_age

```{r gest-age}
nmiss <- function(x) {sum(is.na(x))}

bf |>
	summarize_if(is.numeric, nmiss)

bf |>
	summarize(
		mean_age_stop=mean(age_stop, na.rm=TRUE),
		sd_age_stop=sd(age_stop, na.rm=TRUE),
		n_missing=sum(is.na(age_stop)))


bf |>
	summarize(
		mean_ga=mean(gest_age),
		sd_ga=sd(gest_age),
		n_missing=sum(is.na(gest_age)))
```