---
title: "Hypothetical litter weights"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

```{r}
#| label: 12-02-setup
#| message: false
#| warning: false

library(glue)
library(tidyverse)
library(yaml)

load("../data/module12.RData")
```

## Description of litter weights data, 1 of 3

```{r}
#| label: 12-02-ddf-1

f1 <- "https://raw.githubusercontent.com"
f2 <- "pmean/data/refs/heads/main/files"
f3 <- "litter-weights.yaml"

dd <- read_yaml(glue("{f1}/{f2}/{f3}"))

cat(as.yaml(dd["data_dictionary"]))
cat(as.yaml(dd["description"]))
cat(as.yaml(dd["source"]))
```



## Description of litter weights data, 2 of 3

```{r}
#| label: 12-02-ddf-2

cat(as.yaml(dd$vars[1:3]))
```

## Description of litter weights data, 3 of 3

```{r}
#| label: 12-02-ddf-3

cat(as.yaml(dd$vars[4:5]))
```

## Layout of full dataset

![](../images/litter-graph-1 "Plot showing 6 rats per litter, 3 male, 3 female,across 20 litters, 10 in group 1 and 10 in group 2")

## Layout for two sample t-test

![](../images/litter-graph-2 "Plot showing subset of full data, 1 random male per litter in the 10 group 1 litters and in the 10 group 2 litters")

## Layout for the paired t-test

![](../images/litter-graph-3 "Plot showing subset of full data, 1 random male and 1 random female per litter in the 10 group 1 litters")

## Why you need to accommodate non-independence

-   Changes in power and precision
-   Combine multiple tests into single framework
-   Handle missing data 
-   Allows more sophisticated analyses

## How does power/precision change?

-   Within cluster comparisons standard error
    -   $\sqrt{\frac{\sigma^2_{within}}{an}}$
    -   $\sqrt{\frac{\sigma^2_{total}}{an}(1-\rho_{ICC})}$
-   Between cluster comparisons standard error
    -   $\sqrt{\frac{\sigma^2_{within}}{an}+\frac{\sigma^2_{between}}{a}}$
    -   $\sqrt{\frac{\sigma^2_{total}}{an}(1+(n-1) \rho_{ICC})}$

::: notes

These calculations only work for the simple case where you have the exact same number of observations in each cluster. The formulas illustrate an important issue that also holds when the number of observations does vary from cluster to cluster. The actual formulas, though, become quite tedious in this setting.

:::
    
## Pseudo-replication

-   Hypothetical study of water contamination
    -   Drill 5 wells
    -   Sample 4 times at each well
    -   Treat as 20 independent observations
-   Standard error is not
    -   $\sqrt{\frac{\sigma^2_{total}}{an}}$
-   Standard error is
    -   $\sqrt{\frac{\sigma^2_{total}}{an}(1+(n-1) \rho_{ICC})}$

::: notes

Stuart H. Hurlbert. (1984), Pseudoreplication and the Design of Ecological Field Experiments. Ecological Monographs, 54: 187-211. https://doi.org/10.2307/1942661

Festing, M.F.W. The “completely randomised” and the “randomised block” are the only experimental designs suitable for widespread use in pre-clinical research. Sci Rep 10, 17577 (2020). https://doi.org/10.1038/s41598-020-74538-3

:::

## Design considerations

-   Within cluster comparisons always have greater precision and power
    -   Not always possible
-   Best to increase number of clusters
    -   Not always economical

## Degrees of freedom

-   df = N - k (for a simpler design)
    -   k = number of estimated parameters
    -   What is N?
        -   N=5? (Number of wells)
        -   N=20? (Number of water samples)
