---
title: "simon-5502-12-solution"
format: docx
editor: source
execute: 
  error: false
---

## File details

This program was written by Steve Simon on 2025-04-29 and is placed in the public domain. You can use this program any way you please.

```{r}
#| message: false
#| warning: false
#| label: setup

library(lme4)
library(tidyverse)

R.version.string
Sys.Date()
```

## Read fasting-turtles

```{r}
#| label: read

turtles <- read_tsv(
  file="../data/fasting-turtles.txt",
  col_names=TRUE,
  col_types="ncnnn")

glimpse(turtles)
```

## Question 1

You will not be graded on this, but get in the habit of drawing a few graphs and computing a few statistics that will help you better understand the dataset you are working with. Only examine variables that will be part of the further analyses.

It would also be fine to wait until after restructuring to calculate descriptive statistics.

```{r}
#| label: descriptives

turtles |>
	summarize(
		baseline_mean=mean(Fed),
		baseline_sd=sd(Fed),
		baseline_min=min(Fed),
		baseline_max=max(Fed))

turtles |>
	summarize(
		time_10_mean=mean(Fasted10),
		time_10_sd=sd(Fasted10),
		time_10_min=min(Fasted10),
		time_10_max=max(Fasted10))

turtles |>
	summarize(
		time_20_mean=mean(Fasted20),
		time_20_sd=sd(Fasted20),
		time_20_min=min(Fasted20),
		time_20_max=max(Fasted20))

```

## Question 2

Restructure the data so Fed, Fasted10, and Fasted20 are in the same column. Create a new time variable from Fed, Fasted10, and Fasted20 that has values of 0, 10, and 20 respectively. The restructured dataset should have 24 rows. Show a glimpse of the new dataset.

## Restructure into long format

```{r}
#| label: long

turtles |>
  pivot_longer(
    cols=Fed:Fasted20,
    names_to="measure",
    values_to="plasma_protein") |>
  mutate(time=case_when(
  	measure=="Fed" ~ 0,
  	measure=="Fasted10" ~ 10,
  	measure=="Fasted20" ~ 20)) -> turtles_1

glimpse(turtles_1)
```

## Question 3

Draw a line graph, showing the change in plasma protein over time for each turtle. Ignore sex in this graph. Do the data show a consistent pattern (e.g., patients with large values at baseline tend to have large values after 10 and 20 days of fasting (relative to turtles that have small values at baseline)?

```{r}
#| label: plot-1

turtles_1 |>
  ggplot() +
  aes(
    x = time, 
    y = plasma_protein, 
    group = Subject) +
  geom_line() +
	labs(
		x="Time (days)",
		y="Plasma protein (mg/ml)",
		title="Line graph of individual turtles over time",
		caption="Simon, 2025-04-29")
```

## Question 4

Fit a random intercepts model with plasma protein as the dependent variable and time as the independent variable. Is there a statistically significant decline in plasma protein over time?

```{r}
#| label: mixed

m1 <- lmer(
  plasma_protein ~ time + (1 | Subject),
  data=turtles_1)

summary(m1)
```

## Question 5

Calculate the intraclass correlation coefficient. Interpret this value.

```{r}
#| label: icc

icc <- 10.7945/(10.7945+0.5811)
icc
```
