---
title: "07-01, Multiple linear regression"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

## Model

-   $Y_i=\beta_0+\beta_1 X_{1i}+\beta_2 X_{2i}+\epsilon_i$
-   Least squares estimates: $b_0,\ b_1,\ b_2$

::: notes
Add note.
:::

## Interpretations

-   $b_0$ is the estimated average value of Y when X1 and X2 both equal zero.
-   $b_1$ is the estimated average change in Y
    -   when $X_1$ increases by one unit, and
    -   $X_2$ is held constant
-   $b_2$ is the estimated average change in Y
    -   when $X_2$ increases by one unit, and
    -   $X_1$ is held constant
    
## Unadjusted relationship between height and FEV

```{r}
library(broom)
library(tidyverse)
load("../data/module07.RData")
lm(fev~height, data=pulmonary) %>%
  tidy %>%
  pull(estimate) -> bu
ggplot(pulmonary, aes(height, fev, label=age)) +
  geom_text() +
  geom_abline(slope=bu[2], intercept=bu[1])
```

## Relationship between height and FEV controlling at Age=3

```{r}
lm(fev~height+age, data=pulmonary) %>%
  tidy %>%
  pull(estimate) -> b
control_age <- function(a) {
  co=ifelse(pulmonary$age==a, "black", "gray")
  o <- order(co)
  pulmonary[o, ] %>%
    ggplot(aes(height, fev, label=age)) +
      geom_text(color=co[o]) +
      geom_abline(slope=b[2], intercept=b[1]+b[3]*a) +
      geom_abline(slope=bu[2], intercept=bu[1], color="gray")
}
a <- 3
```

## Relationship between height and FEV controlling at Age=`r a`

```{r}
control_age(a)
a <- a+1
```

## Relationship between height and FEV controlling at Age=`r a`

```{r}
control_age(a)
a <- a+1
```

## Relationship between height and FEV controlling at Age=`r a`

```{r}
control_age(a)
a <- a+1
```

## Relationship between height and FEV controlling at Age=`r a`

```{r}
control_age(a)
a <- a+1
```

## Relationship between height and FEV controlling at Age=`r a`

```{r}
control_age(a)
a <- a+1
```

## Relationship between height and FEV controlling at Age=`r a`

```{r}
control_age(a)
a <- a+1
```

## Relationship between height and FEV controlling at Age=`r a`

```{r}
control_age(a)
a <- a+1
```

## Relationship between height and FEV controlling at Age=`r a`

```{r}
control_age(a)
a <- a+1
```

## Relationship between height and FEV controlling at Age=`r a`

```{r}
control_age(a)
a <- a+1
```

## Relationship between height and FEV controlling at Age=`r a`

```{r}
control_age(a)
a <- a+1
```

## Relationship between height and FEV controlling at Age=`r a`

```{r}
control_age(a)
a <- a+1
```

## Relationship between height and FEV controlling at Age=`r a`

```{r}
control_age(a)
a <- a+1
```

## Relationship between height and FEV controlling at Age=`r a`

```{r}
control_age(a)
a <- a+1
```

## Relationship between height and FEV controlling at Age=`r a`

```{r}
control_age(a)
a <- a+1
```

## Relationship between height and FEV controlling at Age=`r a`

```{r}
control_age(a)
a <- a+1
```

## Relationship between height and FEV controlling at Age=`r a`

```{r}
control_age(a)
a <- a+1
```

## Relationship between height and FEV controlling at Age=`r a`

```{r}
control_age(a)
a <- a+1
```

## Unadjusted relationship between age and fev

```{r}
lm(fev~age, data=pulmonary) %>%
  tidy %>%
  pull(estimate) -> bu
ggplot(pulmonary, aes(age, fev, label=height)) +
  geom_text() +
  geom_abline(slope=bu[2], intercept=bu[1])
h <- 48
```

## Relationship between age and FEV controlling for height between `r h-2` and `r h+1.5`

```{r}
g <- 4*floor((pulmonary$height-2)/4)+4
control_height <- function(a) {
  co=ifelse(g==h, "black", "gray")
  o <- order(co)
  pulmonary[o, ] %>%
    ggplot(aes(age, fev, label=height)) +
      geom_text(color=co[o]) +
      geom_abline(slope=b[3], intercept=b[1]+b[2]*h) +
      geom_abline(slope=bu[2], intercept=bu[1], color="gray")
}
h <- 48
control_height(h)
h <- h+4
```

## Relationship between age and FEV controlling for height between `r h-2` and `r h+1.5`

```{r}
control_height(h)
h <- h+4
```

## Relationship between age and FEV controlling for height between `r h-2` and `r h+1.5`

```{r}
control_height(h)
h <- h+4
```

## Relationship between age and FEV controlling for height between `r h-2` and `r h+1.5`

```{r}
control_height(h)
h <- h+4
```

## Relationship between age and FEV controlling for height between `r h-2` and `r h+1.5`

```{r}
control_height(h)
h <- h+4
```

## Relationship between age and FEV controlling for height between `r h-2` and `r h+1.5`

```{r}
control_height(h)
h <- h+4
```

## Relationship between age and FEV controlling for height between `r h-2` and `r h+1.5`

```{r}
control_height(h)
h <- h+4
```
