---
title: "07-02, R code for categorical independent variables"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

## fev data dictionary, 1

```{r data1}
load("../data/module07.RData")

cat(datac$chunk[datac$lab=="data_dictionary"])
cat(datac$chunk[datac$lab=="copyright"])
cat(datac$chunk[datac$lab=="description"])
cat(datac$chunk[datac$lab=="additional_description"])
```

## fev data dictionary, 2

```{r data2}
cat(datac$chunk[datac$lab=="download_url"])
cat(datac$chunk[datac$lab=="format"])
cat(datac$chunk[datac$lab=="varnames"])
cat(datac$chunk[datac$lab=="missing_value_code"])
cat(datac$chunk[datac$lab=="size"])
```

## fev data dictionary, 3

```{r data3}
cat(datav$chunk[1])
cat(datav$chunk[2])
cat(datav$chunk[3])
```

## fev data dictionary, 4

```{r data4}
cat(datav$chunk[4])
cat(datav$chunk[5])
```

## simon-5501-07-fev.qmd, 1

```{r write1}
cat(codec[1])
```

:::notes
The first few lines are the documentation header
:::

## simon-5501-07-fev.qmd, 2

```{r write2}
cat(codec[2])
```

## simon-5501-07-fev.qmd, 3

```{r write3}
cat(codec[3])
```

## simon-5501-07-fev.qmd, 4

```{r write4}
cat(codec[4])
```

## simon-5501-07-fev.qmd, 5

```{r write5}
cat(codec[5])
```

## simon-5501-07-fev.qmd, 6

```{r write6}
cat(codec[6])
```

## simon-5501-07-fev.qmd, 7

```{r write7}
cat(codec[7])
```

## simon-5501-07-fev.qmd, 8

```{r write8}
cat(codec[8])
```

## simon-5501-07-fev.qmd, 9

```{r write9}
cat(codec[9])
```

## simon-5501-07-fev.qmd, 10

```{r write10}
cat(codec[10])
```

## simon-5501-07-fev.qmd, 11

```{r write11}
cat(codec[11])
```

## simon-5501-07-fev.qmd, 12

```{r write12}
cat(codec[12])
```

## simon-5501-07-fev.qmd, 13

```{r write13}
cat(codec[13])
```

## simon-5501-07-fev.qmd, 14

```{r write14}
cat(codec[14])
```

## simon-5501-07-fev.qmd, 15

```{r write15}
cat(codec[15])
```

## simon-5501-07-fev.qmd, 16

```{r write16}
cat(codec[16])
```

## simon-5501-07-fev.qmd, 17

```{r write17}
cat(codec[17])
```

## simon-5501-07-fev.qmd, 18

```{r write18}
cat(codec[18])
```

## simon-5501-07-fev.qmd, 19

```{r write19}
cat(codec[19])
```

## simon-5501-07-fev.qmd, 20

```{r write20}
cat(codec[20])
```

## simon-5501-07-fev.qmd, 21

```{r write21}
cat(codec[21])
```

## simon-5501-07-fev.qmd, 22

```{r write22}
cat(codec[22])
```

## simon-5501-07-fev.qmd, 23

```{r write23}
cat(codec[23])
```

