---
title: "11-02, Log transformation"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

## Analysis of variance model

-   Sample 1: $Y_{11}, Y_{12}, \cdots, Y_{1n_1}$ are $N(\mu_1, \sigma)$
-   Sample 2: $Y_{21}, Y_{22}, \cdots, Y_{2n_2}$ are $N(\mu_2, \sigma)$
-   $\cdots$
-   Sample k: $Y_{k1}, Y_{k2}, \cdots, Y_{kn_k}$ are $N(\mu_k, \sigma)$
$\ $                                          
-   $H_0:\ \mu_1=\mu_2=...=\mu_k$
-   $H_1:\ \mu_i \ne \mu_j$ for some i, j


## Violation of assumptions

-   Non-normality
-   Heterogeneity
-   Lack of independence

## When to consider a log transformation

-   Only positive values
-   Max/min > 3
-   Skewed distribution
-   Groups with larger means have more variation

## Log transformation, 1

```{r log-longevity-anova-1}
fly |>
  mutate(log_longevity=log10(longevity)) -> log_fly
m3 <- aov(log_longevity ~ cage, data=log_fly)
tidy(m3)
```

:::notes
Although there are no problems with heterogeneity or non-normality with this particular dataset, here is an illustration of how to use a log transformation with analysis of variance.
:::

## Log transformation, 2

```{r log-longevity-anova-2}
t3 <- TukeyHSD(m3, ordered=TRUE)
t3
```

:::notes
Here are the results of the Tukey post hoc tests on the log scale. These intervals are difficult to interpret.
:::

## Log transformation, 3

```{r log-longevity-anova-3}
t3$cage |>
  data.frame() |>
  select(diff, lwr, upr) |> 
  mutate(
    diff=10^diff,
    lwr=10^lwr,
    upr=10^upr)
```

:::notes
Here are the results translated back to the original scale. The confidence intervals are intervals for the ratio of two geometric means.
:::

