---
title: "11-01, Relationship between analysis of variance and linear regression"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

## Indicator variables, 1

-  Two levels is less complex with three or more levels
   -   R assigns 0 to first category (alphabetically)
   -   R assigns 1 to second category
-  Interpretation
   -   Intercept is estimated average outcome for first category
   -   Slope is the estimated average change
       -   Second category average minus first category average

## Indicator variables, 2

-  With k levels, you need k-1 indicators
   -   First indicator
        -   R assigns 1 to second category (alphabetically)
        -   R assigns 0 to all other categories
   -   Second indicator
        -   R assigns 1 to third category (alphabetically)
        -   R assigns 0 to all other categories
   -   And so on

## Indicator variables, 3

-  Interpretation
   -   Intercept is estimated average outcome for first category
   -   First slope is the estimated average change
       -   Second category average minus first category average
   -   Second slope is the estimated average change
       -   Third category average minus first category average
   -   And so on
   
## Example using fruitfly lifespans, 1

-   Experiment with 125 cages
    -   Does fruitfly mating affect average male lifespan?
    -   Isolate a male fruitfly with 
        -   virgin females, 
        -   pregnant females, or
        -   no females
-   Cages 1-25 have one male fruitfly, no female fruit flies
    -   Set partners=0, type=9

## Example using fruitfly lifespans, 2

-   Cages 26-50 have one male fruitfly, one pregnant female
    -   Set partners=1, type=0
    -   Males will not mate with pregnant females
-   Cages 51-75 have one male fruitfly, one virgin female
    -   Set partners=1, type=1
-   Cages 76-100 have one male fruitfly, eight pregnant females
    -   Set partners=8, type=0
-   Cages 101-125 have one male fruitfly, eight virgin females
    -   Set partners=8, type=1

```{r 11-01-setup}
#| message: false
#| warning: false
library(tidyverse)
load("../data/chopper.RData")
```

```{r 11-01-dictionary}
#| output: asis
partition_yaml("fruitfly.yaml") |>
	cat(sep="\n")
```

## Fruitfly lifespan boxplots

```{r 11-01-read-fruitfly-data}
#| message: false
#| warning: false
library(broom)
library(tidyverse)
load("../data/fruitfly.RData")
```

```{r 11-01-boxplots}
#| fig.width: 6
#| fig.height: 2.5
fly |>
  ggplot(aes(cage, longevity)) +
    geom_boxplot() +
    ggtitle("Graph drawn by Steve Simon on 2024-10-23") +
    xlab("Cage residents") +
    ylab("Lifespan (days)") +
    coord_flip()
```

## Fruitfly lifespan group means

```{r 11-01-group-means}
fly |>
  group_by(cage) |>
  summarize(
    longevity_mn=mean(longevity),
    longevity_sd=sd(longevity),
    n=n())
```

## Fruitfly lifespan analysis using aov

```{r 11-01-aov}
#| echo: true
m1 <- aov(longevity ~ cage, data=fly)
anova(m1)
```

## Fruitfly lifespan analysis using lm, 1

```{r 11-01-lm-anova}
#| echo: true
m2 <- lm(longevity ~ cage, data=fly)
anova(m2)
```


## Fruitfly lifespan analysis using lm, 2

```{r 11-01-lm-tidy}
#| echo: true
m2 <- lm(longevity ~ cage, data=fly)
tidy(m2)
```

## Fruitfly lifespan analysis using lm, 3

```{r 11-01-lm-means}
fly |>
  group_by(cage) |>
  summarize(longevity_mn=mean(longevity)) |>
  mutate(reference_mn=first(longevity_mn)) |>
  mutate(mean_difference=longevity_mn-reference_mn)
```

