---
title: "Analysis of fruitfly data"
format: 
  html:
    embed-resources: true
---

This program reads data on fruit fly longevity. Find more information in the [data dictionary][dd].

[dd]: https://github.com/pmean/data/blob/master/files/fruitfly.yaml

This code was written by Steve Simon on 2024-10-23 and is placed in the public domain.

## Load the tidyverse library

```{r setup}
#| message: false
#| warning: false
library(broom)
library(tidyverse)
```

For most of your programs, you should load the tidyverse library. The broom library converts your output to a nicely arranged dataframe. The messages and warnings are suppressed.

## List the variable names

```{r variable-list}
fn <- "https://jse.amstat.org/datasets/fruitfly.dat.txt"
vlist <- c(
  "id",
  "partners",
  "type",
  "longevity",
  "thorax",
  "sleep")
```

When a dataset does not have variables on the first line, you need to specify them in the code.

## Read the data and view a brief summary

```{r read}
fly <- read_fwf(
  "../data/fruitfly.txt",
  col_types="nnnnnn",
  fwf_widths(
    widths=c(2, 2, 2, 3, 5, 3),
    col_names=vlist))
glimpse(fly)
```

The fruitfly dataset has a fixed width format (fwf). You need to specify the columns that each variable uses.

## Create cage groups

```{r cage}
fly$cage <- 
  case_when(
    fly$partners==0 & fly$type==9 ~ "No females",
    fly$partners==1 & fly$type==0 ~ "One pregnant female",
    fly$partners==1 & fly$type==1 ~ "One virgin female",
    fly$partners==8 & fly$type==0 ~ "Eight pregnant females",
    fly$partners==8 & fly$type==1 ~ "Eight virgin females")
```

The five categories represent different combinations of partners and type.

## Calculate descriptive statistics

```{r longevity-means}
fly |>
  group_by(cage) |>
  summarize(
    longevity_mn=mean(longevity),
    longevity_sd=sd(longevity),
    n=n())
```

The mean lifespan is much lower for the eight virgin females group. The standard deviations are reasonably small and somewhat consistent across all groups.

## Draw boxplot

```{r longevity-boxplot}
#| fig.width: 6
#| fig.height: 2.5
fly |>
  ggplot(aes(cage, longevity)) +
    geom_boxplot() +
    ggtitle("Graph drawn by Steve Simon on 2024-10-23") +
    xlab("Cage residents") +
    ylab("Lifespan (days)") +
    coord_flip()
```

The boxplot also shows that the eight virgin females group has lower lifespans.
The variation is roughly comparable across all groups. The distributions are
reasonably symmetric with no outliers.

## One factor analysis of variance for longevity

```{r longevity-one-factor-anova}
m1 <- aov(longevity ~ cage, data=fly)
tidy(m1)
```

The F-ratio is large and the p-value is small. Conclude that there is a 
difference among some or all of the population mean lifespans.

## Linear model for longevity, 1

```{r longevity-lm-1}
m2 <- lm(longevity ~ cage, data=fly)
anova(m2)
```

You can use linear regression to reach the same conclusion. The sums of squares,
degrees of freedom, F-ratio, and p-value all match.

## Linear model for longevity, 2

```{r longevity-lm-2}
tidy(m2)
```

The linear model creates indicator variable for four out of the five category
levels. The regression slopes represent the estimated average difference in 
means between each category and the reference category.

## Linear model for longevity, 3

```{r longevity-lm-3}
fly |>
  group_by(cage) |>
  summarize(longevity_mn=mean(longevity)) |>
  mutate(reference_mn=first(longevity_mn)) |>
  mutate(mean_difference=longevity_mn-reference_mn)
```

This table shows how to interpret the intercept and the five slope terms. The
intercept is the estimated average lifespan for the first level of the
categorical variable. The slope terms are the estimated average changes from the
other levels of the categorical variable and the first level.

## Re-order cage groups, 1

```{r re-order-1}
fly$cage_1 <- 
  factor(
    fly$cage,
    levels = c(
      "No females",
      "One pregnant female",
      "Eight pregnant females",
      "One virgin female",
      "Eight virgin females"))
```

It makes more sense to compare every other category to the No females category
because No females represents a control condition.

## Re-order cage groups, 2

```{r re-order-2}
m3 <- lm(longevity ~ cage_1, data=fly)
tidy(m3)
```

Use the factor function to re-order the categories so that the No females
level is listed as the first level.

## Log transformation, 1

```{r log-longevity-anova-1}
fly |>
  mutate(log_longevity=log10(longevity)) -> log_fly
m3 <- aov(log_longevity ~ cage, data=log_fly)
tidy(m3)
```

Although there are no problems with heterogeneity or non-normality, here is an illustration of how to use a log transformation with analysis of variance.

## Log transformation, 2

```{r log-longevity-anova-2}
t3 <- TukeyHSD(m3, ordered=TRUE)
t3
```

## Log transformation, 3

```{r log-longevity-anova-3}
t3$cage |>
  data.frame() |>
  select(diff, lwr, upr) |> 
  mutate(
    diff=10^diff,
    lwr=10^lwr,
    upr=10^upr)
```

## Log transformation, 4

```{r log-longevity-anova-4}
t3$cage |>
  data.frame() |>
  select(diff, lwr, upr) |> 
  mutate(
    diff=10^(-diff),
    upr=10^(-lwr),
    lwr=10^(-upr))
```

## Kruskal-Wallis test

```{r kw}
kruskal.test(longevity ~ cage, data=fly)
```

It may not be needed with this particular dataset, but this is an illustration
of how to use the Kruskal-Wallis test. 

## Create a subset of fruitfly dataset

```{r subset}
fly |>
  filter(type != 9) -> fly_subset
```

If you exclude the pure control group (No females), you can analyze the two
factors, partners and type individually. Partners has two category levels, 1 for
when one female was included in the cage and 8 for when eight females were 
included in the cage. Type also has two category levels, 0 for pregnant female
fly/flies and 1 for virgin fly/flies. A male fly will not mate with a pregnant
females, so you can think of this as a second level of controls. The two factors
are crossed, meaning that every possible combination of partners and type has 
outcomes measured.

## Draw boxplot of longevity by partners and type

```{r boxplots-partners-type}
#| fig.width: 6
#| fig.height: 2.5
fly_subset |>
  ggplot(aes(factor(partners), longevity, fill=factor(type))) +
    geom_boxplot() +
    xlab("Number of partners") +
    ylab("Lifespan in days") +
    ggtitle("Graphs drawn by Steve Simon on 2024-10-28") +
    labs(fill="Female fly type") +
    scale_fill_discrete(labels=c("Pregnant", "Virgin")) +
    coord_flip()
```

## Calculate average longevity by partners and type

```{r by-partners-and-type}
#| message: false
fly_subset |>
  group_by(partners, type) |>
  summarize(
    longevity_mn=mean(longevity),
    longevity_sd=sd(longevity),
    n=n()) -> fly_means
fly_means
```

## Draw line graph

```{r line-graph}
fly_means |>
  ggplot(aes(
    factor(partners), 
    longevity_mn, 
    group=factor(type),
    color=factor(type))) +
    geom_line() +
  xlab("Number of partners") +
  ylab("Lifespan in days") +
  labs(color="Female fly type") +
  scale_color_discrete(labels=c("Pregnant", "Virgin"))
```

## Interpretation of the interaction

Male fruitflies sharing a cage with pregnant females have longer average lifespans than sharing a cage with virgin females. The difference in lifespans is much more extreme when sharing with eight female fruitflies compared to one female fruitfly.

The line for pregnant (sharing a cage with pregnant females) is close to flat, indicating that there is little of the number of pregnant females on average longevity.

## Analysis of variance including the interaction

```{r anova-with-interaction}
m4 <- aov(longevity ~ factor(type)*factor(partners), data=fly_subset)
anova(m4)
```

There is a statistically significant interaction between the number of female partners and the type of partners (pregnant or virgin).

## Save important files for later use

```{r save}
save(
  fly,
  log_fly,
  fly_subset,
  file="../data/fruitfly.RData")
```