---
title: "Analysis of fruitfly data"
format: 
  html:
    embed-resources: true
---

This program reads data on fruit fly longevity. Find more information in the [data dictionary][dd].

[dd]: https://github.com/pmean/data/blob/master/files/fruitfly.yaml

This code was written by Steve Simon on 2024-10-23 and is placed in the public domain.

## Load the tidyverse library

```{r setup}
#| message: false
#| warning: false
library(broom)
library(tidyverse)
```

For most of your programs, you should load the tidyverse library. The broom library converts your output to a nicely arranged dataframe. The messages and warnings are suppressed.

## List the variable names

```{r variable-list}
fn <- "https://jse.amstat.org/datasets/fruitfly.dat.txt"
vlist <- c(
  "id",
  "partners",
  "type",
  "longevity",
  "thorax",
  "sleep")
```

When a dataset does not have variables on the first line, you need to specify them in the code.

## Read the data and view a brief summary

```{r read}
fly <- read_fwf(
  "../data/fruitfly.txt",
  col_types="nnnnnn",
  fwf_widths(
    widths=c(2, 2, 2, 3, 5, 3),
    col_names=vlist))
glimpse(fly)
```

The fruitfly dataset has a fixed width format (fwf). You need to specify the columns that each variable uses.

## Create cage groups

```{r cage}
fly$cage <- 
  case_when(
    fly$partners==0 & fly$type==9 ~ "No females",
    fly$partners==1 & fly$type==0 ~ "One pregnant female",
    fly$partners==1 & fly$type==1 ~ "One virgin female",
    fly$partners==8 & fly$type==0 ~ "Eight pregnant females",
    fly$partners==8 & fly$type==1 ~ "Eight virgin females")
```

The five categories represent different combinations of partners and type.

## Calculate descriptive statistics

```{r longevity-means}
fly |>
  group_by(cage) |>
  summarize(
    longevity_mn=mean(longevity),
    longevity_sd=sd(longevity),
    n=n())
```

The mean lifespan is much lower for the eight virgin females group. The standard deviations are reasonably small and somewhat consistent across all groups.

## Draw boxplot

```{r longevity-boxplot}
#| fig.width: 6
#| fig.height: 2.5
fly |>
  ggplot(aes(cage, longevity)) +
    geom_boxplot() +
    ggtitle("Graph drawn by Steve Simon on 2024-10-23") +
    xlab("Cage residents") +
    ylab("Lifespan (days)") +
    coord_flip()
```

The boxplot also shows that the eight virgin females group has lower lifespans. The variation is roughly comparable across all groups. The distributions are reasonably symmetric with no outliers.

## Analysis of variance for longevity

```{r longevity-anova}
m1 <- aov(longevity ~ cage, data=fly)
tidy(m1)
```

The F-ratio is large and the p-value is small. Conclude that there is a difference among some or all of the population mean lifespans.

## Linear model for longevity, 1

```{r longevity-lm-1}
m2 <- lm(longevity ~ cage, data=fly)
anova(m2)
```

You can use linear regression to reach the same conclusion. The sums of squares, degrees of freedom, F-ratio, and p-value all match.

## Linear model for longevity, 2

```{r longevity-lm-2}
tidy(m2)
```

The linear model creates indicator variable for four out of the five category levels. The regression slopes represent the estimated average difference in means between each category and the reference category.

## Linear model for longevity, 3

```{r longevity-lm-3}
fly |>
  group_by(cage) |>
  summarize(longevity_mn=mean(longevity)) |>
  mutate(reference_mn=first(longevity_mn)) |>
  mutate(mean_difference=longevity_mn-reference_mn)
```

## Re-order cage groups, 1

```{r re-order-1}
fly$cage_1 <- 
  factor(
    fly$cage,
    levels = c(
      "No females",
      "One pregnant female",
      "Eight pregnant females",
      "One virgin female",
      "Eight virgin females"))
```

## Re-order cage groups, 2

```{r re-order-2}
m3 <- lm(longevity ~ cage_1, data=fly)
tidy(m3)
```

## Tukey post hoc comparisons

```{r longevity-tukey}
t1 <- TukeyHSD(m1, ordered=TRUE)
```

## Log transformation, 1

```{r log-longevity-anova-1}
fly |>
  mutate(log_longevity=log10(longevity)) -> log_fly
m3 <- aov(log_longevity ~ cage, data=log_fly)
tidy(m3)
```

## Log transformation, 2

```{r log-longevity-anova-2}
t3 <- TukeyHSD(m3, ordered=TRUE)
t3
```

## Log transformation, 3

```{r log-longevity-anova-3}
t3$cage |>
  data.frame() |>
  select(diff, lwr, upr) |> 
  mutate(
    diff=10^diff,
    lwr=10^lwr,
    upr=10^upr)
```

## Log transformation, 4

```{r log-longevity-anova-4}
t3$cage |>
  data.frame() |>
  select(diff, lwr, upr) |> 
  mutate(
    diff=10^(-diff),
    upr=10^(-lwr),
    lwr=10^(-upr))
```

## Kruskal-Wallis test

```{r kw}
kruskal.test(longevity ~ cage, data=fly)
```

## Save important files for later use

```{r save}
save(
  fly,
  log_fly,
  m1,
  m2,
  m3,
  t3,
  file="../data/firefly.RData")
```