---
title: "10-02, The F-test"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

## The F-distribution

-   Ratio of two measures of variation
-   Two measures are comparable if F is close to 1
-   F distribution is skewed and can never be negative
-   Two measures of degrees of freedom
    -   $df_1$ = degrees of freedom for numerator
    -   $df_2$ = degrees of freedom for denominator

:::notes
The F distribution appears many times in Statistics when you are comparing two different measures of variation.
:::

## Graph of the F distribution
  
```{r 10-02-f-plot}
#| message: false
#| warning: false

library(tidyverse)
load("../data/module10.RData")
x <- seq(0, 6, length=99)
y <- df(x, 5, 28)
data.frame(x, y) |>
	ggplot(aes(x, y)) + 
	geom_line() +
	ggtitle("Graph drawn by Steve Simon on 2024-10-22")
```

:::notes
Here is a picture of the F distribution. 
:::

## The F-test

-   $H_0:\ \mu_1=\mu_2=...=\mu_k$
-   $H_1:\ \mu_i \ne \mu_j$ for some i, j
    -   Compute F = MS(Between) / MS(Within)
    -   Accept H0 if the F < $F(1-\alpha, k-1, N-k)$
-   Only reject $H_0$ for large values

:::notes
You only reject the null hypothesis for large values of F. If the variation between groups is about equal to the variation within groups or if it is much smaller than the variation within groups, you should accept the null hypothesis.
:::

## The p-value for the F test

-   p-value = P[F(k-1, N-k) > F]
-   Accept $H_0$ if p-value > $\alpha$.

:::notes
R will compute a p-value for you and you compare the p-value to alpha. With a large p-value (greater than alpha), you should accept the null hypothesis. With a small p-value (less than or equal to alpha), you should reject the null hypothesis.
:::

## R code for the F test

```{r 10-02-f-test}
m1 <- aov(y ~ factor(g), data=aa)
tidy(m1)
```

Because the F-ratio is large and the p-value is less than 0.05, you should reject $H_0$ and conclude that there is a difference among the means.

## R-squared

-   $R^2 = \frac{SS(Between)}{SS(Total)}$
    -   Proportion of variation explained by groups
    
```{r 10-02-r-squared}
#| echo: true
glance(m1)$r.squared
```

Approximately 74% of the variation in measurements can be accounted for by the grouping.
