---
title: "Paired t-test for postural sway data and alternatives"
format: 
  html:
    embed-resources: true
execute: 
  error: true
---

This program runs a paired t-test. Consult the [data dictionary][dic] for information about the data itself.

[dic]: https://github.com/pmean/data/blob/main/files/postural-sway.yaml

This program was written by Steve Simon on 2024-10-17. It is placed in the public domain.

## Libraries

```{r setup}
#| message: false
#| warning: false
library(broom)
library(tidyverse)
```

## Read data

```{r read-sway}
sway <- read_tsv(
  file="../data/postural-sway.txt",
  col_types="cnn")
names(sway) <- tolower(names(sway))
glimpse(sway)
```

## Paired t-test

The research team is also interested in whether the front-to-back sway is larger than the side-to-side sway, combining both the elderly and young patients into one group.

## Compute and graph differences

```{r compute-difference-1}
sway |>
  mutate(diff_sway=fbsway-sidesway) -> paired_differences
paired_differences |>
  summarize(
    diff_mn=mean(diff_sway),
    diff_sd=sd(diff_sway))
```

The average difference is positive, indicating that front-to-back-sways are larger on average. The standard deviation of the differences is large. 

## Boxplot of differences

```{r compute-difference-2}
#| fig.height: 1.5
#| fig.width: 6
paired_differences |>
  ggplot(aes(diff_sway, "Combined")) +
    geom_boxplot() +
    ggtitle("Graph drawn by Steve Simon on 2024-10-13") +
    xlab("Front-to-back sway minus side-to-side sway") +
    ylab(" ")
```

More than 75% of the differences are positive, also indicating that front-to-back sways tend to be larger.

## Paired t-test

```{r paired-t-test}
m1 <- t.test(
  sway$fbsway, 
  sway$sidesway,
  paired=TRUE,
  alternative="two.sided")
m1
```
The p-value is small, indicating that front-to-back sways are significantly larger than side-to-side sways. The 95% confidence intervals shows that the mean difference is at least 0.21 units larger for front-to-back sway and possibly as large as 7.0 units.

## Compute a log transformation

```{r compute-log}
sway |>
  mutate(log_fb=log10(fbsway)) |>
  mutate(log_side=log10(sidesway)) -> log_sway
```

## Descriptive statistics for log data

```{r log-descriptives}
log_sway |>
  summarize(
    mn_log_fb=mean(log_fb),
    mn_log_side=mean(log_side)) -> log_means
log_means
10^log_means
```

## Paired t-test after log transform

```{r log-paired-t-test}
m2 <- t.test(
  log10(sway$fbsway), 
  log10(sway$sidesway),
  paired=TRUE,
  alternative="two.sided")
```



## Equivalent analysis using linear regression, 1

```{r pivot}
sway$patient_id <- 1:length(sway$age)
sway |>
  pivot_longer(
    cols=c("fbsway", "sidesway"),
    names_to="direction",
    values_to="sway") -> sway_long
glimpse(sway_long)
```

## Equivalent analysis using linear regression, 2

```{r lm-analysis}
m1 <- lm(
  sway ~ direction + factor(patient_id), 
  data=sway_long)
tidy(m1)
confint(m1)
```