---
title: "r-02-02, Using mutate, summarize"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
execute:
  echo: true
  eval: false
---

## Data dictionary for sleep

The next few examples will use a different data file, sleep. Refer to the [data dictionary][ref21] for details about this file.

[ref21]: https://github.com/pmean/data/blob/main/files/sleep.yaml

## Read text file

```{r}
#| label: 02-02-setup
#| message: false
#| warning: false
#| echo: true

library(tidyverse)
```

```{r}
#| label: 02-02-read-sleep
#| echo: true

sleep <- read_tsv(
  file="../data/sleep.txt",
  col_types="cnnnnnnnnnn")
glimpse(sleep)
```

## Convert brain weight from grams to kilograms

```{r}
#| label: 02-02-convert-units

sleep |>
  mutate(brain_wt_kg=BrainWt/1000) -> sleep
sleep |>
  slice(1:10) |>
  select(Species, BrainWt, brain_wt_kg)
```

## Calculate brain to body weight ratio

```{r}
#| label: 02-02-calculate-percent

sleep |>
  mutate(brain_pct=100*brain_wt_kg/BodyWt) |>
  mutate(brain_pct=round(brain_pct, 2)) |>
  mutate(brain_pct=paste0(brain_pct, "%")) -> sleep

sleep |>
  slice(1:10) |>
  select(Species, brain_wt_kg, BodyWt, brain_pct)
```

## Summarize body weight

```{r}
#| label: 02-02-summarize

sleep |>
  summarize(
    body_wt_mean=round(mean(BodyWt), 0),
    body_wt_sd=round(sd(BodyWt), 0),
    body_wt_min=min(BodyWt),
    body_wt_max=max(BodyWt))
```

## Interpretation of body weight statistics

The mean body weight is just a bit below 200 kilograms with a standard devation of about 900. The range of weights is extremely broad. The smallest animal is 0.005 kilograms (or 5 grams) and the largest is over 6,600 kilograms.

