---
title: "r-02-02, Using mutate, summarize"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

## Data dictionary for burger-calories

The next few examples will use a different data file, sleep. Refer to the [data dictionary][ref21] for details about this file.

[ref21]: https://github.com/pmean/data/blob/main/files/sleep.yaml

## Read text file

```{r}
#| label: setup2
#| message: false
#| warning: false

library(tidyverse)
```

```{r}
#| label: read-sleep
#| echo: true

sleep <- read_tsv(
  file="../data/sleep.txt",
  col_types="cnnnnnnnnnn")
glimpse(sleep)
```

::: notes

The sleep data is too big to display on a single slide. The [glimpse][ref22] function provides a brief overview. It lists the number of rows and columns. This is followed with the name for each variable, its type, and the first few values.

[ref22]: https://dplyr.tidyverse.org/reference/glimpse.html

:::

## Convert brain weight from grams to kilograms

```{r}
#| label: convert-units
#| echo: true

sleep |>
  mutate(brain_wt_kg=BrainWt/1000) -> sleep
sleep |>
  slice(1:10) |>
  select(Species, BrainWt, brain_wt_kg)
```

::: notes



:::


## Calculate brain to body weight ratio

```{r}
#| label: calculate-percent
#| echo: true

sleep |>
  mutate(brain_pct=100*brain_wt_kg/BodyWt) |>
  mutate(brain_pct=round(brain_pct, 2)) |>
  mutate(brain_pct=paste0(brain_pct, "%")) -> sleep

sleep |>
  slice(1:10) |>
  select(Species, brain_wt_kg, BodyWt, brain_pct)
```

## Summarize

```{r}
#| label: summarize
#| echo: true

sleep |>
  summarize(
    body_wt_mean=mean(BodyWt),
    body_wt_sd=sd(BodyWt))
```
