---
title: "r-01-01, Subsetting a tibble/dataframe"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

## Read burger-calories

```{r}
#| label: read-yaml
#| message: false
#| warning: false

library(tidyverse)
library(yaml)
b <- read_yaml("https://raw.githubusercontent.com/pmean/data/refs/heads/main/files/burger-calories.yaml")
```


```{r}
#| label: read-tsv
#| echo: true

burgers <- read_tsv(
  file="../data/burger-calories.txt", 
  col_types="nnn")
burgers
```


:::notes

:::

## Select, 1

```{r}
#| label: select
#| echo: true

burgers |>
  select(Calories)
```

::: notes

The [select][ref01] function will include or exclude variables from a tibble/dataframe.

[ref01]: https://dplyr.tidyverse.org/reference/select.html

:::

## Select, 2

```{r}
#| label: select-2
#| echo: true

burgers |>
  select(-Fat)
```


## Alternatives to select, 1

```{r}
#| label: select-alternative-1
#| echo: true

burgers$Calories
```

::: notes

There are some "old school" ways to select a single column of data. These will produce a vector rather than a tibble/dataframe.

:::

## Alternatives to select, 2

```{r}
#| label: select-alternative-2
#| echo: true

burgers[ , 1]
```

## Alternatives to select, 3

```{r}
#| label: select-alternative-3
#| echo: true

burgers[ , 3]
```

## Slice, 1

```{r}
#| label: slice-1
#| echo: true

burgers |>
  slice(1)
```

::: notes

The [slice][ref01] function selects rows based on their position.

[ref01]: https://dplyr.tidyverse.org/reference/slice.html

:::

## Slice, 2

```{r}
#| label: slice-2
#| echo: true

burgers |>
  slice_max(Calories)
```

::: notes

The slice_max function, described in the previous link, selects rows based on the size of one of the variables.

:::

## Filter, 1

```{r}
#| label: filter-1
#| echo: true

burgers |>
  filter(Sodium == 1500)
```

::: notes

The [filter][ref02] function selects rows of a tibble or dataframe based on a logical condition (an R expression that produces true/false values).

[ref02]: https://dplyr.tidyverse.org/reference/filter.html

:::

## Filter, 2

```{r}
#| label: filter-2
#| echo: true

burgers |>
  filter(Calories >= 600)
```

## Alternatives to slice/filter