---
title: "r-01-04, Histograms, scatterplots"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

## Default histogram for body weight, 1

```{r}
#| label: setup-4
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
```

```{r}
#| label: read-sleep-4
#| echo: false

sleep <- read_tsv(
  file="../data/sleep.txt",
  col_types="cnnnnnnnnnn")
```

```{r}
#| label: default-histogram-1
#| echo: true
sleep |>
  ggplot() +
  aes(x=BodyWt) +
  geom_histogram() -> basic_histogram
```

::: notes

R has at least three different approaches for graphs: base graphs, lattice, and ggplot2. All are excellent. I will rely almost entirely on ggplot2 for graphs that I use. 

:::

## Default histogram for body weight, 2


```{r}
#| label: default-histogram-2
#| echo: true
basic_histogram
```

::: notes

Never settle for the default. I did change one default right off the bat. The default in ggplot for a histogram or bar chart is a dark gray bar with a dark gray border. There are a few times that this is a reasonable choice, but more often than not choosing the same color value for the border and the interior is confusing.

:::

## Better colors for the histogram, 1

```{r}
#| label: better-histogram-1
#| echo: true
sleep |>
  ggplot() +
  aes(x=BodyWt) +
  geom_histogram(
    color="black",
    fill="white") -> better_histogram
```

::: notes

I usually try a black border. The color argument specifies the color of the border and the fill argument specifies the color or the interior.

:::

## Better colors for the histogram, 2

```{r}
#| label: better-histogram-2
#| echo: true

better_histogram
```

::: notes

It may not be so obvious with this graph, but I do think that black borders with white interiors improve the appearace of most bar charts and histograms.

::::

## Labeled histogram, 1

```{r}
#| label: histogram-with-labels-1
#| echo: true

better_histogram +
  xlab("Body weight (kg)") +
  ggtitle("Graph drawn by Steve Simon on 2025-01-02") -> labeled_histogram
```

::: notes

You should always use the [xlab, ylab, and ggtitle][ref41] functions to make your graphs more interpretable. The xlab function changes the label on the x-axis from the default (just the variable name itself) to a more descriptive label. There is a ylab function as well, but I did not need it for the histogram because the word "count" that ggplot uses on the y-axis for its histograms is sufficiently descriptive. The ggtitle function places a title at the top of the graph. For this class, I require you to put your name and the date as the title. If you are preparing a graph for a publication, you will probably need a different title.

[ref41]: https://ggplot2.tidyverse.org/reference/labs.html

:::

## Labeled histogram, 2

```{r}
#| label: histogram-with-labels-2
#| echo: true

labeled_histogram
```

::: notes

This is what the labeled histogram looks like.

:::

## Log transformed histogram, 1

```{r}
#| label: histogram-with-log-scale-1
#| echo: true

ticks <- 10^(-2:3)
labeled_histogram +
  scale_x_log10(
    breaks=ticks,
    minor_breaks=NULL,
    labels=ticks) -> histogram_with_log10
```

::: notes

The [scale_x_log10][ref42] function is one of a series of functions that controls how the scale of the x axis appears.

[ref42]: https://ggplot2.tidyverse.org/reference/scale_continuous.html

:::


## Log transformed histogram, 2

```{r}
#| label: histogram-with-log-scale-2
#| echo: true

histogram_with_log10
```

## Different scales

```{r}
#| echo: false
i <- 10^(-3:3)
j <- c(1, 3)
ticks <- kronecker(i, j)
```

## Labeled scatterplot, 1

```{r}
#| label: scatterplot-1
#| echo: true

sleep |>
  ggplot() +
  aes(x=BodyWt, y=BrainWt) +
  geom_point() + 
  xlab("Body weight (kg)") +
  ylab("Brain weight (g)") +
  ggtitle("This graph was drawn by Steve Simon on 2025-01-02") -> labeled_scatterplot
```

## Labeled scatterplot, 2

```{r}
#| label: scatterplot-2
#| echo: true

labeled_scatterplot
```

## Log transformed scatterplot, 1

```{r}
#| label: log-transformed-scatterplot-1
#| echo: true

xticks <- 10^(-2:3)
yticks <- 10^(-2:3)
labeled_scatterplot +
  scale_x_log10(
    breaks=xticks,
    minor_breaks=NULL,
    labels=xticks) +
  scale_y_log10(
    breaks=yticks,
    minor_breaks=NULL,
    labels=yticks)-> log_scatterplot
```

## Log transformed scatterplot, 1

```{r}
#| label: log-transformed-scatterplot-2
#| echo: true

log_scatterplot
```

## Add a trend line, 1

```{r}
#| label: trend-line-1
#| echo: true

log_scatterplot +
  geom_smooth() -> trend_line_graph
```

## Add a trend line, 2

```{r}
#| label: trend-line-2
#| echo: true

trend_line_graph
```
