---
title: "Examples from the video"
format: 
  html:
    embed-resources: true
execute: 
  error: true
---

#### Program details

This program will read in data from an RData file. It will then list all the objects that you just read in, and list the first few lines for one of those objects.

This program was written by Steve Simon on 2024-12-18 and is placed in the public domain. You can use it any way that you please.

```{r}
#| label: setup
#| message: false
#| warning: false
R.version.string
Sys.Date()
library(tidyverse)
```

## Read a small RData file

```{r}
#| label: read-rdata
fat <- read_csv(
  "../data/fat.csv",
  col_types="n")
glimpse(fat)
```

#### Comments on the code

When you use unusual or novel R code, include a brief description.

I will defer discussion of the read_csv function until a later module.

The [glimpse][ref3] function provides a listing of each variable in a data frame
along with the first few values of that variable. It is useful for orienting 
yourself to the way that your data is organized.

#### Interpretation of the output

For all except the most basic output, include a brief interpretation.

In this program, none of the output from any of the program chunks needs an
interpretation.

[ref3]: https://dplyr.tidyverse.org/reference/glimpse.html

## Save important files for later use

```{r}
#| eval: false
save(
  fat,
  log_fly,
  file="../data/fruitfly.RData")
```
