---
title: "Examples from the video"
format: 
  html:
    embed-resources: true
execute: 
  error: true
---

#### Program details

This program will read in data from an RData file. It will then list all the objects that you just read in, and list the first few lines for one of those objects.

This program was written by Steve Simon on 2024-12-18 and is placed in the public domain. You can use it any way that you please.

```{r}
#| label: setup
#| message: false
#| warning: false
R.version.string
Sys.Date()
library(tidyverse)
```

## Read a small RData file

```{r}
#| label: read-rdata
load("../data/two-small-dataframes.RData")
ls()
fat <- read_csv(
  "../data/fat.csv",
  col_types="n")
glimpse(fat)
```

#### Comments on the code

When you use unusual or novel R code, include a brief description.

The [load][ref1] function places objects from an R binary file into your 
environment. The [ls][ref2] function displays all the objects in your environment.

[ref1]: https://bookdown.org/rwnahhas/IntroToR/loading-an-r-dataset.html
[ref2]: https://stat.ethz.ch/R-manual/R-devel/library/base/html/ls.html

#### Interpretation of the output

For all except the most basic output, include a brief interpretation.

In this program, none of the output from any of the program chunks needs an
interpretation.

## Peek at the fat data

```{r}
#| label: glimpse-fat
glimpse(fat)
```

#### Comments on the code

The [glimpse][ref3] function provides a listing of each variable in a data frame along
with the first few values of that variable. It is useful for orienting yourself
to the way that your data is organized.

[ref3]: https://dplyr.tidyverse.org/reference/glimpse.html

## Save important files for later use

```{r}
#| eval: false
save(
  fat,
  log_fly,
  file="../data/fruitfly.RData")
```
