---
title: "r-03-03, Tab delimited files"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

## Tab delimited files

+ Use the tab key instead of the space bar.

```{r tab, echo=FALSE, comment=""}
tx <- read_lines("../data/simple.tsv")
cat(tx, sep="\n")
```

+ Save as simple.tsv

::: notes

Tabs are little bit different and you might want to listen to the live video where I show how the tab key works.

Type the following into notepad and what appears to be multiple blanks, I want you to press the tab key, but only once.

:::

## Using the read_tsv function

```{r tsv}
fn <- "../data/simple.tsv"
raw_data <- read_tsv(fn, 
  col_types="nn")
```

::: notes

The read_tsv function works when you have a tab character as the delimiter. It uses the same arguments as the read_csv function.

:::

## Display tab-delimited data

```{r display-tab}
raw_data
```

::: notes

The data has been read in correctly.

:::

## Break #3

+ What have you learned
  + Reading tab delimited files  
+ What's next?
  + Anything can be a delimiter

## Anything can be a delimiter

+ Type the following into notepad

```{r anything, echo=FALSE, comment=""}
tx <- read_lines("../data/tilde.txt")
cat(tx, sep="\n")
```

+ Save as tilde.txt

::: notes

Type in the same data, but using the tilde (~) as a separator (e.g., 1~4). Save the file as tilde.txt.

Here is how you would use read_delim to read this file.

:::

## Using the read_delim function with delim="~"

```{r tilde}
fn <- "../data/tilde.txt"
raw_data <- read_delim(fn,
  delim="~",
  col_types="nn")
```

::: notes

The data reads in just fine. Why would you use a tilde as a delimiter? Sometimes your data itself includes delimiters like spaces and commas, and then you might want to choose an obscure out of the way symbol to serve as a delimiter.

Another obscure character that is sometimes used as a delimiter is the veritical bar (|).

:::

## Display tilde-delimted data

```{r display-tilde}
raw_data
```

::: notes

The data has been read in correctly.

:::

## Break #4

+ What have you learned
  + Anything can be a delimiter
+ What's next?
  + Reading fixed width files

