---
title: "r-03-04, Other delimiters"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
execute: 
  echo: true
  eval: false
---

## Anything can be a delimiter

```{r}
#| label: 03-04-setup
#| echo: false
#| eval: true
#| message: false
#| warning: false

library(tidyverse)

load("../data/module03.RData")
```

```{r}
#| echo: false
#| eval: true
#| label: 03-04-read-lines

read_lines("../data/tilde.txt")|> 
  cat(sep="\n")
```

::: notes

This file uses the tilde (~) as a separator (e.g., 1~4).

Why would you use a tilde as a delimiter? Sometimes your data itself includes delimiters like spaces and commas, and then you might want to choose an obscure out of the way symbol to serve as a delimiter.

Another obscure character that is sometimes used as a delimiter is the vertical bar (|).

:::

## Using the read_delim function

`r code["tilde"]`

## Live demonstration, 3

Now, you will see a live demonstration of part 3 of [simon-5505-03-demo][ref-demo].
