---
title: "Reading comma delimited files"
format: 
  html:
    embed-resources: true
execute: 
  error: true
editor: source
---

## File details

This program reads data from a text file with comma delimiters. It was written by Steve Simon on 2025-01-06 and is placed in the public domain. You can use this program any way you please.

The data itself is small and artificial, so there is no need for a data dictionary.

```{r}
#| label: setup
#| message: false
#| warning: false

R.version.string
Sys.Date()
library(tidyverse)
```

## Read in data

```{r}
#| label: read-comma

raw_data <- read_csv(
  file="../data/simple.csv",
  col_names=TRUE,
  col_types="nn")

glimpse(raw_data)
```

#### Comments on the code

The [read_csv][ref01] function opens a text file. If it is structured properly, it will create a tibble from the data in the text file. The col_names=TRUE argument tells R that the first row of the file contains the variable names. The col_types="nn" tells R that there are two columns of data, both numeric.

The base R program has a similar function, read.csv, that also works. This function is a bit slower, which can be a problem for very large files. It also has fewer features.

[ref01]: https://readr.tidyverse.org/reference/read_delim.html
[ref02]: https://stat.ethz.ch/R-manual/R-devel/library/utils/html/read.table.html