---
title: "Real world text file examples"
format: 
  html:
    embed-resources: true
execute: 
  error: true
editor: source
---

## File details

This program reads data from a variety of text files. It was written by Steve Simon on 2025-01-09 and is placed in the public domain. You can use this program any way you please.

```{r}
#| label: setup
#| message: false
#| warning: false

R.version.string
Sys.Date()
library(tidyverse)
```

## Program outline

This program will read in three very different text data files. Here are some links.

-   Example 1: binary.csv
    -   [Description][ref01]
    -   [Download][ref02]
-   Example 2: barbershop-music.txt
    -   [Description][ref03]
    -   [Download][ref04]
-   Example 3: airport.txt
    -   [Description][ref05]
    -   [Download][ref06]

[ref01]: https://stats.idre.ucla.edu/r/dae/logit-regression/
[ref02]: https://stats.idre.ucla.edu/stat/data/binary.csv
[ref03]: https://dasl.datadescription.com/datafile/barbershop-music/
[ref04]: https://dasl.datadescription.com/download/data/3061
[ref05]: http://jse.amstat.org/datasets/airport.txt
[ref06]: http://jse.amstat.org/datasets/airport.dat.txt

## Example 1, quick peek

```{r}
#| label: peek-binary

url_binary <- "https://stats.idre.ucla.edu/stat/data/binary.csv"
read_lines(
  file=url_binary, 
  n_max=10)
```

#### Comments on the code

The [read_lines][ref07] function will read each line of the text file into a single string vector. This doesn't allow you to separate out the individual data values. It is, however, a quick way to review the contents of a file to better understand how it is structured. 

[ref07]: https://readr.tidyverse.org/reference/read_lines.html

#### Interpretation of the output

Peeking at the first ten lines of data, you can see that this file uses commas as delimiters and the first line contains variable names. six lines of data, you can see clearly that it is uses comma delimiter and the first row of data contains the variable names.

## Example 1, read_csv

```{r}
#| label: read-binary

example_binary <- read_csv(
  file=fn,
  col_names=TRUE,
  col_types="nnnn")

glimpse(example_binary)
```

#### Interpretation of the output

The data shown here appears to match the description of the file on the web.


## Example 2, quick peek

```{r}
#| label: peek-barbershop

url_barbershop <- "https://dasl.datadescription.com/download/data/3061"
read_lines(
  file=url_barbershop,
  n_max=10)
```

#### Interpretation of the output

It looks like it is a tab delimited file. The read_lines file replaces the tab with a key code (backslash-t). The first row lists the variable names.

## Example 2, read_tsv

```{r}
#| label: read-barbershop

raw_data <- read_tsv(
  file=url_barbershop,
  col_names=TRUE,
  col_types="nnn")

glimpse(raw_data)
```

#### Interpretation of the output

The data appears to match the description.

The read_tsv function works well here.

## Example 3, peek

```{r}
#| label: peek-airport

url_airport <- "http://jse.amstat.org/datasets/airport.dat.txt"
read_lines(
  file=url_airport,
  n_max=10)
```

#### Interpretation of the output

This could be a tab delimited file, but the numbers are right justified. In a tab delimited file, the numbers would be left justified. There are no variable names in the first row of data.

## Example 3, define data columns and names

```{r}
#| label: define-airport

start_column <- c( 1, 22, 44, 51, 58, 67, 77)
end_column <-   c(21, 43, 49, 56, 65, 75, 85)
variable_names <- c(
  "airport",
  "city",
  "scheduled_departures",
  "performed_departures",
  "enplaned_passengers",
  "enplaned_freight",
  "enplaned_mail")
```

## Example 3, read_fwf

```{r}
#| label: read-airport

example_3 <- read_fwf(
  file=url_3, 
  fwf_positions(
    start=start_column, 
    end=end_column),
  col_names=variable_names,
  col_types="ccnnnnn")

glimpse(example_3)
```

#### Interpretation of the output

The data appears to match the description.
