---
title: "Reading fixed width files"
format: 
  html:
    embed-resources: true
execute: 
  error: true
editor: source
---

## File details

This program reads data from a fixed width text file. It was written by Steve Simon on 2025-01-08 and is placed in the public domain. You can use this program any way you please.

The data itself is small and artificial, so there is no need for a data dictionary.

```{r}
#| label: setup
#| message: false
#| warning: false

R.version.string
Sys.Date()
library(tidyverse)
```

## Read in data

```{r}
# label: read-fixed

raw_data <- read_fwf(
  file="../data/fixed.txt", 
  col_names=c("x", "y"),
  col_positions = fwf_cols(1, 2),
  col_types="nn")

glimpse(raw_data)
```


#### Comments on the code

The [read_fwf][ref01] function reads data from is described in the same page that describes the read_csv  and read_tsv functions. There is a similar function, [read.fwf][ref02], in the base R package that works similarly.

Most fixed width files do not include the variable names in the first row. You can provide the actual names by supplying the col_names argument with a character vector of names.

There are several helpful functions that will allow you to specify column positions. The fwf_cols function will allow you to enter the widths of each column of data.

[ref01]: https://readr.tidyverse.org/reference/read_fwf.html
[ref02]: https://stat.ethz.ch/R-manual/R-devel/library/utils/html/read.fwf.html