---
title: "Cluster randomized trials"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

## Design

-   Designate two treatments
    -   i=1,2
    -   Possible to use three or more treatments
-   Randomly assign clusters
    -   j=1,...,b
    -   Entire cluster gets same treatment
    -   Possible to have unequal numbers within each treatment
-   Randomly select patients within clusters
    -   k=1,...,n
    -   Possible to have unequal numbers within each cluster
    
## Physical activity study, data dictionary

```{r}
#| label: 11-03-setup
#| message: false
#| warning: false

library(glue)
library(tidyverse)
library(yaml)
load("../data/demo11.RData")
```

```{r}
#| label: 11-03-pa-dictionary

f1 <- "https://raw.githubusercontent.com/pmean"
f2 <- "data/refs/heads/main/files"
f3 <- "arthritis-treatments.yaml"

# dd <- read_yaml(glue("{f1}/{f2}/{f3}"))
```

This data set is from a cluster randomized trial in Norway. Students from 57 schools were recruited for the study. 29 schools were randomly assigned to physical activity intervention and 28 schools served as a control group. The outcomes were changes in various measures of cardiometabolic health.

## Glimpse of original data

```{r}
#| label: 11-03-oa

glimpse(oa)
```

## Physical activity study, glimpse of restructured data

```{r}
#| label: 11-03-oa-2

glimpse(oa_2)
```

## Descriptive statistics

```{r}
#| label: 11-03-descriptives

oa_means
```

## Plot by subject

```{r}
#| label: 11-03-plot

oa_plot
```

## Mixed model, 1

```{r}
#| label: 11-03-mixed
#| echo: true

library(lme4)

m1 <- lmer(
  rom ~ treatment + (1 | Subject),
  data=oa_2)
```

::: notes

Information about this code is in simon-5502-11-demo

:::

## Mixed model, 2

```{}
Linear mixed model fit by REML ['lmerMod']
Formula: rom ~ treatment + (1 | Subject)
   Data: oa_2

REML criterion at convergence: 253.5

Scaled residuals: 
    Min      1Q  Median      3Q     Max 
-1.8326 -0.4790  0.1129  0.6702  1.4533 
```

-   (You can ignore this part of the output)

## Mixed model, 3

```{}
Random effects:
 Groups   Name        Variance Std.Dev.
 Subject  (Intercept) 481.7    21.95   
 Residual             301.8    17.37   
Number of obs: 30, groups:  Subject, 10
```

-   $\sigma_{between}$ = 21.95
-   $\sigma_{within}$ = 17.37
-   ICC = 481.7/(481.7+301.8)
    -   = `r round(481.7/(481.7+301.8), 2)`

## Mixed model, 4

```{}
Fixed effects:
              Estimate Std. Error t value
(Intercept)     98.400      8.851  11.117
treatmentswd     4.100      7.769   0.528
treatmenttens  -14.200      7.769  -1.828
```

-   $\bar{X}_{swd}-\bar{X}_{no}$ = 4.1
    -   t is close to zero, accept H0
-   $\bar{X}_{tens}-\bar{X}_{no}$ = -14.2
    -   Goes in the wrong direction!!!
    -   t is close to zero, accept H0

## Mixed model, 5

```{}
Correlation of Fixed Effects:
            (Intr) trtmnts
treatmntswd -0.439        
treatmnttns -0.439  0.500 
```

-   (You can ignore this part of the output)

