---
title: "Within cluster comparisons"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

## Physical activity study, data dictionary

```{r}
#| label: 11-04-setup
#| message: false
#| warning: false

library(glue)
library(tidyverse)
library(yaml)
load("../data/demo11.RData")
```

```{r}
#| label: 11-04-pa-dictionary

f1 <- "https://raw.githubusercontent.com/pmean"
f2 <- "data/refs/heads/main/files"
f3 <- "physical-activity.yaml"

dd <- read_yaml(glue("{f1}/{f2}/{f3}"))
```

`r dd$description`

## Counts by school (partial listing)

```{r}
#| label: 11-04-counts

pa_counts[1:10, ]
```

## Boxplot of change scores

```{r}
#| label: 11-04-pa-box
#| fig.height: 1.5
#| fig.width: 6

pa_box
```

## Mixed model, 1

```{r}
#| label: 11-04-mixed
#| echo: true

library(lme4)

m2 <- lmer(
  change_score ~ Group_allocation + (1 | School),
  data=pa_1)
```

::: notes

Information about this code is in simon-5502-11-demo

:::

## Mixed model, 2

```{}
Linear mixed model fit by REML ['lmerMod']
Formula: change_score ~ Group_allocation + (1 | School)
   Data: pa_1

REML criterion at convergence: 5013.3

Scaled residuals: 
    Min      1Q  Median      3Q     Max 
-4.6825 -0.5667  0.0021  0.5684  4.5686 
```

-   (You can ignore this part of the output)

## Mixed model, 3

```{}
Random effects:
 Groups   Name        Variance Std.Dev.
 School   (Intercept) 1.590    1.261   
 Residual             6.578    2.565   
Number of obs: 1043, groups:  School, 57
```

-   $\sigma_{between}$ = 1.261
-   $\sigma_{within}$ = 2.565
-   ICC = 1.590/(1.590+6.578)
    -   = `r round(1.590/(1.590+6.578), 2)`

## Mixed model, 4

```{}
Fixed effects:
                             Estimate Std. Error t value
(Intercept)                   1.34079    0.27058   4.955
Group_allocationIntervention -0.01049    0.38135  -0.028
```

-   $\bar{X}_{Intervention}-\bar{X}_{Control}$ = -0.01
    -   t is close to zero, accept H0

## Mixed model, 5

```{}
Correlation of Fixed Effects:
            (Intr)
Grp_llctnIn -0.710
```

-   (You can ignore this part of the output)



