---
title: "simon-5502-11-demo"
format: 
  html:
    embed-resources: true
editor: source
execute: 
  error: false
---

## File details

This program was written by Steve Simon on 2025-02-23 and is placed in the public domain. You can use this program any way you please.

-   Data used in this program
    -   postural-sway
        -   [tab delimited file][ref-sway]
        -   [dictionary][ref-sway-dd]

[ref-sway]: https://github.com/pmean/data/blob/main/files/postural-sway.txt
[ref-sway-dd]: https://github.com/pmean/data/blob/main/files/postural-sway.yaml

-   Tibbles created by this program
    -   sway: Original data from postural-sway.txt
-   Models created by this program
    -   m1: 

```{r}
#| message: false
#| warning: false
#| label: setup

library(broom)
library(lme4)
library(nlme)
library(tidyverse)

R.version.string
Sys.Date()
```

## Read litter-weights

```{r}
#| label: read-litter

library(foreign)
w <- read.spss(file="../data/litter-weights.sav")
w <- data.frame(w)
names(w) <- tolower(names(w))
glimpse(w)
```

```{r}
count(w, grp)
l <- unique(w$litter)
w$x <- rep(c(1:10, 12:21), each=6)
w$y <- rep(1:6, 20)
w$s <- rep(rep(c("M", "F"), each=3), 20)
```

```{r}
w |>
  ggplot() +
  aes(x, y, label=s) +
  geom_text() + 
  geom_text(x=5.5, y=6.5, label="Group 1") +
  geom_text(x=16.5, y=6.5, label="Group 2") +
  scale_x_continuous(
    name="Litter",
    breaks=unique(w$x),
    minor_breaks=NULL,
    labels=l) +
  scale_y_continuous(
    name="Animal",
    breaks=1:6, 
    minor_break=NULL) +
  expand_limits(x=c(-1, 24), y=c(0,7))
```

## Read postural-sway

```{r}
#| label: sway-read

sway <- read_tsv(
  file="../data/postural-sway.txt",
  col_types="cnn",
  col_names=TRUE)

glimpse(sway) 
```

