---
title: "simon-5502-11-demo"
format: 
  html:
    embed-resources: true
editor: source
execute: 
  error: false
---

## File details

This program was written by Steve Simon on 2025-02-23 and is placed in the public domain. You can use this program any way you please.

-   Data used in this program
    -   physical-activity
        -   [tab delimited file][ref-pa]
        -   [dictionary][ref-pa-dd]
    -   oa
        -   [tab delimited file][ref-oa]
        -   [dictionary][ref-oa-dd]
    
-   Tibbles created by this program
    -   oa: Original data from arthritis-treatments
    -   oa_1: Subset with only VAS values
    -   oa_2: Restructure to long format
    -   pa: Original data from physical-activity
    -   pa_1: Compute change score
-   Models created by this program
    -   m1: Within cluster analysis of arthritis treatments data
    -   m2: Between cluster analysis of physical activity data

[ref-oa]: https://github.com/pmean/data/blob/main/files/oa.txt
[ref-oa-dd]: https://github.com/pmean/data/blob/main/files/oa.yaml
[ref-pa]: https://github.com/pmean/data/blob/main/files/physical-activity.sav
[ref-pa-dd]: https://github.com/pmean/data/blob/main/files/physical-activity.yaml


```{r}
#| message: false
#| warning: false
#| label: setup

library(broom)
library(foreign)
library(lme4)
library(nlme)
library(tidyverse)

R.version.string
Sys.Date()
```

------------------------------Begin part 1--------------------------------------

## Read oa

```{r}
#| label: read-oa

oa <- read_table(
  file="../data/arthritis-treatments.txt",
  col_names=TRUE,
  col_types="nnnnnnn")

glimpse(oa)
```

## Create subset for ROM

```{r}
#| label: subset

oa |>
  select(
    Subject,
    NoROM,
    TENSROM,
    SWDROM) |>
  rename(
    no=NoROM,
    tens=TENSROM,
    swd=SWDROM) -> oa_1
```

## Restructure into long format

```{r}
#| label: long

oa_1 |>
  pivot_longer(
    cols=no:swd,
    names_to="treatment",
    values_to="rom") -> oa_2

glimpse(oa_2)
```

#### Comments on the code

The [pivot_longer][ref-pivot-nodate] function is in the [tidyr][ref-tidyr-nodate] package, which is part of the tidyverse.

[ref-pivot-nodate]: https://tidyr.tidyverse.org/reference/pivot_longer.html
[ref-tidyr-nodate]: https://tidyr.tidyverse.org/


## Descriptive statistics

```{r}
#| label: descriptives-oa

oa_2 |>
  group_by(treatment) |>
  summarize(
    rom_mean=mean(rom),
    rom_sd=sd(rom)) -> oa_means

oa_means
```

#### Interpretation of the output

There is a slight improvement in range of motion using swd versus no treatment. Using tens actually decreases the range of motion.

## Plot by subject

```{r}
#| label: plot-1

oa_2 |>
  ggplot() +
  aes(
    x = treatment, 
    y = rom, 
    group = Subject) +
  geom_line() -> oa_plot

oa_plot
```

#### Interpretation of the output

There is a tendency for some subjects to see the best range of motion under swd, but there are a few exceptions.

## Mixed model for oa

```{r}
#| label: mixed-oa

m1 <- lmer(
  rom ~ treatment + (1 | Subject),
  data=oa_2)

summary(m1)
```

#### Comments on the code

The [lmer][ref-lmer-nodate] function is part of the [lme4][ref-lme4-nodate] package. The packageis quite detailed, but you only need to use some basic parts of it. The lmer function uses a similar approach as the lm function. You specify a model with a formula that includes a tilde(~). The dependent variable goes on the left side of the tilde and the independent variables plus the cluster variable goes on the right side. The cluster variable needs to be defined as (1 | *cluster-variable*). This might seem a bit strange, but the syntax is required for more complex models that what you will be using, at least at the start.

The output from lmer is quite verbose and there is no easy way to select just what you want. You do want to use the summary function, which provides a list of fixed and random effects.

I have not found a simple function that will compute the intraclass correlation. It is easy enough for you to calculate by hand.

[ref-lmer-nodate]: https://cran.r-project.org/web/packages/multilevelTools/vignettes/lmer-vignette.html
[ref-lme4-nodate]: https://github.com/lme4/lme4

#### Interpretation of the output

The estimated range of motion in the reference category (treament=no) is 98 degrees. The range of motion for treatment=swd is 4 degree larger than treatment=no, which represents a very small is only slightly better result. The t-value is close to zero, so you would accept the null hypothesis and conclude that there is no statistically significant difference between the groups.

The estimated range of motion is smaller for treatment=swd than treatment=no. This is a surprising and disappointing result. The t statistic is borderline but probably needs to be larger than plus or minus two before you might consider this to be a statistically significant decline in range of motion.

There is substantial variation between clusters (subjects are the cluster). The intraclass correlation is 481.7 / (481.7 + 301.8) = 0.61. This indicates a large degree of correlation for two measurements on the same subject.

------------------------------Begin part 2--------------------------------------

## Read physical-activity

```{r}
#| label: read-pa

pa <- read.spss(
  file="../data/physical-activity.sav",
  to.data.frame=TRUE)

glimpse(pa)
```

## Descriptive statistics, 1

```{r}
#| label: stats-1

pa |>
  summarize(
    mean_pre=mean(WC_PRE, na.rm=TRUE),
    mean_post=mean(WC_POST, na.rm=TRUE)) -> pa_means

pa_means
```

#### Interpretation of the output

There is a small increase in waist circumference after the study is over. This may be due to natural growth over time.

## Descriptive statistics, 2

```{r}
#| label: stats-2

pa |>
  count(School, Group_allocation) -> pa_counts

pa_counts
```

#### Interpretation of the output

There is a lot of variation in the number of subjects per school, but every school has at least 7.


## Compute change score

```{r}
#| label: change-score

pa |>
  mutate(change_score=WC_POST - WC_PRE) -> pa_1
```

## Draw boxplot by group

```{r}
#| label: box-2
#| fig.height: 1.5
#| fig.width: 6

pa_1 |>
  filter(!is.na(change_score)) |>
  ggplot() +
  aes(
    x=Group_allocation, 
    y=change_score) +
  geom_boxplot() +
  coord_flip() +
  ggtitle("Graph drawn by Steve Simon on 2025-04-06") +
  xlab("Group") +
  ylab("Change score") -> pa_box

pa_box
```

### Interpretation of the output

The amount of increase in waist circumference is about the same in the treatment and control groups.

## Fit mixed model

```{r}
#| label: mixed

m2 <- lmer(
  change_score ~ Group_allocation + (1 | School),
  data=pa_1)

summary(m2)
```

#### Interpretation of the output

The estimated change in waist circumference in the reference category (Group_allocation=Control) is 1.34 centimeters. The estimated change in the treatment group versus the control group is -0.01 which is so very very slightly smaller that you might as well call it no difference at all. The t-statistic is close to zero indicating that this difference is not statistically significant.

There is a moderate amount of variation between clusters (schools are the cluster). The intraclass correlation is 1.590 / (1.590 + 6.578) = 0.19. This indicates a weak degree of correlation for two measurements within the same school.

## Save everything

```{r}
#| label: save

save.image(file="../data/demo11.RData")
```
