---
title: "simon-5502-11-solution"
format: docx
editor: source
execute: 
  error: false
---

## File details

This program was written by Steve Simon on 2025-03-30 and is placed in the public domain. You can use this program any way you please.

-   Data used in this program
    -    [arthritis-treatments.txt][ref-oa]
    -    [data dictionary][ref-oa-dd]

[ref-oa]: https://github.com/pmean/data/blob/main/files/arthritis-treatments.txt
[ref-oa-dd]: https://github.com/pmean/data/blob/main/files/arthritis-treatments.yaml


-   Tibbles created by this program
    -   oa: Original data from arthritis-treatments.txt
    -   oa_1: Renamed variables, exclude ROM variables
    -   oa_2: Pivot to a wide format
-   Models created by this program
    -   m1: Random effects model

```{r}
#| message: false
#| warning: false
#| label: setup

library(broom)
library(foreign)
library(lme4)
library(tidyverse)

R.version.string
Sys.Date()
```

## Read oa

```{r}
#| label: read-oa

oa <- read_table(
  file="../data/arthritis-treatments.txt",
  col_names=TRUE,
  col_types="nnnnnnn")

glimpse(oa)
```

## Create subset for VAS

```{r}
#| label: subset

oa |>
  select(
    Subject,
    NoVAS,
    TENSVAS,
    SWDVAS) |>
  rename(
    no=NoVAS,
    tens=TENSVAS,
    swd=SWDVAS) -> oa_1
```

## Restructure into long format

```{r}
#| label: long

oa_1 |>
  pivot_longer(
    cols=no:swd,
    names_to="treatment",
    values_to="vas") -> oa_2

glimpse(oa_2)
```


## Descriptive statistics

```{r}
#| label: descriptives-oa

oa_2 |>
  group_by(treatment) |>
  summarize(
    vas_mean=mean(vas),
    vas_sd=sd(vas)) -> oa_means

oa_means
```

#### Interpretation of the output

You should provide the interpretation here and after other parts of the output.

## Plot by subject

```{r}
#| label: plot-1

oa_2 |>
  ggplot() +
  aes(
    x = treatment, 
    y = vas, 
    group = Subject) +
  geom_line() +
	labs(
		x="Treatment",
		y="Pain (visual analog scale)",
		title="Line graph of pain across treatments",
		caption="Simon, 2025-04-29") -> oa_plot

oa_plot
```

## Mixed model for oa

```{r}
#| label: mixed-oa

m1 <- lmer(
  vas ~ treatment + (1 | Subject),
  data=oa_2)

summary(m1)
```
