---
title: "simon-5502-11-demo"
format: 
  html:
    embed-resources: true
editor: source
execute: 
  error: false
---

## File details

This program was written by Steve Simon on 2025-02-23 and is placed in the public domain. You can use this program any way you please.

-   Data used in this program
    -   postural-sway
        -   [tab delimited file][ref-sway]
        -   [dictionary][ref-sway-dd]

[ref-sway]: https://github.com/pmean/data/blob/main/files/postural-sway.txt
[ref-sway-dd]: https://github.com/pmean/data/blob/main/files/postural-sway.yaml

-   Tibbles created by this program
    -   sway: Original data from postural-sway.txt
-   Models created by this program
    -   m1: 

```{r}
#| message: false
#| warning: false
#| label: setup

library(broom)
library(foreign)
library(lme4)
library(nlme)
library(tidyverse)

R.version.string
Sys.Date()
```

## Read physical-activity

```{r}
#| label: read-pa

pa <- read.spss(
  file="../data/physical-activity.sav",
  to.data.frame=TRUE)

glimpse(pa)
```


## Read postural-sway

```{r}
#| label: sway-read

sway <- read_tsv(
  file="../data/postural-sway.txt",
  col_types="cnn",
  col_names=TRUE)

glimpse(sway) 
```

## Descriptive statistics

```{r}
#| label: descriptives

sway |>
  mutate(sway_difference=FBSway-SideSway) |>
  summarize(
    fb_sd=sd(FBSway),
    side_sd=sd(SideSway),
    diff_sd=sd(sway_difference)) -> sway_descriptives

with(sway, cor(FBSway, SideSway))
```

## Restructure

```{r}
#| label: restructure

sway |>
  mutate(id=1:nrow(sway)) |>
  pivot_longer(
    cols=FBSway:SideSway,
    names_to="direction",
    values_to="sway") -> sway_2

glimpse(sway_2)
```

## Plots

```{r}
#| label: boxplot

sway_2 |>
  ggplot() +
  aes(x=direction, y=sway) +
  geom_boxplot() +
  scale_y_log10()
```

```{r}
#| label: line plot

sway_2 |>
  ggplot() +
  aes(x=direction, y=sway, group=id) +
  geom_line() +
  scale_y_log10()
```



## Analyze as a paired t-test

```{r}
#| label: paired

m1 <- with(sway, t.test(x=FBSway, y=SideSway, paired=TRUE))
tidy(m1)

m2 <- lmer(sway~direction + (1 | id), data=sway_2, REML=TRUE)
summary(m2)
```

#### Comments on the code

The [t.test][ref-t-test] function will conduct three different t-tests, a single sample t-test, a two-sample t-test, and a paired t-test. I normally do not use the t.test function because I prefer to provide a comparable test using the [lm][ref-lm] function.

[ref-t-test]: https://stat.ethz.ch/R-manual/R-devel/library/stats/html/t.test.html
[ref-lm]: https://stat.ethz.ch/R-manual/R-devel/library/stats/html/lm.html

## Analyze as a two-sample t-test

```{r}
#| label: two-sample

m3 <- with(sway_2, t.test(sway ~ direction, var.equal=TRUE))
tidy(m3)

m4 <- lmer(sway~direction + (1 | id), data=sway_2, REML=TRUE)
summary(m4)
```

## Analyze as a mixed model

```{r}
#| label: mixed

m5 <- lmer(sway~Age*direction + (1 | id), data=sway_2, REML=TRUE)
summary(m5)
```

## Save everything

```{r}
#| label: save

save.image(file="../data/module11.RData")
```
