---
title: "simon-5502-11-demo"
format: 
  html:
    embed-resources: true
editor: source
execute: 
  error: false
---

## File details

This program was written by Steve Simon on 2025-02-23 and is placed in the public domain. You can use this program any way you please.

-   Data used in this program
    -   physical-activity
        -   [tab delimited file][ref-pa]
        -   [dictionary][ref-pa-dd]
    -   oa
        -   [tab delimited file][ref-oa]
        -   [dictionary][ref-oa-dd]
    
-   Tibbles created by this program
    -   oa: Original data from arthritis-treatments
    -   oa_1: Subset with only VAS values
    -   oa_2: Restructure to long format
    -   pa: Original data from physical-activity
    -   pa_1: Compute change score
-   Models created by this program
    -   m1: Within cluster analysis of arthritis treatments data
    -   m2: Between cluster analysis of physical activity data

[ref-oa]: https://github.com/pmean/data/blob/main/files/oa.txt
[ref-oa-dd]: https://github.com/pmean/data/blob/main/files/oa.yaml
[ref-pa]: https://github.com/pmean/data/blob/main/files/physical-activity.sav
[ref-pa-dd]: https://github.com/pmean/data/blob/main/files/physical-activity.yaml


```{r}
#| message: false
#| warning: false
#| label: setup

library(broom)
library(foreign)
library(lme4)
library(nlme)
library(tidyverse)

R.version.string
Sys.Date()
```

------------------------------Begin part 1--------------------------------------

## Read oa

```{r}
#| label: read-oa

oa <- read_table(
  file="../data/arthritis-treatments.txt",
  col_names=TRUE,
  col_types="nnnnnnn")

glimpse(oa)
```

## Create subset for ROM

```{r}
#| label: subset

oa |>
  select(
    Subject,
    NoROM,
    TENSROM,
    SWDROM) |>
  rename(
    no=NoROM,
    tens=TENSROM,
    swd=SWDROM) -> oa_1
```

## Restructure into long format

```{r}
#| label: long

oa_1 |>
  pivot_longer(
    cols=no:swd,
    names_to="treatment",
    values_to="rom") -> oa_2

glimpse(oa_2)
```


## Descriptive statistics

```{r}
#| label: descriptives-oa

oa_2 |>
  group_by(treatment) |>
  summarize(
    rom_mean=mean(rom),
    rom_sd=sd(rom)) -> oa_means

oa_means
```

#### Interpretation of the output

There is a slight improvement in range of motion using swd versus no treatment. Using tens actually decreases the range of motion.

## Plot by subject

```{r}
#| label: plot-1

oa_2 |>
  ggplot() +
  aes(
    x = treatment, 
    y = rom, 
    group = Subject) +
  geom_line() -> oa_plot

oa_plot
```

#### Interpretation of the output

There is a tendency for some subjects to see the best range of motion under swd, but there are a few exceptions.

## Mixed model for oa

```{r}
#| label: mixed-oa

m1 <- lmer(
  rom ~ treatment + (1 | Subject),
  data=oa_2)

summary(m1)
```

------------------------------Begin part 2--------------------------------------

## Read physical-activity

```{r}
#| label: read-pa

pa <- read.spss(
  file="../data/physical-activity.sav",
  to.data.frame=TRUE)

glimpse(pa)
```

## Descriptive statistics, 1

```{r}
#| label: stats-1

pa |>
  summarize(
    mean_pre=mean(WC_PRE, na.rm=TRUE),
    mean_post=mean(WC_POST, na.rm=TRUE)) -> pa_means

pa_means
```

#### Interpretation of the output

There is a small increase in waist circumference after the study is over. This may be due to natural growth over time.

## Descriptive statistics, 2

```{r}
#| label: stats-2

pa |>
  count(School, Group_allocation) -> pa_counts

pa_counts
```

#### Interpretation of the output

There is a lot of variation in the number of subjects per school, but every school has at least 7.


## Compute change score

```{r}
#| label: change-score

pa |>
  mutate(change_score=WC_POST - WC_PRE) -> pa_1
```

## Draw boxplot by group

```{r}
#| label: box-2
#| fig.height: 1.5
#| fig.width: 6

pa_1 |>
  filter(!is.na(change_score)) |>
  ggplot() +
  aes(
    x=Group_allocation, 
    y=change_score) +
  geom_boxplot() +
  coord_flip() +
  ggtitle("Graph drawn by Steve Simon on 2025-04-06") +
  xlab("Group") +
  ylab("Change score") -> pa_box

pa_box
```

### Interpretation of the output

The amount of increase in waist circumference is about the same in the treatment and control groups.

## Fit mixed model

```{r}
#| label: mixed

m2 <- lmer(
  change_score ~ Group_allocation + (1 | School),
  data=pa_1)

summary(m2)
```

## Save everything

```{r}
#| label: save

save.image(file="../data/demo11.RData")
```
