---
title: "simon-5502-11-demo"
format: 
  html:
    embed-resources: true
editor: source
execute: 
  error: false
---

## File details

This program was written by Steve Simon on 2025-02-23 and is placed in the public domain. You can use this program any way you please.

-   Data used in this program
    -   postural-sway
        -   [tab delimited file][ref-sway]
        -   [dictionary][ref-sway-dd]

[ref-sway]: https://github.com/pmean/data/blob/main/files/postural-sway.txt
[ref-sway-dd]: https://github.com/pmean/data/blob/main/files/postural-sway.yaml

-   Tibbles created by this program
    -   sway: Original data from postural-sway.txt
-   Models created by this program
    -   m1: 

```{r}
#| message: false
#| warning: false
#| label: setup

library(broom)
library(lme4)
library(nlme)
library(tidyverse)

R.version.string
Sys.Date()
```


## Read postural-sway

```{r}
#| label: sway-read

sway <- read_tsv(
  file="../data/postural-sway.txt",
  col_types="cnn",
  col_names=TRUE)

glimpse(sway) 
```

## Analyze as a two-sample test

```{r}
#| label: two-sample

sway |>
  mutate(id=1:nrow(sway)) |>
  select(-Age) |>
  pivot_longer(
    cols=FBSway:SideSway,
    names_to="direction",
    values_to="sway") -> sway_1

sd(sway$FBSway - sway$SideSway)

lm(sway~direction, data=sway_1) |> confint()
# lme(sway~direction + 1 | id, data=sway_1)
lmer(sway~direction + (1 | id), data=sway_1, REML=FALSE) |> summary()

t.test(sway$FBSway, sway$SideSway, var.equal=TRUE) |> tidy()
t.test(sway$FBSway, sway$SideSway, paired=TRUE, var.equal=TRUE) |> tidy()
```

## Save everything

```{r}
#| label: save

save.image(file="../data/module11.RData")
```
