---
title: "r-04-04, Barplots"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

```{r}
#| label: 04-04-setup
#| message: false
#| warning: false

library(glue)
library(tidyverse)
```

## Hypothetical datasets, 3

```{r}
#| label: 04-04-hypo-2

hypo_3 <- tibble(
  intervention=rep(c("c", "t1", "t2"), c(3, 4, 2)),
  result=rep(c("f", "s", "f", "s", "f", "s"), c(2, 1, 1, 3, 1, 1)))

hypo_3
```

## Barplot of counts, 1

```{r}
#| echo: true
#| label: 04-04-code-1

hypo_3 |>
  ggplot() +
  aes(x=intervention) +
  geom_bar() + 
  labs(
    title = "Subject counts by intervention",
    x = "Intervention (hypothetical data)",
    y = "Counts",
    caption = "Simon, 2025-04-17")  +
	coord_flip() -> bar_1
```

::: notes

I'm not a big fan of bar plots, but they sometimes have their uses. You can get a barplot for the frequency count.

Because of the way Rmarkdown displays graphs, I have to put the code on a separate slide from the graph.

:::

## Barplot of counts, 2

```{r}
#| fig.width: 6
#| fig.height: 2.5
#| label: 04-04-bar-1

bar_1
```

## Barplot of percents, 1

```{r}
#| echo: true
#| label: 04-04-code-2

pct_ticks <- seq(from = 0, to = 50, by = 10)
hypo_3 |>
  count(intervention) |>
  mutate(pct=100*n/sum(n)) |>
  ggplot() +
  aes(x=intervention, y=pct) +
  geom_col() + 
  scale_y_continuous(
    breaks=pct_ticks,
    labels=glue("{pct_ticks}%")) +
  ggtitle("Graph drawn by Steve Simon on 2025-04-17") +
  xlab("Intervention (hypothetical data)") +
  ylab("Percent") +
	coord_flip() -> bar_2
```

## Barplot of percents, 2

```{r}
#| fig.width: 6
#| fig.height: 2.5
#| label: 04-04-bar-2

bar_2
```

::: notes

Here is what the plot looks like.

:::


## Barplot with two categorical counts, 1

```{r}
#| echo: true
#| label: 04-04-code-3

hypo_3 |>
  ggplot() +
  aes(
    x=intervention, 
    fill=result) +
  geom_bar(position=position_dodge()) + 
  ggtitle("Graph drawn by Steve Simon on 2025-04-17") +
  xlab("Inteverntion (hypothetical data)") +
	coord_flip() -> bar_3
```

## Barplot with two categorical counts, 2

```{r}
#| fig.width: 6
#| fig.height: 2.5
#| label: 04-04-bar-3

bar_3
```

::: notes

Here is what the plot looks like.

:::

## Barplot with two categorical counts, 3

```{r}
#| echo: true
#| label: 04-04-code-4

hypo_3 |>
  ggplot() +
  aes(
    x=intervention, 
    fill=result) +
  geom_bar(position=position_stack()) + 
  ggtitle("Graph drawn by Steve Simon on 2025-04-17") +
  xlab("Intervention (hypothetical data)") +
	coord_flip() -> bar_4
```

## Barplot with two categorical counts, 4

```{r}
#| fig.width: 6
#| fig.height: 2.5
#| label: 04-04-bar-4

bar_4
```

::: notes

Here is what the plot looks like.

:::

## Barplot with two categorical percentages, 1

```{r}
#| echo: true
#| label: 04-04-code-5

hypo_3 |>
  count(intervention, result) |>
	group_by(result) |>
	mutate(total = sum(n)) |>
	mutate(pct = round(100*n/total)) |>
  ggplot() +
  aes(
    x = result, 
    y = pct,
    fill = intervention) +
  geom_col(position = position_stack()) + 
  ggtitle("Graph drawn by Steve Simon on 2025-04-17") +
  xlab("Intervention (hypothetical data)") +
	ylab("Percent") +
	coord_flip() -> bar_5
```

## Barplot with two categorical percentages, 2

```{r}
#| fig.width: 6
#| fig.height: 2.5
#| label: 04-04-bar-5

bar_5
```

::: notes

Here is what the plot looks like.

:::

## Barplot with two categorical percentages, 3

```{r}
#| echo: true
#| label: 04-04-code-6

hypo_3 |>
  count(intervention, result) |>
	group_by(intervention) |>
	mutate(total = sum(n)) |>
	mutate(pct = round(100*n/total)) |>
  ggplot() +
  aes(
    x = intervention, 
    y = pct,
    fill = result) +
  geom_col(position = position_stack()) + 
  ggtitle("Graph drawn by Steve Simon on 2025-04-17") +
  xlab("Intervention (hypothetical data)") + 
  ylab("Percent") +
  coord_flip() -> bar_6
```

## Barplot with two categorical percentages, 4

```{r}
#| fig.width: 6
#| fig.height: 2.5
#| label: 04-04-bar-6

bar_6
```

::: notes

Here is what the plot looks like.

:::

## Barplot with two categorical percentages, 5

```{r}
#| echo: true
#| label: 04-04-code-7

hypo_3 |>
  count(intervention, result) |>
	group_by(intervention) |>
	mutate(total = sum(n)) |>
	mutate(pct = round(100*n/total)) |>
  filter(result=="s") |>
  ggplot() +
  aes(
    x = intervention, 
    y = pct) +
  geom_col() + 
  ggtitle("Graph drawn by Steve Simon on 2025-04-17") +
  xlab("Intervention (hypothetical data)") +
  ylab("Percent") +
  coord_flip() -> bar_7
```

## Barplot with two categorical percentages, 6

```{r}
#| fig.width: 6
#| fig.height: 2.5
#| label: 04-04-bar-7

bar_7
```

::: notes

Here is what the plot looks like.

:::

