---
title: "r-04-01, Review statistics for continuous outcomes"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

## The Titanic dataset, 1

```{r}
#| label: 04-01-setup
#| message: false
#| warning: false

library(glue)
library(tidyverse)
load("../data/module-04.RData")
load("../data/demo-04.RData")
```

titanic_dictionary$description

::: notes

Here is the description of the Titanic dataset, found in the data dictionary that I created.

:::

## The Titanic dataset, 2

```{r}
#| label: 04-01-glimpse

glimpse(ti)
```

::: notes

Here are the key variables.

:::

## Peek at the bottom

`r tail`

::: notes

Comments on the code and interpretation of the output appears in simon-5505-04-demo.

:::

## Descriptive statistics on age

`r age`


## Print out information on youngest and oldest passengers

`r extremes`

## Counts and percentages, 1

## Counts and percentages, 1

```{r}
#| label: 04-01-counts-1

ti |>
  count(pclass)
```

## Counts and percentages, 2

```{r}
#| label: 04-01-counts-2

ti |>
  count(pclass) |>
  mutate(total=sum(n))   
  mutate(pct=100*n/total)
```

## Counts and percentages, 3

```{r}
#| label: 04-01-counts-3

ti |>
  count(pclass) |>
  mutate(total=sum(n)) |>
  mutate(pct=100*n/total) |>
  mutate(pct=glue("{n}/{total} ({round(pct)}%)")) |>
  select(-n, -total)
```

