---
title: "04-00, Resources needed for module 04 slides"
format: 
  html:
    embed-resources: true
editor: source
---

## Preliminaries

```{r}
#| message: false
#| warning: false
#| label: 04-00-setup

library(glue)
library(tidyverse)
```

## Read and parse simon-5505-04-demo.qmd

```{r}
#| label: 04-00-parse

fn <- "simon-5505-04-demo.qmd"

parse_code <- function(fn) {
  # Extract code from individual chunks of a Quarto or 
  # RMarkdown program.
  
  # For my programs, code always starts at s1 and ends at s2.
  # I want to display the code inside another Quarto program,
  # so I add a prefix and postfix that tells Quarto to
  # display the code without any reformatting.
  
  s1 <- fixed("#| label: ") # Beginning boundary
  s2 <- "^```$"             # Ending boundary

  prefix <- "```{}"         # Goes before R chunk
  postfix <- "```"          # Goes after R chunk

  demo_text <- read_lines(fn)

  t1 <- str_which(demo_text, s1)
  t2 <- str_which(demo_text, s2)
  
  # I should check here that the lengths of t1 and t2 are the same.

  code <- NULL

  for (i in 1:length(t1)) {
    # Find the label value for a chunk and store it in v.
    demo_text[t1[i]] |>
      str_remove(s1) |>
      str_replace_all(fixed("-"), "_") -> v
    
    # Select two lines after s1 and one line before s2
    # and add prefix and postfix. Store in a list with
    # the name stored in v.
    r <- (t1[i]+2):(t2[i]-1) 
    prefix |>
      append(demo_text[r]) |>
      append(postfix) |>
      paste0(collapse="\n") -> code[v]
  }
  # Display the results here as a quality check.
  for (v in names(code)) {
    cat("\n----------")
    cat(v)
    cat("----------\n")
    cat(code[v])
  } 
  return(code)
}
```

```{r}
#| label: parse

code <- parse_code("simon-5505-04-demo.qmd")
```

## Save everything important

```{r save}
save(code, file="../data/module-04.RData")
```
