---
title: "r-04-01, The Titanic dataset"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

## The Titanic dataset, 1

```{r}
#| label: r-04-01-setup
#| message: false
#| warning: false

library(tidyverse)
load("../data/module04.RData")
```

`r titanic_dictionary$description`

::: notes

Here is the description of the Titanic dataset, found in the data dictionary that I created.

:::

## The Titanic dataset, 2

```{r}
#| label: 04-01-glimpse

glimpse(ti)
```

::: notes

Here are the key variables.

:::


## Descriptive statistics on age

```{r}
#| label: 04-01-age-1

ti |>
  summarize(
    age_mean=mean(age, na.rm=TRUE),
    age_sd=sd(age, na.rm=TRUE),
    age_min=min(age, na.rm=TRUE),
    age_max=max(age, na.rm=TRUE),
    missing=sum(is.na(age))
  )

```

::: notes

You learned how to get descriptive statistics for continuous variables in a previous module. The youngest passenger is `r min(ti$Age, na.rm=TRUE)` years old and the oldest is `r max(ti$Age, na.rm=TRUE)` years old. Let's peek at their data values.

:::

## Print out information on youngest

```{r min}
ti |>
  slice_min(Age, na.rm=TRUE)
```

::: notes

:::

## Print out information on oldest

```{r max}
ti |>
  slice_max(Age, na.rm=TRUE)
```

::: notes

Here is the oldest passenger. It turns out that the little baby survived, but not the old man.

:::

## Break #1
+ What have you learned
  + Reading in the Titanic dataset
+ What's next
  + Categorical versus continuous variables
  + Counts, proportions, and percentage

