---
title: "Examples from module03"
author: "Steve Simon"
date: "Created 2022-03-09"
output: html_document
---

This is a small program written in Rmarkdown to show how to read in some simple text datasets.

This program is in the public domain. You can use it any way that you please.

```{r}
#| label: version-and-current-date
#| message: false
#| warning: false

R.version.string
Sys.Date()
library(tidyverse)
```

This shows how to read in a comma delimited file.

```{r read-csv}
#| label: read-csv

fn <- "../data/simple.csv"
raw_data <- read_csv(file=fn, col_types="nn")
raw_data
```

This shows how to read in a space delimited file.

```{r read-space}
#| label: read-space

fn <- "../data/simple.txt"
raw_data <- read_delim(
  file=fn, 
  delim=" ",
  col_types="nn")
raw_data
```

This shows how to read in a tab delimited file.

```{r}
#| label: read-tsv

fn <- "../data/simple.tsv"
raw_data <- read_tsv(file=fn, col_types="nn")
raw_data
```

This shows how to read in a tilde delimited file.

```{r}
#| label: read-tilde

fn <- "../data/tilde.txt"
raw_data <- read_delim(
  file=fn, 
  delim="~",
  col_types="nn")
raw_data
```

This shows how to read in a fixed-width file.

```{r}
#| label:  read-fixed-width

fn <- "../data/fixed.txt"
raw_data <- read_fwf(fn, 
  fwf_cols(1, 2),
  col_types="nn")
raw_data
```

This shows how to read in string data.

```{r}
#| label: string-data

fn <- "../data/string.csv"
raw_data <- read_csv(fn,
  col_types="ccn")
raw_data
```

This shows how to store a data frame as a text file.

```{r store-data}
raw_data <- data.frame(
  x=c(1, 2, 3, 4),
  y=c(4, 8, 12, 16))
fn <- "../results/output_data.txt"
write_csv(raw_data, fn)
```

This shows how to read the airport data.

```{r read-airport}
fn <- "../data/airport.txt"
raw_data <- read_fwf(fn,
  fwf_cols(21, 22, 6, 7, 9, 10, 10),
  col_types="ccnnnnn")
names(raw_data) <- c(
  "airport", 
  "city", 
  "scheduled_departures", 
  "performed_departures", 
  "enplaned_passengers", 
  "enplaned_freight",
  "enplaned_mail")
raw_data
```
