---
title: "Interrupted time series designs"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

## Interrupted time series design

$\ $

`NR E: O1 O2 O3 X O4 O5 O6`

$\ $

-   Best with three or more measures at baseline
-   Check for most temporal trends

::: notes

If you have the option of taking multiple measurements over time, then this can help avoid the problems with using a single group.

If the measurements are flat from O1 through O3, show a jump from O3 to O4, and then becomes flat again from O4 through O6, then you have pretty good evidence against most temporal trends, except for a temporal trend with the bad luck of occurring right at the time of your intervention.

If you see a jump between O3 and O4, but you also see jumps at most of the other time points, that's bad news.

:::

## Hypothetical patterns in the interrupted time series design

```{r time-series-2, echo=FALSE, comment=""}
x <- 1:6
y1 <- c(1.1, 1.3, 1.2, 2.2, 2.1, 2.3)
y2 <- c(1.1, 1.4, 1.3, 1.8, 2.2, 2.3)
par(mar=c(2.6, 0.6, 1.6, 0.1), mfrow=c(2, 1))
plot(x, y1, axes=FALSE, type="b", ylim=c(0.9, 2.4))
text(3.5, 1, "X")
segments(3.5, 1.1, 3.5, 2.4, lty="dotted", col="gray")
axis(side=1, at=1:6, labels=paste0("O", 1:6))
title("Good news")
plot(x, y2, axes=FALSE, type="b", ylim=c(0.9, 2.4))
text(3.5, 1, "X")
segments(3.5, 1.1, 3.5, 2.4, lty="dotted", col="gray")
axis(side=1, at=1:6, labels=paste0("O", 1:6))
title("Bad news")
```

## Phased design, 1 of 2

$\ $

`NR E: O1 O2 X1 O3 O4 O5 X2 O6 O7 O8 X3 O9 O10`

$\ $

-   Split intervention into three or more pieces
-   Phase in the intervention piece by piece

::: notes

:::

## Phased design, 2 of 2

```{r time-series-4, echo=FALSE, comment=""}
x <- 1:10
y1 <- c(1.1, 1.3, 2.4, 2.2, 2.3, 3.1, 3.4, 3.2, 4.3, 4.2)
y2 <- c(1.1, 1.7, 1.6, 2.1, 2.7, 2.9, 3.3, 3.9, 4.3, 4.2)
par(mar=c(2.6, 0.6, 1.6, 0.1), mfrow=c(2, 1))
plot(x, y1, axes=FALSE, type="b", ylim=c(0.7, 4.4))
i1 <- c(2.5, 5.5, 8.5) 
text(i1, 1, paste0("X", 1:3))
segments(i1, 1.1, i1, 4.4, lty="dotted", col="gray")
axis(side=1, at=1:10, labels=paste0("O", 1:10))
title("Good news")
plot(x, y2, axes=FALSE, type="b", ylim=c(0.7, 4.4))
text(i1, 1, paste0("X", 1:3))
segments(i1, 1.1, i1, 4.4, lty="dotted", col="gray")
axis(side=1, at=1:10, labels=paste0("O", 1:10))
title("Bad news")
```

::: notes

:::

## Stepped wedge design, 1 of 2

$\ $

`NR E1: O1 O2 X O3 O4 O5   O6 O7 O8   O9 O10`

`NR E2: O1 O2   O3 O4 O5 X O6 O7 O8   O9 O10`

`NR E3: O1 O2   O3 O4 O5   O6 O7 O8 X O9 O10` 

$\ $

-   Wait for your turn.
-   Useful for very small sample sizes.

::: notes
If you can control the timing of an intervention and you can apply the intervention at different times to different patients, then you have an extra bit of evidence to rule out temporal trends as a competing explanation. When each patient sees a shift at the specific time that he/she experiences the intervention, and is flat at the remaining times, then you have assurances that a temporal trend is not the cause. 

This is a great research design if you have a very small sample size. It works well, for example, if you are implementing a change on three different floors of a hospital, for exam, and it is actually easier in many settings to stagger the interventions rather than run them all at the same time.
:::

## Stepped wedge design, 2 of 2

```{r time-series-6, echo=FALSE, comment=""}
x <- 1:10
y1 <- c(1.1, 1.3, 2.4, 2.2, 2.3, 2.1, 2.4, 2.2, 2.3, 2.2)
y2 <- c(1.2, 1.1, 1.4, 1.1, 1.3, 2.2, 2.3, 2.4, 2.1, 2.4)
y3 <- c(1.3, 1.2, 1.1, 1.3, 1.2, 1.2, 1.3, 1.2, 2.2, 2.3)
par(mar=c(2.6, 0.6, 1.6, 0.1), mfrow=c(3, 1))
plot(x, y1, axes=FALSE, type="b", ylim=c(0.7, 2.5))
text(2.5, 1, "X")
segments(2.5, 1.1, 2.5, 2.5, lty="dotted", col="gray")
axis(side=1, at=1:10, labels=paste0("O", 1:10))
title("First patient")
plot(x, y2, axes=FALSE, type="b", ylim=c(0.7, 2.5))
text(5.5, 1, "X")
segments(5.5, 1.1, 5.5, 2.5, lty="dotted", col="gray")
axis(side=1, at=1:10, labels=paste0("O", 1:10))
title("Second patient")
plot(x, y3, axes=FALSE, type="b", ylim=c(0.7, 2.5))
text(8.5, 1, "X")
segments(8.5, 1.1, 8.5, 2.5, lty="dotted", col="gray")
axis(side=1, at=1:10, labels=paste0("O", 1:10))
title("Third patient")
```

::: notes
The flatness for patients 2 and 3 early rule out the possibility of a temporal trend as an explanation for the jump in patient 1 and the flatness for patients 1 and 2 late rule out the possibility of a temporal trend as an explanation for the jump in patient 3.
:::