---
title: "09-02, One-tailed tests"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

## One-tailed tests

-   $H_0:\ \mu_1 - \mu_2 = 0$
-   $H_1:\ \mu_1 - \mu_2 > 0$
    -   Accept $H_0$ if $\bar{X}_1-\bar{X}_2$ is large negative
    -   Accept $H_0$ if $\bar{X}_1-\bar{X}_2$ is close to zero
    -   Reject $H_0$ if $\bar{X}_1-\bar{X}_2$ is large positive
-   More precisely,
    -   Accept $H_0$ if $T \le t(1-\alpha, df)$
    -   Reject $H_0$ if $T \gt t(1-\alpha, df)$

:::notes
Sometimes you are only interested in changes in a positive direction. This means the first population mean is larger than the second population mean.
:::

## Critical value for a one tailed test
    
```{r 09-02-critical-values-1}
#| message: false
#| warning: false
library(tidyverse)
x <- seq(-6, 6, length=100)
y <- dt(x, 4)

data.frame(x, y) |>
  ggplot(aes(x, y)) + 
    geom_line(
      color="black",
      size=1) +
    expand_limits(y=0.4) +
    ggtitle("Graph drawn by Steve Simon on 2024-10-06") -> t_curve
a <- qt(0.95, 4)
eq0 <- expression(paste("Area = 1 - ", alpha))
eq1 <- expression(paste("t(", alpha, ", df)"))
eq2 <- expression(paste("t(1-", alpha, ", df)"))
eq3 <- expression(paste("Area = ", alpha))
t_curve +
  geom_segment(x= a, y=0, xend= a, yend=0.38) +
  geom_label(x=0, y=0, label=eq0) +
  geom_label(x= a, y=0.4, label=eq2) +
  geom_label(x= a+0.9, y=0, label=eq3)
```


:::notes
The area to the right of the critical value is $\alpha$ and the area to the left is 1-$\alpha$.
:::

## One-tailed tests

-   $H_0:\ \mu_1 - \mu_2 = 0$
-   $H_1:\ \mu_1 - \mu_2 < 0$
    -   Reject $H_0$ if $\bar{X}_1-\bar{X}_2$ is large negative
    -   Accept $H_0$ if $\bar{X}_1-\bar{X}_2$ is close to zero
    -   Accept $H_0$ if $\bar{X}_1-\bar{X}_2$ is large positive
-   More precisely,
    -   Accept $H_0$ if $T \ge t(\alpha, df)$
    -   Reject $H_0$ if $T \lt t(\alpha, df)$

:::notes
There is a second one-sided hypothesis. This tests the hypothesis that the difference is negative, or equivalently that the first population mean is smaller than the second population mean.
:::

## Critical value
  
```{r 09-02-critical-values-2}
data.frame(x, y) |>
  ggplot(aes(x, y)) + 
    geom_line(
      color="black",
      size=1) +
    expand_limits(y=0.4) +
    ggtitle("Graph drawn by Steve Simon on 2024-10-06") -> t_curve
a <- qt(0.05, 4)
t_curve +
  geom_segment(x=a, y=0, xend=a, yend=0.38) +
  geom_label(x=0, y=0, label=eq0) +
  geom_label(x=a, y=0.4, label=eq1) +
  geom_label(x=a-0.9, y=0, label=eq3)
```

:::notes
The area to the left is $\alpha$ and the area to the right is 1-$\alpha$.
:::

## Calculation of the p-value

-   For testing $H_1:\ \mu_1 - \mu_2 \ne 0$
    -   p-value = $2 P[t(n_1+n_2-2) > |T|]$
-   For testing $H_1:\ \mu_1 - \mu_2 > 0$
    -   p-value = $P[t(n_1+n_2-2) > T]$    
-   For testing $H_1:\ \mu_1 - \mu_2 < 0$
    -   p-value = $P[t(n_1+n_2-2) < T]$        
-   For any of these hypothesis,
    -   Accept $H_0$ if p-value > $\alpha$
    
:::notes
The calculation of the p-value changes slightly if you have a one-sided hypothesis. But the decision rule is always the same, accept the null hypothesis if the p-value is large, larger than alpha.
:::

## R code for one-sided hypotheses

-   For testing $H_1:\ \mu_1 - \mu_2 \ne 0$
     -   `alternative="two-sided"`
-   For testing $H_1:\ \mu_1 - \mu_2 > 0$
    -   `alternative="greater"`
-   For testing $H_1:\ \mu_1 - \mu_2 < 0$
    -   `alternative="less"`

:::notes
The t.test function in R has an argument, alternative. The default is "two-sided". Change this to "greater" or "less" to produce the appropriate one-sided hypothesis.
:::
