---
title: "Paired t-test for postural sway data"
format: 
  html:
    embed-resources: true
---

This program runs a paired t-test. Consult the [data dictionary][dic] for information about the data itself.

[dic]: https://github.com/pmean/data/blob/main/files/postural-sway.yaml

This program was written by Steve Simon on 2024-10-15. It is placed in the public domain.

## Libraries

```{r setup}
#| message: false
#| warning: false
library(tidyverse)
```

## Read data

```{r read-sway}
sway <- read_tsv(
  file="../data/postural-sway.txt",
  col_types="cnn")
names(sway) <- tolower(names(sway))
glimpse(sway)
```

## Paired t-test

The research team is also interested in whether the front-to-back sway is larger than the side-to-side sway, combining both the elderly and young patients into one group.

## Compute and graph differences

```{r compute-difference-1}
sway |>
  mutate(diff_sway=fbsway-sidesway) -> paired_differences
paired_differences |>
    summarize(
      diff_mn=mean(diff_sway),
      diff_sd=sd(diff_sway))
```

The average difference is positive, indicating that front-to-back-sways are larger on average. The standard deviation of the differences is large. 

## Boxplot of differences

```{r compute-difference-2}
#| fig.height: 1.5
#| fig.width: 6
paired_differences |>
  ggplot(aes(diff_sway, "Combined")) +
    geom_boxplot() +
    ggtitle("Graph drawn by Steve Simon on 2024-10-13") +
    xlab("Front-to-back sway minus side-to-side sway") +
    ylab(" ")
```

More than 75% of the differences are positive, also indicating that front-to-back sways tend to be larger.

## Paired t-test

```{r paired-t-test}
t.test(
  sway$fbsway, 
  sway$sidesway,
  paired=TRUE,
  alternative="two.sided")
```
The p-value is small, indicating that front-to-back sways are significantly larger than side-to-side sways. The 95% confidence intervals shows that the mean difference is at least 0.21 units larger for front-to-back sway and possibly as large as 7.0 units.