---
title: "Alternative analysis of postural sway data"
format: 
  html:
    embed-resources: true
---

This program runs some alternatives to the two-sample t-test. Consult the [data dictionary][dic] for information about the data itself.

[dic]: https://github.com/pmean/data/blob/main/files/postural-sway.yaml

This program was written by Steve Simon on 2024-10-13. It is placed in the public domain.


## Libraries

```{r setup}
#| message: false
#| warning: false
library(tidyverse)
```

## Read data

```{r read-sway}
sway <- read_tsv(
  file="../data/postural-sway.txt",
  col_types="cnn")
names(sway) <- tolower(names(sway))
glimpse(sway)
```

## Boxplot of front-to-back sway by age

```{r boxplot-1}
#| fig.height: 2
#| fig.width: 6
sway |>
  ggplot(aes(age, fbsway)) +
    geom_boxplot() +
    ggtitle("Graph drawn by Steve Simon on 2024-10-13") +
    xlab("Treatment group") +
    ylab("Front to back sway") +
    coord_flip()
```

The outlier causes some concern about the validity of the two-sample t-test.

## Descriptive statistics for front-to-back sway by age

```{r group-means}
sway |>
  group_by(age) |>
  summarize(
    fb_mn=mean(fbsway),
    fb_sd=sd(fbsway),
    n=n())
```

In addition to the outlier, notice that the group with the larger mean (elderly) has the larger standard deviation. This indicates that a log transformation may produce better results.

## Log transformation, 1

```{r log-transform-1}
sway |>
  mutate(log_fbsway=log10(fbsway)) -> log_sway
```

## Log transformation, 2

```{r log-transform-2}
#| fig.height: 2
#| fig.width: 6
log_sway |>
  ggplot(aes(age, log_fbsway)) +
    geom_boxplot() +
    ggtitle("Graph drawn by Steve Simon on 2024-10-13") +
    xlab("Treatment group") +
    ylab("Front to back sway") +
    coord_flip()
```

There are no outliers after a log transformation.

## Log transformation, 3

```{r compare-means-on-log-scale}
log_sway |>
  group_by(age) |>
  summarize(
    log_mn=mean(log_fbsway),
    log_sd=sd(log_fbsway),
    n=n())
```

The standard deviations on the log scale are quite a bit more similar than they were on the original scale.

## Two-sample t-test using the log transformation

```{r t-test}
m2 <- t.test(
  log_fbsway ~ age, 
  data=log_sway,
  alternative="two.sided",
  var.equal=TRUE)
m2
```

There is a statistically significant difference between the log front-to-back sway between elderly patients and young patients. The confidence interval will be interpreted after transforming back to the original scale of measurement.

## Back-transform confidence interval to the original scale.

```{r back-transform}
10^(m2$conf.int)
```

We are 95% confident that the geometric mean front-to-back sway in elderly patients is somewhere between 1.24 times higher and 8 times higher than the geometric mean for young patients. This indicates a statistically significantly higher mean for elderly patients. The confidence interval is still very wide, indicating a lot of sampling error in these two small samples.

## Wilcoxon-Mann-Whitney

```{r wmw-test}
wilcox.test(fbsway ~ age, data=sway)
```

Since the p-value is small, you would reject the null hypothesis and conclude that there is a statistically significant difference in front-to-back sway values between elderly and young patients.
