---
title: "06-02, R code for diagnostic plots"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

## albuquerque-housing, 1

```{r data1}
load("../data/module06.RData")

cat(datac$chunk[datac$lab=="data_dictionary"])
cat(datac$chunk[datac$lab=="format"])
cat(datac$chunk[datac$lab=="varnames"])
cat(datac$chunk[datac$lab=="missing_value_code"])
```

## albuquerque-housing, 2

```{r data2}
cat(datac$chunk[datac$lab=="description"])
cat(datac$chunk[datac$lab=="download_url"])
```

## albuquerque-housing, 3

```{r data3}
cat(datac$chunk[datac$lab=="source"])
cat(datac$chunk[datac$lab=="copyright"])
cat(datac$chunk[datac$lab=="size"])
```

## albuquerque-housing, 4

```{r data4}
cat(datav$chunk[1])
cat(datav$chunk[2])
cat(datav$chunk[3])
```

## albuquerque-housing, 5

```{r data5}
cat(datav$chunk[4])
cat(datav$chunk[5])
```

## albuquerque-housing, 6

```{r data6}
cat(datav$chunk[6])
cat(datav$chunk[7])
```

## simon-5501-06-albuquerque.qmd, 1

```{r write1}
cat(codec[1])
```

:::notes
The first few lines are the documentation header
:::

## simon-5501-06-albuquerque.qmd, 2

```{r write2}
cat(codec[2])
```

## simon-5501-06-albuquerque.qmd, 3

```{r write3}
cat(codec[3])
```

## simon-5501-06-albuquerque.qmd, 4

```{r write4}
cat(codec[4])
```

## simon-5501-06-albuquerque.qmd, 5

```{r write5}
cat(codec[5])
```

## simon-5501-06-albuquerque.qmd, 6

```{r write6}
cat(codec[6])
```

## simon-5501-06-albuquerque.qmd, 7

```{r write7}
cat(codec[7])
```

## simon-5501-06-albuquerque.qmd, 8

```{r write8}
cat(codec[8])
```

## simon-5501-06-albuquerque.qmd, 9

```{r write9}
cat(codec[9])
```

## simon-5501-06-albuquerque.qmd, 10

```{r write10}
cat(codec[10])
```

## simon-5501-06-albuquerque.qmd, 11

```{r write11}
cat(codec[11])
```

## simon-5501-06-albuquerque.qmd, 12

```{r write12}
cat(codec[12])
```

## simon-5501-06-albuquerque.qmd, 13

```{r write13}
cat(codec[13])
```
