---
title: "06-02, R code for diagnostic plots"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

## albuquerque-housing, 1 of 6

```{r data1}
library(yaml)
fn <- "https://raw.githubusercontent.com/pmean/data/main/files/albuquerque-housing.yaml"
dd <- read_yaml(fn)
part1 <- c(
  "data_dictionary",
  "format",
  "varnames",
  "missing_value_code")
for (i in part1) {
  cat(as.yaml(dd[i]))
	cat("\n")
}
```

## albuquerque-housing, 2 of 6

```{r data2}
part2 <- c(
  "description",
  "download_url")
for (i in part2) {
  cat(as.yaml(dd[i]))
	cat("\n")
}
```

## albuquerque-housing, 3 of 6

```{r data3}
part3 <- c(
  "source",
  "copyright",
  "size")
for (i in part3) {
  cat(as.yaml(dd[i]))
	cat("\n")
}
```

## albuquerque-housing, 4 of 6

```{r data4}
cat(as.yaml(dd[["vars"]][1:3]))
```

## albuquerque-housing, 5 of 6

```{r data5}
cat(as.yaml(dd[["vars"]][4:5]))
```

## albuquerque-housing, 6 of 6

```{r data6}
cat(as.yaml(dd[["vars"]][6:7]))
```

## simon-5501-06-albuquerque.qmd, `r (i <- 1)`

```{r write1}
library(tidyverse)
fn <- "simon-5501-06-albuquerque.qmd"
template <- read_lines(file=fn)
h <- str_which(template, "##")
chunk_beg <- c(1, h)
chunk_end <- c(h-1, length(template))
```

```{r}
template[chunk_beg[i]:chunk_end[i]] |>
	paste0(collapse="\n") |>
	cat()
```

:::notes
The first few lines are the documentation header
:::

## simon-5501-06-albuquerque.qmd, `r (i <- 2)`

```{r}
template[chunk_beg[i]:chunk_end[i]] |>
	paste0(collapse="\n") |>
	cat()
```


## simon-5501-06-albuquerque.qmd, `r (i <- 3)`

```{r}
template[chunk_beg[i]:chunk_end[i]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-albuquerque.qmd, `r (i <- 4)`

```{r}
template[chunk_beg[i]:chunk_end[i]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-albuquerque.qmd, `r (i <- 5)`

```{r}
template[chunk_beg[i]:chunk_end[i]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-albuquerque.qmd, `r (i <- 6)`

```{r}
template[chunk_beg[i]:chunk_end[i]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-albuquerque.qmd, `r (i <- 7)`

```{r}
template[chunk_beg[i]:chunk_end[i]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-albuquerque.qmd, `r (i <- 8)`

```{r}
template[chunk_beg[i]:chunk_end[i]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-albuquerque.qmd, `r (i <- 9)`

```{r}
template[chunk_beg[i]:chunk_end[i]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-albuquerque.qmd, `r (i <- 10)`

```{r}
template[chunk_beg[i]:chunk_end[i]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-albuquerque.qmd, `r (i <- 11)`

```{r}
template[chunk_beg[i]:chunk_end[i]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-albuquerque.qmd, `r (i <- 12)`

```{r}
template[chunk_beg[i]:chunk_end[i]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-albuquerque.qmd, `r (i <- 13)`

```{r}
template[chunk_beg[i]:chunk_end[i]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-albuquerque.qmd, `r (i <- 14)`

```{r}
template[chunk_beg[i]:chunk_end[i]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-albuquerque.qmd, `r (i <- 15)`

```{r}
template[chunk_beg[i]:chunk_end[i]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-albuquerque.qmd, `r (i <- 16)`

```{r}
template[chunk_beg[i]:chunk_end[i]] |>
	paste0(collapse="\n") |>
	cat()
```

