---
title: "06-02, R code for diagnostic plots"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

## albuquerque-housing, 1 of 6

```{r data1}
library(yaml)
fn <- "https://raw.githubusercontent.com/pmean/data/main/files/albuquerque-housing.yaml"
dd <- read_yaml(fn)
part1 <- c(
  "data_dictionary",
  "format",
  "varnames",
  "missing_value_code")
for (i in part1) {
  cat(as.yaml(dd[i]))
	cat("\n")
}
```

## albuquerque-housing, 2 of 6

```{r data2}
part2 <- c(
  "description",
  "download_url")
for (i in part2) {
  cat(as.yaml(dd[i]))
	cat("\n")
}
```

## albuquerque-housing, 3 of 6

```{r data3}
part3 <- c(
  "source",
  "copyright",
  "size")
for (i in part3) {
  cat(as.yaml(dd[i]))
	cat("\n")
}
```

## albuquerque-housing, 4 of 6

```{r data4}
cat(as.yaml(dd[["vars"]][1:3]))
```

## albuquerque-housing, 5 of 6

```{r data5}
cat(as.yaml(dd[["vars"]][4:5]))
```

## albuquerque-housing, 6 of 6

```{r data6}
cat(as.yaml(dd[["vars"]][6:7]))
```

## simon-5501-06-albuquerque.qmd, 1

```{r write1}
library(tidyverse)
fn <- "simon-5501-06-albuquerque.qmd"
template <- read_lines(file=fn)
h <- str_which(template, "##")
g <- h-1
template[1:g[1]] |>
	paste0(collapse="\n") |>
	cat()
```

:::notes
The first few lines are the documentation header
:::

## simon-5501-06-albuquerque.qmd, 2

```{r write2}
template[h[1]:g[2]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-albuquerque.qmd, 3

```{r write3}
template[h[2]:g[3]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-albuquerque.qmd, 4

```{r write4}
template[h[3]:g[4]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-albuquerque.qmd, 5

```{r write5}
template[h[4]:g[5]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-albuquerque.qmd, 6

```{r write6}
template[h[5]:g[6]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-albuquerque.qmd, 7

```{r write7}
template[h[6]:g[7]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-albuquerque.qmd, 8

```{r write8}
template[h[7]:g[8]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-albuquerque.qmd, 9

```{r write9}
template[h[8]:g[9]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-albuquerque.qmd, 10

```{r write10}
template[h[9]:g[10]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-albuquerque.qmd, 11

```{r write11}
template[h[10]:g[11]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-albuquerque.qmd, 12

```{r write12}
template[h[11]:g[12]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-albuquerque.qmd, 13

```{r write13}
template[h[12]:g[13]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-albuquerque.qmd, 14

```{r write14}
template[h[13]:g[14]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-albuquerque.qmd, 15

```{r write15}
template[h[14]:g[15]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-albuquerque.qmd, 16

```{r write16}
template[h[15]:g[16]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-albuquerque.qmd, 17

```{r write17}
template[h[16]:g[17]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-albuquerque.qmd, 18

```{r write18}
template[h[10]:g[11]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-albuquerque.qmd, 19

```{r write19}
template[h[18]:g[19]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-albuquerque.qmd, 20

```{r write20}
template[h[19]:g[20]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-albuquerque.qmd, 21

```{r write21}
template[h[20]:length(template)] |>
	paste0(collapse="\n") |>
	cat()
```

