---
title: "06-07, Your homework"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

## simon-5501-06-directions.qmd, `r (i <- 1)`

```{r read}
library(tidyverse)
fn <- "simon-5501-06-directions.md"
template <- read_lines(file=fn)
h <- str_which(template, "##")
chunk_beg <- c(1, h)
chunk_end <- c(h-1, length(template))
```

```{r}
template[chunk_beg[i]:chunk_end[i]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-directions.qmd, `r (i <- i+1)`

```{r}
template[chunk_beg[i]:chunk_end[i]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-directions.qmd, `r (i <- i+1)`

```{r}
template[chunk_beg[i]:chunk_end[i]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-directions.qmd, `r (i <- i+1)`

```{r}
template[chunk_beg[i]:chunk_end[i]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-directions.qmd, `r (i <- i+1)`

```{r}
template[chunk_beg[i]:chunk_end[i]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-directions.qmd, `r (i <- i+1)`

```{r}
template[chunk_beg[i]:chunk_end[i]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-directions.qmd, `r (i <- i+1)`

```{r}
template[chunk_beg[i]:chunk_end[i]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-directions.qmd, `r (i <- i+1)`

```{r}
template[chunk_beg[i]:chunk_end[i]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-directions.qmd, `r (i <- i+1)`

```{r}
template[chunk_beg[i]:chunk_end[i]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-directions.qmd, `r (i <- i+1)`

```{r}
template[chunk_beg[i]:chunk_end[i]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-directions.qmd, `r (i <- i+1)`

```{r}
template[chunk_beg[i]:chunk_end[i]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-directions.qmd, `r (i <- i+1)`

```{r}
template[chunk_beg[i]:chunk_end[i]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-directions.qmd, `r (i <- i+1)`

```{r}
template[chunk_beg[i]:chunk_end[i]] |>
	paste0(collapse="\n") |>
	cat()
```

## simon-5501-06-directions.qmd, `r (i <- i+1)`

```{r}
template[chunk_beg[i]:chunk_end[i]] |>
	paste0(collapse="\n") |>
	cat()
```

