---
title: "02-02, Computing counts and percentages using R"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
execute:
  echo: false
  message: false
  warning: false
---

## Data dictionary for sharing, 1

```{r data0a}
library(glue)
library(readr)
library(stringr)
library(yaml)
f1 <- "https://raw.githubusercontent.com/pmean"
f2 <- "data/main/files/sharing.yaml"
dd <- read_lines(glue("{f1}/{f2}"))
# if I am not online use the line below
# dd <- read_lines("../../../../data/files/sharing.yaml")
h <- str_which(str_trim(dd), "^$")
chunk_beg <- c(1, h+1)
chunk_end <- c(h, length(dd))
```

```{r data1a}
cat(dd[chunk_beg[1]:chunk_end[1]], sep="\n")
```

```{r data2a}
cat(dd[chunk_beg[2]:chunk_end[2]], sep="\n")
```

:::notes
This data comes from a paper on sharing services: sharing cars, sharing bikes, even sharing housing. It has a large number of variables, but for this module, we are only interested in some basic demographics.
:::

## Data dictionary for sharing, 2

```{r data3a}
cat(dd[chunk_beg[3]:chunk_end[3]], sep="\n")
```

:::notes
The bulk of the data is a series of Likert scale items asking the degree of agreement or disagreement to various statements about sharing services.
:::

## Data dictionary for sharing, 3

```{r data4a}
cat(dd[chunk_beg[4]:chunk_end[4]], sep="\n")
```

:::notes
This is rare and greatly appreciated. You have information about the conditions under which you can use the data. This is a common open source license that allows you to use the dataset as long as you give appropriate credit.
:::

## Data dictionary for sharing, 4

```{r data5a}
cat(dd[chunk_beg[5]:chunk_end[5]], sep="\n")
```

```{r data6a}
cat(dd[chunk_beg[6]:chunk_end[6]], sep="\n")
```

```{r data7a}
cat(dd[chunk_beg[7]:chunk_end[7]], sep="\n")
```

```{r data8a}
cat(dd[chunk_beg[8]:chunk_end[8]], sep="\n")
```

:::notes
This is an Excel file with lots of columns. You will not be asked to analyze all 31 columns of data, just 3 related to basic demographics.
:::

## Data dictionary for sharing, 5

```{r data12a}
cat(dd[chunk_beg[13]:chunk_end[13]], sep="\n")
```

:::notes
This is the first demographic variable.
:::

## Data dictionary for sharing, 6

```{r data13a}
cat(dd[chunk_beg[14]:chunk_end[14]], sep="\n")
```

:::notes
This is the second demographic variable.
:::

## Data dictionary for sharing, 7

```{r data14a}
cat(dd[chunk_beg[15]:chunk_end[15]], sep="\n")
```

:::notes
This is the third demographic variable.
:::

## simon-5501-02-sharing.qmd, 1

```{r write0a}
fn <- "simon-5501-02-sharing.qmd"
template <- read_lines(file=fn)
h <- str_which(template, "##")
chunk_beg <- c(1, h)
chunk_end <- c(h-1, length(template))
```

```{r write1a}
cat(template[chunk_beg[1]:chunk_end[1]], sep="\n")
```

:::notes
Here is a program template that illustrates how to calculate counts and percentages in R. You will make minor additions to this program in your programming assignment.

The first few lines are the documentation header.
:::

## simon-5501-01-template.qmd, 2

```{r write2a}
cat(template[chunk_beg[2]:chunk_end[2]], sep="\n")
```

:::notes
Here is a bit of additional documentation.
:::

## simon-5501-01-template.qmd, 3

```{r write3a}
cat(template[chunk_beg[3]:chunk_end[3]], sep="\n")
```

:::notes
Always load the tidyverse library. You also need the readxl library because the data is stored as an Excel file.
:::

## simon-5501-01-template.qmd, 4

```{r write4a}
cat(template[chunk_beg[4]:chunk_end[4]], sep="\n")
```

:::notes
While you might want to see information about all 31 columns of data, I can't display them all on a single slide. Column 1 is the id and columns 5, 6, and 7 are demographic variables.
:::

## simon-5501-01-template.qmd, 5

```{r write5a}
cat(template[chunk_beg[5]:chunk_end[5]], sep="\n")
```

:::notes
Here are counts and percentages for the first demographic variable. There are several alternative methods to get counts and percentages. The table function and the count function are both good alternatives.

Note the interpretation. You should always provide an interpretation.
:::