---
title: "Counts and percentages"
format: 
  html:
    slide-number: true
    embed-resources: true
editor: source
execute:
  echo: true
  message: false
  warning: false
---

## Data source

This program uses data from a study of sharing services (like sharing an automobile) and produces counts and percentages for a few demographic variables. There is a [data dictionary][dd] that provides more details about the data. 

[dd]: https://github.com/pmean/datasets/blob/master/sharing.yaml

## Libraries

Here are the libraries you need for this program.

```{r setup}
library(readxl)
library(tidyverse)
```

## Reading the data

Here is the code to read the data and show a glimpse. There are 31 columns total, but I am showing just a few of the columns here.

```{r read}
fn <- "../data/sharing.xlsx"
sharing <- read_excel(fn)
glimpse(sharing[ , c(1, 5:7)])
```

## Calculate counts and percentages for age group

```{r count-age-groups}
sharing |>
  group_by(age) |>
  summarize(n=n()) |>
  mutate(total=sum(n)) |>
  mutate(pct=100*n/total)
```

The survey respondents were younger than the general population. About half of the survey respondents were 18 to 25 years old. Only 3% were over 60. Six ages were missing.