---
title: "Univariate statistics for Legionnaires disease"
format: 
  html:
    slide-number: true
    embed-resources: true
editor: source
execute:
  echo: true
  message: false
  warning: false
---

## Data source

This program uses data from a fictional study of Legionnaires disease and produces some simple univariate statistics: means, standard deviations, and percentiles. There is a [data dictionary][dd] that provides more details about the data. 

[dd]: https://github.com/pmean/data/blob/main/files/legionnaires-disease.yaml

This is the correct program for the Fall 2026 semester.

This code is placed in the public domain.

## Datasets used in this program

-   ld: the raw data from legionnaires disease
-   ld1: modify a variable name

## Libraries

```{r setup}
library(tidyverse)

R.version.string
Sys.Date()
```

## Reading the data

```{r read}
fn <- "../data/legionnaires-disease.txt"
ld <- read_tsv(fn, col_types="cnn")
glimpse(ld)
```

## Rename variables

```{r rename}
ld |>
  rename(room_number=`Room Number`) 
  rename_with(tolower) -> ld1
glimpse(ld1)
```

#### Comments on the code

Notice how R encloses the first variable name (Room Number) in back-quotes. This is needed when a variable includes an embedded blank. You should rename this variable at your first opportunity.

I find that many of the mistakes that I make are due to inconsistencies in how I name variables. Capitalization is one of the biggest problems. So I have gotten into the habit of converting variable names to all lower case. That way I don't have to worry about whether it is "Before" or "before". Here is the code to convert every capital letter to a lowercase letter.

The [rename function][ref-rename] is part of the dplyr package and will change the name of a variable. The new name goes to the left of the equal sign and the old name goes to the right. The rename_with function, described on the same link as above, will rename multiple variables using a function. The [tolower function][ref-tolower] is part of base R and converts any uppercase letters to lowercase. 

[ref-rename]: https://dplyr.tidyverse.org/reference/rename.html
[ref-tolower]: https://stat.ethz.ch/R-manual/R-devel/library/base/html/chartr.html

## Calculate means and standard deviations before remediation

```{r before-means}
ld |>
  summarize(
    before_mn=mean(before),
    before_sd=sd(before)) 
```

#### Comments on the code

The [sd function][ref-sd] is part of base R and computes a standard deviation. This standard deviation uses the traditional n-1 in the denominator. If you have mssing values, the default for sd is to calculate the standard deviation as missing. If you want to exclude missing values, add the argument na.rm=TRUE. 

[ref-sd]: https://stat.ethz.ch/R-manual/R-devel/library/stats/html/sd.html

#### Interpretation of the output

The average colony count per cubic foot before remediation, 11.3, is quite large. The standard deviation, 2.8, represents a moderate amount of variation in this variable.

## Calculate means and standard deviations after remediation

```{r after-means}
ld |>
  summarize(
    after_mn=mean(after),
    after_sd=sd(after)) 
```

#### Interpretation of the output

The average colony count per cubic foot after remediation, 9.7, is still quite large. The standard deviation, 3.2, represents a moderate amount of variation in this variable and is roughly comparable to the variation before remediation.

## Calculate median and range before intervention

You could also use "median(before)" and  "min(before)" and "max(before)" in the code below.

```{r before-quantiles}
ld |>
  summarize(
    before_median=quantile(before, probs=0.5),
    before_min=quantile(before, probs=0), 
    before_max=quantile(before, probs=1)) 
```

#### Comments on the code

The [quantile function][ref-quantile] is part of base R and compute a quantile between 0 and 1. When you uses the argument probs=0.5, you are computing the median. There is a median function in base R that you could have used instead. The extreme values of 0 and 1 represent the minimum and maximum. and you can use the min and max functions instead. All of these functions can use the na.rm argument to contorl what happens when there are missing values.

The names that you choose for the left hand side of the equal sign are arbitrary. You should choose a descriptive name, but you have lots of options. A median of the before and after values could be called

-   Before_median, After_median
-   Median0, Median1
-   Second_quartile_A, Second_quartile_B
-   or many other reasonable choices.


[ref-quantile]: https://stat.ethz.ch/R-manual/R-devel/library/stats/html/quantile.html

#### Interpretation of the output

The median colony count before remediation, 11.3, is roughly the same as the mean. The data ranges from 7.1 to 14.6 colonies per cubic centimeter, a fairly wide range.

## Calculate median and range after intervention

```{r after-quantiles}
ld |>
  summarize(
    after_q50=quantile(after, probs=0.5),
    after_min=quantile(after, probs=0), 
    after_max=quantile(after, probs=1)) 
```

#### Comments on the code

The names that you choose for the left hand side of the equal sign are arbitrary. You should choose a descriptive name, but you have lots of options. A median of the before and after values could be called

-   Before_median, After_median
-   Median0, Median1
-   Second_quartile_A, Second_quartile_B
-   or many other reasonable choices.

#### Interpretation of the output

The median colony count, 10.3, is slightly lower after remediation. The data range from 3.8 to 13.7 colonies per cubic centimeter and is about as wide as the range before remediation.
