---
title: "03-02, Normal probabilities and quantiles"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

## simon-5501-03-normal-calculations.qmd, 1

```{r write0a}
library(tidyverse)
fn <- "simon-5501-03-normal-calculations.qmd"
template <- read_lines(file=fn)
h <- str_which(template, "##")
chunk_beg <- c(1, h)
chunk_end <- c(h-1, length(template))
```

```{r write1a}
cat(template[chunk_beg[1]:chunk_end[1]], sep="\n")
```

:::notes
Here is a program template that illustrates how to calculate probabilities and quantiles for the standard normal distribution

The first few lines are the documentation header.
:::

## simon-5501-03-normal-calculations.qmd, 2

```{r write3a}
cat(template[chunk_beg[2]:chunk_end[2]], sep="\n")
```

:::notes
You only need the tidyverse libraries for this program.
:::

## simon-5501-03-normal-calculations.qmd, 3

```{r write2a}
cat(template[chunk_beg[3]:chunk_end[3]], sep="\n")
```

:::notes
This code draws the bell shaped curve for a standard normal distribution. The graph is saved as normal_curve to allow for future modifications.
:::

## simon-5501-03-normal-calculations.qmd, 4

```{r write4a}
cat(template[chunk_beg[4]:chunk_end[4]], sep="\n")
```

:::notes
This code computes the probability that a standard normal variable is less than a particular number and displays the area associated with this probability.
:::

## simon-5501-03-normal-calculations.qmd, 5

```{r write5a}
cat(template[chunk_beg[5]:chunk_end[5]], sep="\n")
```

:::notes
Here's a similar calculation.
:::

## simon-5501-03-normal-calculations.qmd, 6

```{r write6a}
cat(template[chunk_beg[6]:chunk_end[6]], sep="\n")
```

:::notes
For greater than probabilities (probabilities corresponding to area to the right), subtract the pnorm result from 1.
:::

## simon-5501-03-normal-calculations.qmd, 7

```{r write7a}
cat(template[chunk_beg[7]:chunk_end[7]], sep="\n")
```

:::notes
Here's another greater than probability calculation.
:::

## simon-5501-03-normal-calculations.qmd, 8

```{r write8a}
cat(template[chunk_beg[8]:chunk_end[8]], sep="\n")
```

:::notes
For probabilities between two values, calculate the difference.
:::

## simon-5501-03-normal-calculations.qmd, 9

```{r write9a}
cat(template[chunk_beg[9]:chunk_end[9]], sep="\n")
```

:::notes
Here's a similar probability calculation.
:::

## simon-5501-03-normal-calculations.qmd, 10

```{r write10a}
cat(template[chunk_beg[10]:chunk_end[10]], sep="\n")
```

:::notes
The normal quantile is the value associated with a specified probability. Use the qnorm function for quantile calculations.
:::

## simon-5501-03-normal-calculations.qmd, 11

```{r write11a}
cat(template[chunk_beg[11]:chunk_end[11]], sep="\n")
```

:::notes
Here is another quantile calculation.
:::

