---
title: "03-02, Normal probabilities"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

## Overview

The normal distribution is an important component of most statistical analyses. The formula for the normal distribution is quite complex, 

$f(x; \mu, \sigma)=\frac{1}{\sqrt{2\pi}}e^{-\frac{(x-\mu)^2}{2\sigma^2}}$

but the shape is readily recognizable.

```{r setup-2}
#| echo: false
library(tidyverse)
```

```{r normal-bell}
#| echo: false
normal_bell <- function(mu=0, sigma=1) {
  x0 <- seq(-8, 8, length=500)
  y0 <- dnorm(x0, mu, sigma)
  normal <- data.frame(
    x = c(-8, x0, 8),
    y = c(0, y0, 0))
  ggplot(data=normal, aes(x, y)) +
    geom_polygon(
      fill="white",
      color="black") +
    expand_limits(y=c(0, 0.8)) +
    scale_x_continuous(breaks=2*(-4:4))
}
```

## Normal distribution with mu=2, sigma=1

```{r normal_a}
normal_bell(mu=2, sigma=1)
```

## Normal distribution with mu=-3, sigma=1

```{r normal_b}
normal_bell(mu=-3, sigma=1)
```

## Normal distribution with mu=1, sigma=1

```{r normal_c}
normal_bell(mu=1, sigma=1)
```


## Normal distribution with mu=1, sigma=2

```{r normal_d}
normal_bell(mu=1, sigma=2)
```

## Normal distribution with mu=1, sigma=2

```{r normal_e}
normal_bell(mu=1, sigma=0.5)
```

## Standard normal distribution (mu=0, sigma=1)

```{r normal_f}
normal_bell()
```


```{r standard-normal}
#| echo: false
draw_bell <- function(
    lo=-3.5, 
    hi=3.5, 
    mu=0,
    sigma=1,
    range=c(-4, 4)) {
  x0 <- seq(
      (-3.5-mu)/sigma, 
      ( 3.5-mu)/sigma, 
      length=100)
  y0 <- dnorm(x0)
  x1 <- seq(
      (lo-mu)/sigma, 
      (hi-mu)/sigma,
      length=100)
  y1 <- dnorm(x1)
  std_normal <- data.frame(
    x = c(-3.5, x0, 3.5),
    y = c(0, y0, 0))
  shaded_subset <- data.frame(
    x = c(lo, x1, hi),
    y = c(0, y1, 0))

  ggplot(data=std_normal, aes(x, y)) +
    geom_polygon(
      fill="white",
      color="black") +
    expand_limits(x=c(-4, 4)) +
    geom_polygon(
      data=shaded_subset,
      aes(x, y),
      fill="gray",
      color="black")
}
```

## P[Z < 1.5]

```{r}
a <- 1.5
draw_bell(hi=a) +
  geom_label(x=a, y=0, label=a) +
  geom_label(x=a-1.5, y=0.5*dnorm(a-1.5), label="?")
```

```{r}
#| echo: true
pnorm(1.5)
```

## P[Z < -0.5]

```{r}
a <- -0.5
draw_bell(hi=a) +
  geom_label(x=a, y=0, label=a) +
  geom_label(x=a-0.5, y=0.5*dnorm(a-0.5), label="?")
```

```{r}
#| echo: true
pnorm(-0.5)
```

## P[Z > 1]

```{r}
a <- 1
draw_bell(lo=a) +
  geom_label(x=a, y=0, label=a) +
  geom_label(x=a+0.5, y=0.5*dnorm(a+0.5), label="?")
```

```{r}
#| echo: true
1- pnorm(1.0)
```

## P[Z > -2]

```{r}
a <- -2
draw_bell(lo=a) +
  geom_label(x=a, y=0, label=a) +
  geom_label(x=a+2, y=0.5*dnorm(a+2), label="?")
```

```{r}
#| echo: true
1- pnorm(-2)
```

## P[-2.5 < Z < 2.5]

```{r}
a <- -2.5
b <- 2.5
draw_bell(lo=a, hi=b) +
  geom_label(x=a, y=0, label=a) +
  geom_label(x=b, y=0, label=b) +
  geom_label(x=0, y=0.5*dnorm(0), label="?")
```

```{r}
#| echo: true
pnorm(2.5)- pnorm(-2.5)
```

## P[-0.5 < Z < 0.5]

```{r}
a <- -0.5
b <- 0.5
draw_bell(lo=a, hi=b) +
  geom_label(x=a, y=0, label=a) +
  geom_label(x=b, y=0, label=b) +
  geom_label(x=0, y=0.5*dnorm(0), label="?")
```

```{r}
#| echo: true
pnorm(0.5) - pnorm(-0.5)
```

## 25th percentile of a standard normal

```{r}
p <- 0.25
a <- qnorm(p)
b <- 0.5
draw_bell(hi=a) +
  geom_label(x=a, y=0, label="?") +
  geom_label(x=a-0.5, y=0.5*dnorm(a-0.5), label=p)
```

```{r}
#| echo: true
qnorm(0.25)
```

## 90th percentile of a standard normal

```{r}
p <- 0.9
a <- qnorm(p)
b <- 0.5
draw_bell(hi=a) +
  geom_label(x=a, y=0, label="?") +
  geom_label(x=0, y=0.5*dnorm(0), label=p)
```

```{r}
#| echo: true
qnorm(0.9)
```

