---
title: "03-01, The normal distribution"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

```{r setup-1}
library(tidyverse)
```

```{r histogram}
draw_histogram <- function(x, w=1) {
  x1 <- round(x*w)/w
  data.frame(x=x1) |>
    ggplot(aes(x)) +
      theme(legend.position="none") +
      xlab(" ") +
      ylab(" ") +
      geom_bar(
        color="black", 
        fill="white",
        width=w)
}
```

## The bell shaped curve

-   Does your variation follow a bell shaped curve?
-   Synonyms: normality, normal distribution
    -   Values in the middle are most common
    -   Frequencies taper off away from the center
    -   Symmetry on either side

::: notes
*Speaker notes*

Much variation in the real world follows a bell shaped curve, alternately called a normal distribution. You can assess whether you have a bell shaped curve using a histogram. Look for values in the middle being most common. The frequencies should taper off slowly as you moved away from the middle. The histogram should have symmetry. The left side of the histogram should be roughly equivalent to the right side of the histogram.
:::

## Bell-shaped curve is a limit

```{r limit-1}
x_normal <- pmax(0, rnorm(2000, 10, 3))
draw_histogram(x_normal[1:10])
```

## Bell-shaped curve is a limit

```{r limit-2}
draw_histogram(x_normal[1:20], w=0.5)
```

## First example of non-normal data, bimodal

```{r bimodal}
x_bimodal <- c(rnorm(1200, 4, 1.5), rnorm(800, 10, 1.5))
draw_histogram(x_bimodal)
```

::: notes
*Speaker notes*

Here's a histogram that shows a bimodal distribution. The frequencies are not highest in the center of the data. This is not a bell shaped curve.
:::

## Second example of non-normal data, skewed

```{r skewed}
x_skewed <- rgamma(2000, 2, scale=6)
draw_histogram(x_skewed)
```

::: notes
*Speaker notes*

Here's a histogram that shows a skewed or asymmetric distribution. This is not a bell shaped curve.
:::

## Third example of non-normal data, light-tailed

```{r light-tailed}
x_light_tailed <- runif(2000, 0, 15)
draw_histogram(x_light_tailed)
```

::: notes
*Speaker notes*

Here's a histogram that shows a symmetric distribution, but the frequencies do not taper off as you move away from the center. This is not a bell shaped curve.
:::

## Fourth example of skewed distribution, heavy-tailed

```{r heavy-tailed}
x0 <- rt(1000, 2)
x1 <- 15 + c(x0, -x0) + rnorm(2000, mean=0, sd=0.2)
x_heavy_tailed <- pmin(pmax(0, x1), 30)
draw_histogram(x_heavy_tailed)  
```

::: notes
*Speaker notes*

Here's a histogram that shows a symmetric distibution, but the frequencies taper off at first, but then flatten out. This is called a heavy tailed distribution and it tends to produce outliers, extreme values, on both sides. This is not a bell shaped curve.
:::

## 

```{r normal}
draw_histogram(x_normal)
```


::: notes
*Speaker notes*

Here's a histogram that shows a symmetric distribution, with the most frequent values in the center and frequencies that taper off on either side.

This is a bell shaped curve.
:::

## Why concern yourself with the bell shaped curve?

-   You can characterize individual observations
-   You can characterize summary measures

