---
title: "Normality"
format: 
  html:
    embed-resources: true
editor: source
---

This program uses data from a study of pulmonary function in children. The key variable is forced expiratory volume (fev) measured in liters There is a [data dictionary][dd] that provides more details about the data. The program was written by Steve Simon on 2024-09-02 and is placed in the public domain.

[dd]: https://github.com/pmean/datasets/blob/master/fev.yaml

## Libraries

The tidyverse library is the only one you need for  this program.

```{r setup}
#| message: false
#| warning: false
library(tidyverse)
```

## Reading the data

Here is the code to read the data and show a glimpse. 

```{r read}
fev <- read_csv(
  file="../data/fev.csv",
  col_names=c(
    "age",
    "fev",
    "ht",
    "sex",
    "smoke"),
  col_types="nnncc")
  glimpse(fev)
```

## Calculate mean and standard deviation for fev

```{r descriptive-fev}
fev |>
  summarize(
    fev_mean=mean(fev),
    fev_stdv=sd(fev))
```

## Histogram for fev, wide bars

```{r histogram-fev-wide}
ggplot(
  data=fev,
  aes(x=fev)) +
  geom_histogram(
    binwidth=0.5,
    color="black",
    fill="white")
```

## Histogram for fev, narrow bars

```{r histogram-fev-narrow}
ggplot(
  data=fev,
  aes(x=fev)) +
  geom_histogram(
    binwidth=0.1,
    color="black",
    fill="white")
```

## Normal probability plot for fev

```{r qqplot-fev}
qqnorm(fev$fev)
```

```{r qqplot-log-fev}
qqnorm(log(fev$fev))
```

