---
title: "Calculations involving the normal distribution"
format: 
  html:
    slide-number: true
    embed-resources: true
editor: source
execute:
  echo: true
  error: true
  message: false
  warning: false
---

## Overview

The normal distribution is an important component of most statistical analyses. The formula for the normal distribution is quite complex, 

$f(x; \mu, \sigma)=\frac{1}{\sqrt{2\pi}}e^{-\frac{(x-\mu)^2}{2\sigma^2}}$

but the shape is readily recognizable.


is program uses data from a study of pulmonary function in children. The key variable is forced expiratory volume (fev) measured in liters There is a [data dictionary][dd] that provides more details about the data. 

[dd]: https://github.com/pmean/datasets/blob/master/fev.yaml

## Libraries

Here are the libraries you need for this program.

```{r setup}
library(tidyverse)
```

## The standard normal distribution

```{r standard-normal}
std_normal <- data.frame(
  x <- seq(-3.5, 3.5, length=1000),
  y <-dnorm(x))
ggplot(data=std_normal, aes(x, y)) +
  geom_line() + 
  geom_segment(
    aes(xend=x, yend=0), 
    color="white") +
  geom_line() +
  geom_segment(x=-3.5, y=0, xend=3.5, yend=0)
```