---
title: "Normal probabilities and quantiles"
format: 
  html:
    embed-resources: true
editor: source
execute: 
  error: true
---

## File details

This program is for internal use, but you are welcome to look things over.

This program displays the standard normal curve along with various probabilities and quantiles. It was written by Steve Simon on 2024-09-01 and is placed in the public domain.

## Load the tidyverse library

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
```

## Functions for drawing

```{r histogram}
draw_histogram <- function(x, w=1) {
  x1 <- round(x/w)*w
  data.frame(x=x1, wt=w/length(x1)) |>
    ggplot(aes(x, weight=wt)) +
      theme(legend.position="none") +
      xlab(" ") +
      ylab(" ") +
      geom_bar(
        color="black", 
        fill="white",
        width=w)
}
```

```{r boxplot}
draw_boxplot <- function(x) {
  data.frame(x) |>
    ggplot(aes(x)) +
      geom_boxplot()
}
```

```{r q-q-plot}
draw_qqplot <- function(y) {
  data.frame(x=y) |>
    mutate(z=(x-mean(y))/sd(y)) |>
    mutate(r=percent_rank(z)) |>
    mutate(q=qnorm(r)) |>
    ggplot(aes(x=x, y=q)) +
      geom_point()
}
```

## Bell-shaped curve is a limit

```{r limit-1}
n <- 100; w <- 1
x_bell <- qnorm((1:n)/(n+1), mean=15, sd=5)
draw_histogram(x_bell, w)
ggsave("../images/histogram-01.png")
```

## Bell-shaped curve is a limit

```{r limit-2}
n <- n*4; w <- w/2
x_bell <- qnorm((1:n)/(n+1), mean=15, sd=5)
draw_histogram(x_bell, w)
ggsave("../images/histogram-02.png")
```

## Bell-shaped curve is a limit

```{r limit-3}
n <- n*4; w <- w/2
x_bell <- qnorm((1:n)/(n+1), mean=15, sd=5)
draw_histogram(x_bell, w)
ggsave("../images/histogram-03.png")
```

## Bell-shaped curve is a limit

```{r limit-4}
n <- n*4; w <- w/2
x_bell <- qnorm((1:n)/(n+1), mean=15, sd=5)
draw_histogram(x_bell, w)
ggsave("../images/histogram-04.png")
```

```{r normal-bell}
#| echo: false

normal_bell <- function(mu=0, sigma=1) {
  x0 <- seq(-8, 8, length=500)
  y0 <- dnorm(x0, mu, sigma)
  normal <- data.frame(
    x = c(-8, x0, 8),
    y = c(0, y0, 0))
  ggplot(data=normal, aes(x, y)) +
    geom_polygon(
      fill="white",
      color="black") +
    expand_limits(y=c(0, 0.8)) +
    scale_x_continuous(breaks=2*(-4:4))
}
```

## Normal distribution with mu=2, sigma=1

```{r normal_a}
normal_bell(mu=2, sigma=1)
ggsave("../images/bell-01.png")
```

## Normal distribution with mu=-3, sigma=1

```{r normal_b}
normal_bell(mu=-3, sigma=1)
ggsave("../images/bell-02.png")
```

## Normal distribution with mu=1, sigma=1

```{r normal_c}
normal_bell(mu=1, sigma=1)
ggsave("../images/bell-03.png")
```


## Normal distribution with mu=1, sigma=2

```{r normal_d}
normal_bell(mu=1, sigma=2)
ggsave("../images/bell-04.png")
```

## Normal distribution with mu=1, sigma=0.5

```{r normal_e}
normal_bell(mu=1, sigma=0.5)
ggsave("../images/bell-05.png")
```

## Standard normal distribution (mu=0, sigma=1)

```{r normal_f}
normal_bell()
ggsave("../images/bell-06.png")
```


## Using R to draw the standard normal curve

```{r}
#| label: standard-normal

x <- seq(from=-4, to=4, length=100)
y <- dnorm(x)
data.frame(x, y) |>
  ggplot() + 
  aes(x, y) +
  geom_polygon(fill="white", color="black") -> normal_curve
normal_curve
ggsave("../images/normal-00.png")
```

#### Comments on the code

The [seq function][ref-seq], part of base R, calculates evenly spaced values within an interval. The code here calculates 100 evenly spaced values between -4 and +4. The [dnorm function][ref-dnorm] is part of base R and calculates the height of the normal curve. You can specify a mean and standard deviation, but if you leave it out, it specifies the height for a standard normal.

The [ggplot function][ref-ggplot] is part of the [ggplot2 library][ref-ggplot2] and included within the tidyverse libary. This function allows you to specify the data for plotting.

The [aes function][ref-aes] is short for aesthetics. It is part of the ggplot2 library and defines which variables are used for axes, color, markers, etc.

The [geom-polygon function][ref-polygon] defines an irregular shape using the X and Y values defined in aes. It allows you to control the style of the border and the color used to fill the polygon.

The [ggsave function][ref-ggsave] is part of the ggplot2 library. It saves the results of a ggplot function as a file. It looks at the extension to decide what format to use.

[ref-seq]: https://stat.ethz.ch/R-manual/R-devel/RHOME/library/base/html/seq.html

[ref-dnorm]: https://stat.ethz.ch/R-manual/R-devel/library/stats/html/Normal.html

[ref-ggplot]: https://ggplot2.tidyverse.org/reference/ggplot.html

[ref-ggplot2]: https://ggplot2.tidyverse.org/

[ref-aes]: https://ggplot2.tidyverse.org/reference/aes.html

[ref-polygon]: https://ggplot2.tidyverse.org/reference/geom_polygon.html

[ref-ggsave]: https://ggplot2.tidyverse.org/reference/ggsave.html

## P[Z < 1.5]

```{r}
#| label: prob-1

a <- 1.5
normal_curve +
  geom_vline(xintercept=a) +
  geom_label(x=a, y=0.4, label=a) +
  geom_label(x=a-0.5, y=0, label="Area = ?") -> prob1
prob1
ggsave("../images/normal-01.png")
pnorm(1.5)
```

#### Comments on the code

Use the [geom_vline function][ref-vline] and [geom_label function][ref-label], both part of the ggplot2 library to draw a vertical reference line and add text to the normal curve.

The [pnorm function][ref-pnorm] computes the standard normal probability. The link listed here also describes the qnorm function used later in this program.

[ref-vline]: https://ggplot2.tidyverse.org/reference/geom_abline.html

[ref-label]: https://ggplot2.tidyverse.org/reference/geom_text.html

[ref-pnorm]: https://stat.ethz.ch/R-manual/R-devel/library/stats/html/Normal.html

#### Interpretation of the output

The probability that a standard normal variable is less than 1.5 is about 93%.

## P[Z < -0.5]

```{r}
#| label: prob-2

a <- -0.5
normal_curve +
  geom_vline(xintercept=a) +
  geom_label(x=a, y=0.4, label=a) +
  geom_label(x=a-0.5, y=0, label="Area = ?") -> prob2
prob2
ggsave("../images/normal-02.png")
pnorm(-0.5)
```

#### Interpretation of the output

The probability that a standard normal variable is less than -0.5 is about 31%.

## P[Z > 1]

```{r}
#| label: prob-3
a <- 3
normal_curve +
  geom_vline(xintercept=a) +
  geom_label(x=a, y=0.4, label=a) +
  geom_label(x=a+0.5, y=0, label="Area = ?") -> prob3
prob3
ggsave("../images/normal-03.png")
1- pnorm(3)
```

#### Comments on the code

When you are calculating the probability on the right (probability greater than some number), use 1-pnorm.

#### Interpretation of the output

The probability that a standard normal variable is larger than 3 is about 0.1%.

## P[Z > -2]

```{r}
#| label: prob-4

a <- -2
normal_curve +
  geom_vline(xintercept=a) +
  geom_label(x=a, y=0.4, label=a) +
  geom_label(x=a+0.5, y=0, label="Area = ?") -> prob4
prob4
ggsave("../images/normal-04.png")
1- pnorm(-2)
```

#### Interpretation of the output

The probability that a standard normal variable is larger than -2 is about 98%.

## P[-2.5 < Z < 2.5]

```{r}
#| label: prob-5

a <- 2.5
normal_curve +
  geom_vline(xintercept=-a) +
  geom_vline(xintercept= a) +
  geom_label(x=-a, y=0.4, label=-a) +
  geom_label(x= a, y=0.4, label= a) +
  geom_label(x=0, y=0, label="Area = ?") -> prob5
prob5
ggsave("../images/normal-05.png")
pnorm(2.5)- pnorm(-2.5)
```

#### Comments on the code

When you are calculating the probability between two values, compute pnorm of the larger value minus pnorm of the smaller value.

#### Interpretation of the output.

The probability that a standard normal variable is between -2.5 and 2.5 is approximately 99%.

## P[-0.5 < Z < 0.5]

```{r}
#| label: prob-6

a <- 0.5
normal_curve +
  geom_vline(xintercept=-a) +
  geom_vline(xintercept= a) +
  geom_label(x=-a, y=0.4, label=-a) +
  geom_label(x= a, y=0.4, label= a) +
  geom_label(x=0, y=0, label="Area = ?") -> prob6
prob6
ggsave("../images/normal-06.png")
pnorm(0.5) - pnorm(-0.5)
```

#### Interpretation of the output

The probability that a standard normal variable is between -0.5 and 0.5 is about 38%.

## 25th percentile of a standard normal

```{r}
#| label: quantile-1

p <- 0.25
a <- qnorm(p)
normal_curve +
  geom_vline(xintercept=a) +
  geom_label(x=a, y=0.4, label="Quantile = ?") +
  geom_label(x=a-0.5, y=0, label=p) -> prob7
prob7
ggsave("../images/normal-07.png")
qnorm(0.25)
```

#### Comments on the code

Use the qnorm function, part of base R, to calculate quantiles of the standard normal distribution. A description of the qnorm function is included in the link describing the pnorm function.

#### Interpretation of the output

The 25th percentile of a standard normal distribution is -0.67.

## 90th percentile of a standard normal

```{r}
#| label: quantile-2

p <- 0.9
a <- qnorm(p)
normal_curve +
  geom_vline(xintercept=a) +
  geom_label(x=a, y=0.4, label="Quantile = ?") +
  geom_label(x=a-0.5, y=0, label=p) -> prob8
prob8
ggsave("../images/normal-08.png")
qnorm(0.9)
```

#### Interpretation of the output

The 90th percentile of a standard normal distribution is 1.28.

## Birthweight calculations

```{r}
#| label: bw

mu <- 3.2
sigma <- 0.5

x <- seq(mu-4*sigma, mu+4*sigma, length=100)
y <- dnorm(x, mu, sigma)

data.frame(x, y) |>
  ggplot() +
  aes(x, y) +
  scale_x_continuous(breaks=seq(1.4, 5.0, by=0.2)) +
  geom_polygon(fill="white", color="black") -> bw
bw
ggsave("../images/birthweight-01.png")
```

```{r}
mu+qnorm(0.9)*sigma
mu+qnorm(0.1)*sigma
pnorm((4-mu)/sigma)
```

## Height calculations

```{r}
#| label: ht

mu <- 160
sigma <- 7

x <- seq(mu-4*sigma, mu+4*sigma, length=100)
y <- dnorm(x, mu, sigma)

data.frame(x, y) |>
  ggplot() +
  aes(x, y) +
  scale_x_continuous(breaks=seq(135, 185, by=5)) +
  geom_polygon(fill="white", color="black") -> bw
bw
ggsave("../images/height-01.png")
```

