---
title: "Analysis of fev data"
format: 
  html:
    embed-resources: true
editor: source
---

This program assesses the normality of variables in a study of pulmonary function in children. There is a [data dictionary][dd] that provides more details about the data. The program was written by Steve Simon on 2024-09-02 and is placed in the public domain.

[dd]: https://github.com/pmean/datasets/blob/master/fev.yaml

## Libraries

The tidyverse library is the only one you need for  this program.

```{r setup}
#| message: false
#| warning: false
library(tidyverse)
```

## List variable names

Since the variable names are not listed in the data file itself, you need to list them here.

```{r names}
fev_names <- c(
    "age",
    "fev",
    "ht",
    "sex",
    "smoke")
```

## Reading the data

Here is the code to read the data and show a glimpse. 

```{r read}
fev <- read_csv(
  file="../data/fev.csv",
  col_names=fev_names,
  col_types="nnncc")
glimpse(fev)
```

## Calculate mean and standard deviation for fev

To orient yourself to the data, calculate a few descriptive statistics.

```{r descriptive-fev}
fev |>
  summarize(
    fev_mean=mean(fev),
    fev_stdv=sd(fev))
```

## Histogram for fev, wide bars

```{r histogram-fev-wide}
ggplot(data=fev, aes(x=fev)) +
  geom_histogram(
    binwidth=0.5,
    color="black",
    fill="white")
```

## Histogram for fev, narrow bars

```{r histogram-fev-narrow}
ggplot(data=fev, aes(x=fev)) +
  geom_histogram(
    binwidth=0.1,
    color="black",
    fill="white")
```

Although some may interpret these histograms as showing a slight skewness, I would interpret them as being approximately normal.

## Normal probability plot for fev

The qqnorm function produces a normal probability plot. The default option for most plots is landscape orientation (the width is larger than the height). The q-q plot, however, looks best if figure width and height are equal.

```{r qqplot-fev}
#| fig-width: 5
#| fig.height: 5
qqnorm(fev$fev)
```

The normal probability plot is reasonably close to a straight line, indicating that the data comes reasonably close to following a normal distribution.
