---
title: "Analysis of variance table"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

## Data for multiple linear regression

-   Dependent variable: $Y_1, Y_2,\ldots,Y_n$
-   First independent variable: $X_{11}, X_{12}, \ldots, X_{1n}$
-   Second independent variable: $X_{21}, X_{22}, \ldots, X_{2n}$
-   ...
-   kth independent variable: $X_{k1}, X_{k2}, \ldots, X_{kn}$

::: notes

For a multiple linear regression model, you need a dependent or outcome variable, traditionally designated by the letter "Y". The are n values of Y, one for each patient in your research study.

Then you have k independent variables. The first one, designated by $X_1$ also has n values. The second independent variable, $X_2$ works similarly. There are k total independent variables.

For now, let's assume that every variable, the one dependent variable and the k independent variables, is continuous.

:::

## Matrix notation, 1 of 3

\begin{matrix}Y=& \begin{bmatrix}
Y_1\\
Y_2\\
\vdots\\ 
Y_{j}\\
\vdots\\ 
Y_n
\end{bmatrix}
\end{matrix}

::: notes

Often you will see multiple linear regression defined with a single column for the dependent variable Y ...

:::

## Matrix notation, 2 of 3

\begin{matrix}X=& \begin{bmatrix}
X_{11} & X_{21} & \ldots & X_{i1} & \ldots & X_{k1}\\
X_{12} & X_{22} & \ldots & X_{i2} & \ldots & X_{k2}\\
\vdots & \vdots &        & \vdots &        & \vdots\\ 
X_{1j} & X_{2j} & \ldots & X_{ij} & \ldots & X_{kj}\\
\vdots & \vdots &        & \vdots &        & \vdots\\ 
X_{1n} & X_{2n} & \ldots & X_{in} & \ldots & X_{kn}
\end{bmatrix}
\end{matrix}

::: notes

... and a matrix representing the independent variables. Each column in the matrix represents values of a particular independent variable for every subject and each row represents all of the independent variables for a particular subject.

:::

## Matrix notation, 3 of 3

\begin{matrix}X=& \begin{bmatrix}
1      & X_{11} & X_{21} & \ldots & X_{i1} & \ldots & X_{k1}\\
1      & X_{12} & X_{22} & \ldots & X_{i2} & \ldots & X_{k2}\\
\vdots & \vdots & \vdots &        & \vdots &        & \vdots\\ 
1      & X_{1j} & X_{2j} & \ldots & X_{ij} & \ldots & X_{kj}\\
\vdots & \vdots & \vdots &        & \vdots &        & \vdots\\ 
1      & X_{1n} & X_{2n} & \ldots & X_{in} & \ldots & X_{kn}
\end{bmatrix}
\end{matrix}

::: notes

You may see a similar matrix, but with an extra column at the beginning with the value of 1 repeated down that column.

There are lots of theoretical reasons why the dependent variable and the independent variables are laid out in this format. I will not talk about this in detail. I just wanted to show you this, so you would not be alarmed if you see it in some of the future readings you might have on multiple linear regression.

:::

## Why use multiple linear regression

-   Two competing purposes
    -   Mechanisms
        -   What variables have an impact on your outcome?
    -   Prediction
        -   What outcome will you see on tomorrow's patient?
        -   "It is difficult to make predictions, especially about the future." Niels Bohr
    
::: notes
It is important to note that there are two competing reasons for using multiple linear regression (and other statistical models as well). The first is to try to understand the underlying mechanisms that explain the "how". You are looking to understand how a disease harms or how a treatment helps. This could be called a "white box" feature that you get with multiple linear regression. The size and signs of the regression coefficients help you understand what factors contribute to the "how".

You might think of a mechanistic goal as being able to peek under the hood. Some people are perfectly happy driving a car without having any idea of how the engine works. This is fine, and you are treating the automobile as a black box. You don't know what goes on underneath the hood of the car. Other people want to understand how a car works and they believe (properly in my opinion) that this understanding makes them a better driver.

Alternatively, you can use a multiple linear regression model to predict the future. You want to tell a patient about his or her prognosis, you want to estimate the staffing needs in a hospital, or you want to estimate the cost savings associated changes in health care delivery.

Predicting the future is a perilous process. Pretty much every investment opportunity carries a warning "Past performance is no guarantee of future results
:::

## Multiple linear regression model

-   $Y_i=\beta_0+\beta_1 X_{1i}+\beta_2 X_{2i}+\ldots+\beta_k X_{ki}+\epsilon_i$
    -   Assumptions: $\epsilon_i$ are
        -   independent, 
        -   same variance,
        -   normally distributed.
    -   Independent variables must be on interval or ratio scale     
        -   Nominal/ordinal scales require some care
        
::: notes
The multiple linear regression model says that $Y_i$ is a linear function of $X_{1i},\ X_{2i},\ \ldots,\ X_{ki}$ plus an error term. The error term accounts for the fact that even the best set of independent variables will still fall a bit short.

The $\epsilon_i$ have to meet three assumptions: independence, equal variances, and normally distributed.

It's often not stated directly, but your independent variables must be on an interval or ratio scale. There are ways accommodate independent variables on an ordinal or interval scale, but these methods require a bit of care. 
:::

## Estimates for the multiple linear regression model

-   Use least squares to estimate $\hat\beta_0,\ \hat\beta_1,\ \hat\beta_2,\ \ldots,\ \hat\beta_k$
    -   $\hat Y_i=\hat\beta_0+\hat\beta_1 X_{1i}+\hat\beta_2 X_{2i}+\ldots+\hat\beta_k X_{ki}$
    -   $e_i=Y_i-\hat Y_i$

::: notes
To estimate the $\beta$ values, use the principle of least squares. In other words find the $\beta$ values that make the predicted values, $\hat Y_i$ as close as possible to the original $Y_i$.

The residual, $e_i$, is the deviation between the actual data value and what you predict based on the k independent variables.
:::

## Sum of squares

-   "Bad" regression results.
    -   All $\hat Y_i$ close  to each other
    -   Few $\hat Y_i$ close  to $Y_i$
-   "Good" regression results.
    -   The $\hat Y_i$ are spread out
    -   Many $\hat Y_i$ close to $Y_i$
    
::: notes

Let's talk about a bad regression result. The word "bad" is a bit judgmental. You might substitute "disappointing" but that is also troublesome. Maybe "not too useful" is what I really mean.

Let me give a hypothetical example. Suppose you paid a statistician a fortune to produce a linear regression model that predicted the average length of stay for heart attack patients of a particular age with particular systolic and diastolic blood pressure. 

An 84 year old patient with 130/90 blood pressure? The model predicts 2.7 days length of stay.

A 35 year old patient with 120/80 blood pressure? The model predicts 2.4 days length of stay.

A 69 year old patient with 190/140 blood pressure? The model predicts 2.6 days length of stay.

You ask yourself, why are we bothering collecting information about age and blood pressure. The predictions are pretty much all the same.

Another way of looking at this is that a "bad" or "not too useful" regression model is one where the predictions are not very close to the actual data values.

In contrast a "good" or "useful" regression model is one where the predictions change a lot for patients with different characteristics. A wide spread of the predicted values means that you have information that can help you. The first patient's predicted length of stay is 6.8, the second is 1.5 and the third is 3.1? That's useful information.

Another way of thinking about a "good" regression model is that it's predictions are generally pretty close to the actual values.

The analysis of variance table helps you identify "good" and "bad" regression models.

:::

## Analysis of variance table for multiple linear regression, 1 of 3

-   SSR or $SS_{regression}=\Sigma(\hat Y_i-\bar Y)^2$
-   SSE or $SS_{error}$ or $SS_{residual}=\Sigma(Y_i-\hat Y_i)^2=\Sigma\ e_i^2$
-   SST or $SS_{total}=\Sigma(Y_i-\bar Y)^2$

::: notes
The first column in the analysis of variance table is the sum of squares. The regression sum of squares is a measure of how spread out the predicted values ($\hat Y_i$) are.The error sum of squares is a measure of how close the predicted values are to the data. The total sum of squares is the variation of the data itself.
:::

## Analysis of variance table for multiple linear regression, 2 of 3

-   $df_{regression}=k$
-   $df_{error}=n-k-1$
-   $df_{total}=n-1$
    -   $MS=SS/df$

::: notes
The degrees of freedom for regression is k, the number of independent variables. The degrees of freedom for error and total are n-k-1 and n-1 respectively. You divide the sum of squares by the degrees of freedom to get the mean squares.
:::

## Analysis of variance table for multiple linear regression, 3 of 3

-   $F=MSR/MSE$
-   This tests the hypotheses
    -   $H_0:\ \beta_1=\beta_2=\ldots=\beta_k=0$
    -   $H_1:\ \beta_j \ne 0$ for at least one j
    -   Accept $H_0$ is F is close to 1
    -   Reject $H_0$ is F is much larger than 1
    
::: notes
You compute the F ratio by dividing the mean square for regression into the mean square for error. Think of the numerator as signal and the denominator as noise. If this ratio is close to 1, then you should accept the null hypothesis.
:::

## Example using fat data

```{r}
#| label: 02-02-setup
#| message: false
#| warning: false

library(broom)
library(glue)
library(tidyverse)
load("../data/module02.RData")
```

```{r}
#| label: anova

anova2 <- anova(m0, m2)
anova2
SSR <- format(round(anova2[2,4], 0), big.mark=",")
SSE <- format(round(anova2[2,2], 1), big.mark=",")
SST <- format(round(anova2[1,2], 1), big.mark=",")
F = round(anova2[2,5], 2)
```

-   SSR = `r SSR`
-   SSE = `r SSE`
-   SST = `r SST`
-   F = `r F`
-   You should report the p-value as p < 0.001

::: notes
Here is an example using the dataset with body fat measurements and circumference measurements at the hip, chest, and abdomen. The sum of squares are 10,500 for regression, 4,500 for residual or error and 15,000 for total. The respective degrees of freedom are 3 (for the three independent variables), 248 (252-3-1) for error and 251 (252-1) for total. The mean squares are quite different 3,500 versus 18. The F-ratio, 192, is a lot larger than 1 and the small p-value indicates that you should reject the null hypothesis and conclude that at least one of the three regression slopes is not equal to zero.

Note that this test does not tell you which variable or variables are needed to predict body fat. It is an overall test of the combined effect of all three variables.
:::

## R-squared

-   $R^2=SSR/SST$ or $1-(SSE/SST)$
    -    Proportion of explained variation
    -    1 - proportion of unexplained variation

::: notes
R-squared is defined pretty much the same way as in simple linear regression (regression with only one independent variable). It is the ratio of explained variation ($SS_{regression}$) to total variation ($SS_{total}$) or 1 minus the ratio of unexplained variation ($SS_{error}$) to total variation. This is a descriptive measure and there is no rule of thumb, but you are happier with the quality of the regression predictions if R-squared is closer to 1. 
:::

## Example using the fat data

```{r}
#| label: 02-02-rsq

m2_glance <- data.frame(glance(m2))
m2_glance
r.squared <- m2_glance$r.squared
rsq <- glue("{round(100*r.squared)}%")
```

-   $R^2$ = `r SSR` / `r SST` = `r r.squared`
    -   Alternately $R^2$ = 1- (`r SSE` / `r SST`)
-   Round and express as a percentage
    -   $R^2$ = `r rsq`

::: notes
You can compute R-squared from the analysis of variance table. The sum of squares for regression is 10,548.480 and the sum of squares total is 15,079.017. When you are calculating, please use the maximum precision for any intermediate calculation, but then round aggressively with the final result.

Often you will see R-squared displayed as a percentage instead of a proportion (e.g., 70% instead of 0.70).
:::

## Adjusted $R^2$

-   $1-\frac{MSE}{MST}$ or 
-   $1-\frac{SSE/(n-k)}{SST/(n-1)}$
    -   Field textbook suggests a more complex formula
    -   Penalizes for model complexity (but not enough)
    
::: notes
Many researchers propose an alternative measure, adjusted R-squared. This statistic modifies R-squared slightly downward. You can view this as a penalty for model complexity, but unfortunately, the penalty is small and does help enough to discourage the use of needlessly complex regression models.
:::
