---
title: "Interactions in analysis of variance"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

```{r}
#| label: 04-04-setup
#| message: false
#| warning: false

library(tidyverse)
load("../data/module04.RData")
```

## Group means for tracking data

```{r}
#| label: trial4-interaction-table

trial4_interaction_table
```

::: notes
The data used here is an exercise in motor skills where a subject is asked to maintain contact with a small area on a rotating disk. There were both male and female subjects and a wide range of ages. I decided to classify age into child=Yes (18 and younger) and child=NO (over 18).

The means for this study are interesting. There is almost no difference between male and female children. Adults do better than children, but the gap is much larger when comparing male adults to male children.
:::

## Line plot of tracking data

```{r}
#| label: trial4-plot1

trial4_plot1
```

::: notes
The two lines are right on top of one another for child=Yes and diverge for child=No. So there is a gender gap, but only for adults.
:::

## Alternative line plot of exercise data

```{r}
#| label: trial4-plot2

trial4_plot2
```

::: notes
This is an alternate display of the interaction. The chld=Yes line is flat indicating no change due to gender. The child=No line is above the child=Yes line, indicating superior performance for adults versus children. The gap is wider for males than for females.

You should try both graphs and then display only one. Display the one that portrays the clearest picture of what is going on.
:::

## Alternative line plot of exercise data

```{r}
#| label: test

anova_trial4_table
```

## When you can't estimate an interaction

-   Special case, n=1
    -   Only one observation for categorical combination
    
::: notes
There is a special case where you have two categorical independent variables and you cannot estimate an interaction. If you have n=1, exactly one observation for each combination of your two categorical variables, then you don't have enough degrees of freedom to estimate an interaction and still be able to test whether that interaction is statistically significant.

It's sort of like that old joke I told about married life (it's okay but you lose a degree of freedom). Interactions cause an even bigger loss of degrees of freedom and in the case with only one observation per combination of categories, you lose enough degrees of freedom that it is not marriage, it being in prison.
:::
