---
title: "Review one factor analysis of variance"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

## Review oneway analysis of variance

-   $H_0:\ \mu_1=\mu_2=...=\mu_k$
-   $H_1:\ \mu_i \ne \mu_j$ for some i, j
    -   Reject $H_0$ if F-ratio is large
-   Note: when k=2, use analysis of variance or t-test

::: notes
In Biostats-1, we discussed the comparison of three or more means using oneway or single factor analysis of variance. You can actually use analysis of variance when comparing only two means, but an equivalent alternative is the t-test.
:::

## Full moon data

```{r}
#| label: 04-01-setup
#| message: false
#| warning: false

library(tidyverse)
load("../data/module04.RData")
```


-   Admission rates to mental health clinic
    -   Before, during, and after full moon.
-   One year of data, starting in August
-   Consult the [data dictionary][ref11] for more details

[ref11]: https://github.com/pmean/data/blob/main/files/full-moon-er-admissions.yaml

::: notes
To illustrate oneway analysis of variance, I found a dataset on mental health clinic admissions.
:::

## Line graph of full moon data

```{r}
#| label: plot-01

plot_01
```

::: notes
For interpretation of this output from R and all other output from R included below, refer to the file simon-5502-04-demo.
:::

## Alternative line graph of full moon data

```{r}
#| label: plot-02

plot_02
```

## Descriptive statistics

```{r}
#| label: table-01

table_01
```

## Analysis of variance table

```{r}
#| label: table-02

table_02
```

## Parameter estimates

```{r}
#| label: table-03

table_03
```

## Tukey post hoc

```{r}
#| label: table-04

table_04
```

