---
title: "Checking assumptions of analysis of variance"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

```{r}
#| label: 04-03-setup
#| message: false
#| warning: false

library(tidyverse)
load("../data/module04.RData")
```

## Assumptions

-   Normality
-   Equal variances
-   Independence
-   Note: No linearity assumption
    -   Only for linear regression and analysis of covariance

::: notes
The assumptions for multiple factor analysis of variance are no different than single factor analysis of variance. You must use residuals to check the assumptions of normality and equal variances. The assumption of independence is usually assessed qualitatively.
:::

## Q-Q plot of residuals

```{r}
#| label: 04-03-qq-residuals-plot

qq_residuals_plot
```

::: notes
For the moon analysis, the normality assumption appears to be satisfied.
:::

## Residual versus predicted value plot

```{r}
#| label: 04-03-predicted-residuals-plot

predicted_residuals_plot
```

::: notes
The homogeneity assumption also appears to be satisfied. A violation of homogeneity is often a fanning out patter where you see more variation for large predicted values than for small predicted values.
:::

## Diagnostic measures that are not needed

-   Variance inflation factor
-   Leverage
-   Studentized deleted residuals
-   Cook's distance

::: notes
There are some diagnostic measures that are important in multiple linear regression that are not at all needed for multiple factor analysis of variance. When your independent variables are categorical, there is little opportunity for collinearity to appear. So the variance inflation factor is not normally computed for multiple factor analysis of variance.

While studentized deleted residuals could be used, they almost never deviate substantially from the plain resisuals, so there is no need to look at them.

It is also difficult to produce outliers among the independent variables because categorical data can't have extreme values. In theory, it could happen if the distribution of data within certain categories was extremely unbalanced. The amount of imbalance, though, before this becomes an issue is never seen in real world datasets.

For the same reason, Cook's distance, which is a measure that combines leverage with studentized deleted residuals, is never used in multiple factor analysis of variance.
:::