---
title: "Template for 5501-01 programming assignment"
author: "Steve Simon"
format: 
  html:
    embed-resources: true
date: 2024-08-18
---

# Start part 1 ----------------------------------------------------------------

## File details

This program was written by Steve Simon on 2025-02-23 and is placed in the public domain. You can use this program any way you please.

This is the correct program for the Fall 2026 semester.

There is one data file used in this program

-   albuquerque-housing, [Comma delimited file][ref-alb-dat], [dictionary][ref-alb-yaml]

[ref-alb-data]: https://github.com/pmean/datasets/blob/master/albuquerque-housing.csv
[ref-alb-yaml]: https://github.com/pmean/datasets/blob/master/albuquerque-housing.yaml

This code is placed in the public domain.

## Load the tidyverse library

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)

R.version.string
Sys.Date()
```

#### Comments on the code

This code includes several [chunk options][ref-chunk]. Every program chunk should include a brief label. The first chunk is traditionally named setup and it loads the appropriate libraries, and prints general information about the program.

The next two chunk options suppress messages and warnings. This simplifies the output, but if you need to track down an unusual error, you might want to change one or both of these options to true.

The [library function][ref-library] loads a package, extra programs that are not part of the base R package. For most of your programs, you should load the [tidyverse library][ref-tidyverse].

[R.version.string][ref-version] is an internal scalar that lists the version of R that is used to create your output. 

The [Sys.Date function][ref-date] lists the date that the output was created.

[ref-chunk]: https://quarto.org/docs/computations/r.html#chunk-options
[ref-library]: https://r-pkgs.org/introduction.html
[ref-tidyverse]: https://tidyverse.org/
[ref-version]: https://stat.ethz.ch/R-manual/R-devel/library/base/html/Version.html
[ref-date]: https://stat.ethz.ch/R-manual/R-devel/library/base/html/Version.html

## Read the data and view a brief summary

```{r}
#| label: read

alb <- read_csv(
  file="../data/albuquerque-housing.csv",
  col_names=TRUE,
  col_types="nnnnccc",
  na=".")
glimpse(alb)
```

#### Comment on the code

The [read_csv function][ref-read-csv] is part of the [readr library][ref-readr] which is included in tidyverse. to read the data. The [glimpse function][ref-glimpse] will produce a brief summary.

[ref-read-csv]: https://readr.tidyverse.org/reference/read_delim.html
[ref-readr]: https://readr.tidyverse.org/index.html
[ref-glimpse]: https://tidyverse.org/

## Calculate the average price

```{r}
#| label: mean-price

alb |>
  summarize(price_mean=mean(price)) |>
  gt()
```

The summarize function computes a variety of statistics. By default, the statistics are computed for all observations in the dataset.
## Summarize price

The average price of a home, 106 thousand dollars, is quite low because the data comes from 1993.

## Summarize sqft

## Summarize age

## Summarize features

