---
title: "simon-5501-01-demo"
author: "Steve Simon"
format: 
  html:
    embed-resources: true
date: 2024-08-18
---

## File details

This program was written by Steve Simon on 2025-02-23 and is placed in the public domain. You can use this program any way you please.

This is the correct program for the Fall 2026 semester.

There is one data file used in this program

-   albuquerque-housing, [Comma delimited file][ref-alb-data], [dictionary][ref-alb-yaml]

[ref-alb-data]: https://github.com/pmean/datasets/blob/master/albuquerque-housing.csv
[ref-alb-yaml]: https://github.com/pmean/datasets/blob/master/albuquerque-housing.yaml

This code is placed in the public domain.

## Load the tidyverse library

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)

R.version.string
Sys.Date()
```

#### Comments on the code

This code includes several [chunk options][ref-chunk]. Every program chunk should include a brief label. The first chunk is traditionally named setup and it loads the appropriate libraries, and prints general information about the program.

The next two chunk options suppress messages and warnings. This simplifies the output, but if you need to track down an unusual error, you might want to change one or both of these options to true.

The [library function][ref-library] loads a package, extra programs that are not part of the base R package. For most of your programs, you should load the [tidyverse library][ref-tidyverse].

[R.version.string][ref-version] is an internal scalar that lists the version of R that is used to create your output. 

The [Sys.Date function][ref-date] lists the date that the output was created.

[ref-chunk]: https://quarto.org/docs/computations/r.html#chunk-options
[ref-library]: https://r-pkgs.org/introduction.html
[ref-tidyverse]: https://tidyverse.org/
[ref-version]: https://stat.ethz.ch/R-manual/R-devel/library/base/html/Version.html
[ref-date]: https://stat.ethz.ch/R-manual/R-devel/RHOME/library/base/html/Sys.time.html

## Read the data and view a brief summary

```{r}
#| label: read

alb <- read_csv(
  file="../data/albuquerque-housing.csv",
  col_names=TRUE,
  col_types="nnnnccc",
  na=".")
glimpse(alb)
```

#### Comments on the code

The [read_csv function][ref-read-csv] is part of the [readr library][ref-readr] which is included in tidyverse. The col_names argument tells R whether the first line of the file lists the variable names. The col_types argument uses the letters "n" and "c" to tell R which variables are numeric variables and which are character (string) variables. The na argument tells R which code is used to represent missing values. The [glimpse function][ref-glimpse] will produce a brief summary of a data frame or tibble.

[ref-read-csv]: https://readr.tidyverse.org/reference/read_delim.html
[ref-readr]: https://readr.tidyverse.org/index.html
[ref-glimpse]: https://dplyr.tidyverse.org/reference/glimpse.html

## Calculate the average price

```{r}
#| label: mean-price

alb |>
  summarize(price_mean=mean(price))
```

#### Comments on the code

The [summarize function][ref-summarize] computes a variety of statistics. By default, the statistics are computed for all observations in the dataset. The [mean function][ref-mean] computes the arithmetic mean or the average.

[ref-mean]: https://stat.ethz.ch/R-manual/R-devel/library/base/html/mean.html
[ref-summarize]: https://dplyr.tidyverse.org/reference/summarise.html

#### Interpretation of the output

The average price of a home, 106 thousand dollars, is quite low because the data comes from 1993.

## Summarize sqft

```{r}
#| label: mean-sqft

alb |>
  summarize(sqft_mean=mean(sqft))
```

#### Interpretation of the output

Place your interpretation here. Your interpretation here and below should mention the units of measurement. It should also comment on the size of the mean. Round the mean to two or three significant digits. Tell your audieice if the mean seems reasonable or if it appears to be much lower/higher than you might have expected. Use your judgment here, but even if I disagree when you describe a mean as "reasonable" or "low" or "high", I will not deduct any points.  The only time you will lose points is if you leave out the interpretation.

## Summarize age

```{r}
#| label: mean-age

alb |>
  summarize(age_mean=mean(age, na.rm=TRUE))
```

#### Comments on the code

The default option in R is to designate a missing value for a mean (or any other simple descriptive statistic) when the the variable has one or more missing values. This is a perfectly logical approach. You may, however, be interested in what the mean would be if the missing values were removed from the variable. The na.rm argument tells R whether to exclude missing values or not before calculating the mean.

#### Interpretation of the output

Place your interpretation here.

## Summarize features

```{r}
#| label: mean-features

alb |>
  summarize(features_mean=mean(features))
```

#### Interpretation of the output

Place your interpretation here.

