---
title: "04-03, R code for scatterplots and boxplots"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

```{r data0a}
library(glue)
library(readr)
library(stringr)
library(yaml)
f1 <- "https://raw.githubusercontent.com/pmean"
f2 <- "data/main/files/fev.yaml"
dd <- read_lines(glue("{f1}/{f2}"))
# if I am not online use the line below
# dd <- read_lines("../../../../data/files/fev.yaml")
h <- str_which(str_trim(dd), "^$")
chunk_beg <- c(1, h+1)
chunk_end <- c(h, length(dd))
```

## Data dictionary for fev, 1

```{r data1a}
cat(dd[chunk_beg[1]:chunk_end[1]], sep="\n")
```

```{r data2a}
cat(dd[chunk_beg[2]:chunk_end[2]], sep="\n")
```

```{r data3a}
cat(dd[chunk_beg[3]:chunk_end[3]], sep="\n")
```

```{r data4a}
cat(dd[chunk_beg[4]:chunk_end[4]], sep="\n")
```

## Data dictionary for fev, 2

```{r data5a}
cat(dd[chunk_beg[5]:chunk_end[5]], sep="\n")
```

```{r data6a}
cat(dd[chunk_beg[6]:chunk_end[6]], sep="\n")
```

```{r data7a}
cat(dd[chunk_beg[7]:chunk_end[7]], sep="\n")
```

```{r data8a}
cat(dd[chunk_beg[8]:chunk_end[8]], sep="\n")
```

```{r data9a}
cat(dd[chunk_beg[9]:chunk_end[9]], sep="\n")
```

:::notes

:::

## Data dictionary for fev, 3

```{r data10a}
cat(dd[chunk_beg[10]:chunk_end[10]], sep="\n")
```

```{r data11a}
cat(dd[chunk_beg[11]:chunk_end[11]], sep="\n")
```

:::notes
This is a small dataset with eight rows and three columns.
:::

## Data dictionary for fev, 4

```{r data12a}
cat(dd[chunk_beg[12]:chunk_end[12]], sep="\n")
```

```{r data13a}
cat(dd[chunk_beg[13]:chunk_end[13]], sep="\n")
```

```{r data14a}
cat(dd[chunk_beg[14]:chunk_end[14]], sep="\n")
```

:::notes
The variables are measurements before and after a major overhaul of the air conditioning system. The units are colonies per cubic foot of air. A pump pushes a certain volume of air through a filter and then bacterial colonies are allowed to grow on that filter.
:::

## simon-5501-04-fev.qmd, 1

```{r write0a}
fn <- "simon-5501-04-fev.qmd"
template <- read_lines(file=fn)
h <- str_which(template, "##")
chunk_beg <- c(1, h)
chunk_end <- c(h-1, length(template))
```

```{r write1a}
cat(template[chunk_beg[1]:chunk_end[1]], sep="\n")
```

## simon-5501-04-fev.qmd, 2

```{r write2a}
cat(template[chunk_beg[2]:chunk_end[2]], sep="\n")
```

## simon-5501-04-fev.qmd, 3

```{r write3a}
cat(template[chunk_beg[3]:chunk_end[3]], sep="\n")
```

## simon-5501-04-fev.qmd, 4

```{r write4a}
cat(template[chunk_beg[4]:chunk_end[4]], sep="\n")
```

## simon-5501-04-fev.qmd, 5

```{r write5a}
cat(template[chunk_beg[5]:chunk_end[5]], sep="\n")
```

## simon-5501-04-fev.qmd, 6

```{r write6a}
cat(template[chunk_beg[6]:chunk_end[6]], sep="\n")
```

## simon-5501-04-fev.qmd, 7

```{r write7a}
cat(template[chunk_beg[7]:chunk_end[7]], sep="\n")
```

## simon-5501-04-fev.qmd, 8

```{r write8a}
cat(template[chunk_beg[8]:chunk_end[8]], sep="\n")
```

## simon-5501-04-fev.qmd, 9

```{r write9a}
cat(template[chunk_beg[9]:chunk_end[9]], sep="\n")
```

## simon-5501-04-fev.qmd, 10

```{r write10a}
cat(template[chunk_beg[10]:chunk_end[10]], sep="\n")
```
## simon-5501-04-fev.qmd, 11

```{r write11a}
cat(template[chunk_beg[11]:chunk_end[11]], sep="\n")
```
