---
title: "Bivariate visualizations"
format: 
  html:
    embed-resources: true
editor: source
execute: 
  error: true
---

## File details

This program is for internal use, but you are welcome to look things over.

This program displays scatterplots, boxplots, and bar charts. It was written by Steve Simon on 2026-09-15 and is placed in the public domain.

## Load the tidyverse library

```{r}
#| label: setup
#| message: false
#| warning: false

library(MASS)
library(tidyverse)
```

## Functions for drawing

```{r}
#| label: correlation

library(tidyverse)

illustrate_correlation <- function(rho) {
  set.seed(123)
  mu2 <- 50
  mu1 <- 15
  s2 <- 10
  s1 <- 3
  hypothetical_covariance <- 
    matrix(
  	  c(s1^2,
  	    s1*s2*rho,
  	    s1*s2*rho, 
  	    s2^2),
  	    ncol = 2)
  mvrnorm(
    n = 1000, 
    mu = c(mu1, mu2), 
    Sigma = hypothetical_covariance) |>
    data.frame() -> bvn
	
  r <- round(cor(bvn$X1, bvn$X2), 2)
  nice_correlation <- paste0("r = ", r)
  
  bvn |>
    ggplot(aes(X1, X2)) + 
	    geom_point() +
	    labs(
        x="X1 (hypothetical data)",
        y="X2 (Hypothetical data)",
        title=nice_correlation,
        caption="Simon, 2026-09-15") +
  	  expand_limits(
  	  	y=c(20, 80),
  	  	x=c( 5, 25))
}
```

```{r}
#| label: strong-positive

illustrate_correlation(0.8)
ggsave("../images/correlation-01.png")
```

## An example of a weak positive relationship

```{r}
#| label: weak-positive
illustrate_correlation(0.5)
ggsave("../images/correlation-02.png")
```

## An example of a little or no relationship

```{r}
#| label: no-relationship

illustrate_correlation(0.1)
ggsave("../images/correlation-03.png")
```


## An example of a weak negative relationship

```{r}
#| label: weak-negative

illustrate_correlation(-0.5)
ggsave("../images/correlation-04.png")
```

## An example of a strong negative relationship

```{r}
#| label: strong-negative

illustrate_correlation(-0.8)
ggsave("../images/correlation-05.png")
```


