---
title: "Analysis of relationships in pulmonary data"
format: 
  html:
    embed-resources: true
editor: source
---

## File details

This program assesses the relationships among variables in a study of pulmonary function in children. There is a [data dictionary][dd] that provides more details about the data. The program was written by Steve Simon on 2024-09-07 and is placed in the public domain.

[dd]: https://github.com/pmean/datasets/blob/master/fev.yaml

This is the correct version of this program for Fall semester 2026

## Libraries

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
```

## List variable names

```{r}
#| label: names

pulmonary_names <- c(
    "age",
    "fev",
    "ht",
    "sex",
    "smoke")
```

## Reading the data

```{r}
#| label: read

pulmonary <- read_csv(
  file="../data/fev.csv",
  col_names=pulmonary_names,
  col_types="nnncc")
glimpse(pulmonary)
```

## Calculate mean, standard deviation for fev

```{r}
#| label: descriptive-fev

pulmonary |>
  summarize(
    fev_mean=mean(fev),
    fev_stdv=sd(fev))
```

#### Interpretation of the output

The mean fev is 2.6 liters and the standard deviation is 0.87 liters. I am not an expert on pulmonary function, but these values do not appear to be unusually large or small.

## Calculate mean, standard deviation for age

```{r}
#| label: descriptive-age

pulmonary |>
  summarize(
    age_mean=mean(age),
    age_stdv=sd(age))
```

#### Interpretation of the output

The mean age is small, 9.9 years, and the standard deviation, 3.0 implies a tightly packed distribution. This is consistent with a pediatric population.

## Calculate counts for smoke

```{r}
#| label: descriptive-smoke

pulmonary |>
  count(smoke) |>
  mutate(total=sum(n)) |>
  mutate(pct=100*n/total)
```

#### Comments on the code

The count function produces a count for each level of smoke. The first mutate function calculates a total count for the entire dataset. The second mutate function calculates a percentage.

#### Interpretation of the output

Almost all of the subjects (90% or 589 out of 654) were non-smokers.

## Plot age versus fev

```{r}
#| label: plot-age-fev

pulmonary |>
  ggplot() +
	  aes(age, fev) +
    geom_point() +
    geom_smooth() +
    labs(
      x="Age (years)",
      y="Forced Expiratory Volume (Liters)",
      caption="Steve Simon, 2024-09-07")
```

#### Comments on the code

The [geom_point][ref-point] function adds points on the X and Y axes to create a scatterplot. The [geom_smooth][ref-smooth] function adds a smoothing curve to graph.

[ref-point]: https://ggplot2.tidyverse.org/reference/geom_point.html
[ref-smooth]: https://ggplot2.tidyverse.org/reference/geom_smooth.html

#### Interpretation of the output

The is a positive relationship between age and fev, though things seems to level off around 14 years.

## Correlation between age and fev

```{r}
#| label: correlation

with(pulmonary, cor(age, fev))
```

#### Comments on the code

The cor function calculates the correlation between age and fev. The with function identifies pulmonary as the dataset where you can find age and fev.

#### Interpretation of the output

The correlation, 0.75, which supports the plot drawn earlier by showing a strong positive association between age and fev.

## Plot smoke versus fev

```{r}
#| label: plot-smoke-fev
#| fig.height: 1.5

pulmonary |>
  ggplot() +
    aes(fev, smoke) +
    geom_boxplot() +
    labs(
      y="Smoker (Yes/No)",
      x="Forced Expiratory Volume (Liters)",
      caption="Steve Simon, 2024-09-07")
```

#### Comments on the code

The [geom_boxplot function][ref-boxplot] produces a set of boxplots. The categorical variable is clearly smoke, so placing it on the y-axis produces a horizontal set of boxplots. 

[ref-boxplot]: https://ggplot2.tidyverse.org/reference/geom_boxplot.html

#### Interpretation of the output

The fev values are larger for smokers versus non-smokers. This is the opposite direction from what we expected.

## Means and standard deviations for smokers and non-smokers.

```{r}
#| label: mean-fev-by-smoke

pulmonary |>
  group_by(smoke) |>
  summarize(
    mean_fev=mean(fev),
    sd_fev=sd(fev))
```

#### Comments on the code

The [group_by function][ref-group-by], part of the dplyr and tidyverse libraries, tells R that any further calculations should be done separately for each level of smoke.

[ref-group-by]: https://dplyr.tidyverse.org/reference/group_by.html

#### Interpretation of the output

The average fev values is 3.3 for smokers and much smaller, 2.6, for non-smokers. This is also opposite from what we expected. The standard deviations, 0.82 and 0.86, are roughly equal.