---
title: "Analysis of relationships in pulmonary data"
format: 
  html:
    embed-resources: true
editor: source
---

This program assesses the relationships among variables in a study of pulmonary function in children. There is a [data dictionary][dd] that provides more details about the data. The program was written by Steve Simon on 2024-09-07 and is placed in the public domain.

[dd]: https://github.com/pmean/datasets/blob/master/fev.yaml

## Libraries

The tidyverse library is the only one you need for  this program.

```{r setup}
#| message: false
#| warning: false
library(tidyverse)
```

## List variable names

Since the variable names are not listed in the data file itself, you need to list them here.

```{r names}
pulmonary_names <- c(
    "age",
    "fev",
    "ht",
    "sex",
    "smoke")
```

## Reading the data

Here is the code to read the data and show a glimpse. 

```{r read}
pulmonary <- read_csv(
  file="../data/fev.csv",
  col_names=pulmonary_names,
  col_types="nnncc")
glimpse(pulmonary)
```

## Calculate mean, quartiles, range for fev

```{r descriptive-fev}
summary(pulmonary$fev)
sd(pulmonary$fev)
```

The mean fev is 2.6 liters and the standard deviation is 0.84 liters. The fev values range from 0.8 to 5.8. I am not an expert on pulmonary function, but these values appear to be reasonable.

## Calculate mean, quartiles, range for age

```{r descriptive-age}
summary(pulmonary$age)
sd(pulmonary$age)
```

The mean age is 9.9 years. The youngest subject is 3 years old and the oldest is 19. This is consistent with a pediatric population.

## Calculate counts for smoke

```{r descriptive-smoke}
pulmonary |>
  count(smoke) |>
  mutate(total=sum(n)) |>
  mutate(pct=round(100*n/total))
```

Almost all of the subjects (90% or 589 out of 654) were non-smokers.

## Plot age versus fev

```{r plot-age-fev}
pulmonary |>
  ggplot(aes(age, fev)) +
    geom_point() +
    geom_smooth() +
    xlab("Age (years)") +
    ylab("Forced Expiratory Volume (Liters)") +
    ggtitle("Plot drawn by Steve Simon on 2024-09-07")
```

## Correlation between age and fev

```{r correlation}
cor(pulmonary$age, pulmonary$fev)
```

The correlation, 0.75, and the plot both show a strong positive association between age and fev.

## Plot smoke versus fev

```{r plot-smoke-fev}
pulmonary |>
  ggplot(aes(smoke, fev)) +
    geom_boxplot() +
    xlab("Smoker (Yes/No)") +
    ylab("Forced Expiratory Volume (Liters)") +
    ggtitle("Plot drawn by Steve Simon on 2024-09-07")
```

The fev values are larger for smokers versus non-smokers. This is the opposite direction from what we expected.

## Means and standard deviations for smokers and non-smokers.

```{r mean-fev-by-smoke}
pulmonary |>
  group_by(smoke) |>
  summarize(
    mean_fev=mean(fev),
    sd_fev=sd(fev))
```

The average fev values is 3.1 for smokers and much smaller, 2.6, for non-smokers. This is also opposite from what we expected. The standard deviations, 0.82 and 0.86, are roughly equal.