---
title: "Analysis of gardasil shots by demographic factors"
format: 
  html:
    embed-resources: true
---

## File details

This program reads data on Gardasil vaccinations in young women. Find more information in the [data dictionary][dd].

[dd]: https://raw.githubusercontent.com/pmean/datasets/master/gardasil.yaml

The program was written by Steve Simon on 2024-09-07 and is placed in the public domain.

## Load the tidyverse library

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
```

## Read the data and view a brief summary

```{r}
#| label: read

gard <- read_csv(
  file="../data/gardasil.csv",
  col_names=TRUE,
  col_types="nnnnnnnnnn")
names(gard) <- tolower(names(gard))
glimpse(gard)
```

## Create factors for agegroup

The factor function identifies a variable as categorical and assigns labels to number codes. You don't necessarily need to use factor if the data you read in is character strings, as R automatically treats those variable as categorical.

```{r}
#| label: agegroup-1

gard$agegroup <- factor(
  gard$agegroup,
  levels=0:1,
  labels=c(
    "11 to 17 years",
    "18 to 26 years"))
```

#### Comments on the code

The [factor function][ref-factor], part of base R, assigns labels to categories. This is most useful for numeric codes. You can also use factor for to spell out the full meaning for letter codes or abbreviations. A final application of factor is to control the order in which the categories are displayed.

[ref-factor]: https://stat.ethz.ch/R-manual/R-devel/library/base/html/factor.html

## Counts and percentages for agegroup

```{r}
#| label: agegroup-2

gard |> 
  count(agegroup) |>
  mutate(total=sum(n)) |>
  mutate(pct=100*n/total)
```

#### Interpretation of the output

There are roughly the same number of patients 11 to 17 years as there are patients 18 to 26 years.

## Create factors for shots

It is a bit silly to replace 1, 2, 3 with One, Two, Three. The main reason is to clearly identify shots as categorical rather than continuous.

```{r}
#| label: shots-1

gard$shots <- factor(
  gard$shots,
  levels=1:3,
  labels=c(
    "One", 
    "Two", 
    "Three"))
```

## Counts and percentages for shots

```{r}
#| label: shots-2

gard |> 
  count(shots) |>
  mutate(total=sum(n)) |>
  mutate(pct=100*n/total)
```

Slightly more patients got three shots than one or two shots, but this is still less than half of the patients overall.

## Compare number of shots by age group

```{r}
#| label: shots-by-age-1

gard |>
  count(agegroup, shots) |>
  group_by(agegroup) |>
  mutate(row_total=sum(n)) |>
  mutate(pct=100*n/row_total)
```

#### Comments on the code

As mentioned in the earlier demo program, the group_by function will require that any further calculations should be done separately for each level of agegroup. This means that the sum function creates a separate total for each agegroup.

#### Interpretation of the output

The probability of getting all three shots was higher in the 11 to 17 year old group compared to the 18 to 26 year old group.

## Bar chart of shots by age group

```{r}
#| label: shots-by-age-2

gard |>
  ggplot(aes(x=agegroup, fill=shots)) +
    geom_bar(position="fill") +
    labs(
      x="Age group",
      y="Proportion",
      caption="Steve Simon, 2024-09-07")
```

#### Interpretation of the output

The bar plot shows the sameprobability of getting all three shots was higher in the 11 to 17 year old group compared to the 18 to 26 year old group.

