---
title: "Analysis of gardasil shots by demographic factors"
format: 
  html:
    embed-resources: true
---

This program reads data on Gardasil vaccinations in young women. Find more information in the [data dictionary][dd].

[dd]: https://raw.githubusercontent.com/pmean/datasets/master/gardasil.yaml

The program was written by Steve Simon on 2024-09-07 and is placed in the public domain.

## Load the tidyverse library

For most of your programs, you should load the tidyverse library. The messages and warnings are suppressed.

```{r setup}
#| message: false
#| warning: false
library(tidyverse)
```

## Read the data and view a brief summary

Use the read_csv function to read the data. The glimpse function will produce a brief summary. Use tolower to convert uppercase to lowercase.

```{r read}
gard <- read_csv(
  file="../data/gardasil.csv",
  col_names=TRUE,
  col_types="nnnnnnnnnn")
names(gard) <- tolower(names(gard))
glimpse(gard)
```

## Create factors for agegroup

The factor function identifies a variable as categorical and assigns labels to number codes. You don't necessarily need to use factor if the data you read in is character strings, as R automatically treats those variable as categorical.

```{r agegroup-1}
gard$agegroup <- factor(
  gard$agegroup,
  levels=0:1,
  labels=c(
    "11 to 17 years",
    "18 to 26 years"))
```

## Counts and percentages for agegroup

```{r agegroup-2}
gard |> 
  count(agegroup) |>
  mutate(total=sum(n)) |>
  mutate(pct=round(100*n/total))
```

There are roughly the same number of patients 11 to 17 years as there are patients 18 to 26 years.

## Create factors for shots

It is a bit silly to replace 1, 2, 3 with One, Two, Three. The main reason is to clearly identify shots as categorical rather than continuous.

```{r shots-1}
gard$shots <- factor(
  gard$shots,
  levels=1:3,
  labels=c(
    "One", 
    "Two", 
    "Three"))
```

## Counts and percentages for shots

```{r shots-2}
gard |> 
  count(shots) |>
  mutate(total=sum(n)) |>
  mutate(pct=round(100*n/total))
```

Slightly more patients got three shots than one or two shots, but this is still less than half of the patients overall.

## Compare number of shots by age group

```{r shots-by-age-1}
gard |>
  count(agegroup, shots) |>
  group_by(agegroup) |>
  mutate(row_total=sum(n)) |>
  mutate(pct=round(100*n/row_total))
```

## Bar chart of shots by age group

```{r shots-by-age-2}
gard |>
  ggplot(aes(x=agegroup, fill=shots)) +
    geom_bar(position="fill") +
    xlab("Age group") +
    ylab("Proportion") +
    ggtitle("Plot drawn by Steve Simon on 2024-09-07")
```

The probability of getting all three shots was higher in the 11 to 17 year old group compared to the 18 to 26 year old group.

