---
title: "Review of simple tests of two proportions"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

## Comparing two binary outcomes

-   Is there a difference in the proportion of deaths between male passengers and female passengers on the Titanic?
-   Is there difference in the proportion of patients finishing the full three doses of HPV vaccine between Black women and White women?
-   Does using a ng tube for feeding in pre-term infants increase the probability of successful breast feeding at six months?

::: notes
Most of the statistics, you have seen so far involve a continuous outcome. You can, however, use a binary outcome. Here are three examples comparing a binary outcome between two groups.
:::

## Other comparisons involving a binary outcome

-   Is there are difference in the proportion of deaths between first class, second class, and third class passengers?
-   Does age influence the proportion of women finishing the full three doses of HPV vaccine?
-   Controlling for the mother's age, does using a ng tube for feeding in pre-term infants increase the probability of successful breast feeding at six months?

::: notes
Here are some more complex comparisons involving a binary outcome. The first example involves a comparison of three proportions, not two. The next example involves a continuous predictor of a binary outcome. The final example involves a comparison of binary outcomes in two groups, but controlling for a third variable.
:::

## Hypothesis framework

-   $H_0:\ \pi_1=\pi_2$
-   $H_1:\ \pi_1=\pi_2$
-   Compute $\hat p_1$ and $\hat p_2$ from samples
-   Accept $H_0$ if $\hat p_1-\hat p_2$ is close to zero.
    -   $T=(\hat p_1-\hat p_2)/s.e.$
    -   95% CI: $(\hat p_1-\hat p_2) \pm Z_{\alpha/2}s.e.$

::: notes
The hypothesis to test two proportions uses the symbols $\pi_1$ and $\pi_2$ to represent the proportions in a population.
:::

## The Titanic dataset

```{r}
#| label: setup-1
#| message: false
#| warning: false

library(broom)
library(epitools)
library(glue)
library(magrittr)
library(tidyverse)

load("../data/module06.RData")
```

```{r}
#| label: 06-01-glimpse

glimpse(ti)
```

## Counts and percentages

```{r}
#| label: 06-01-counts

sv$table1
sv$prop1
```

## Test for difference in proportions

```{r}
#| label: 06-01-test-diff

sv$prop_test
```

## Chi-square test of independence, 1 of 2

-   Equivalent to test of two proportions
-   Lay out data in two by two table

\begin{matrix}
          & No\ event & Event \\
Treatment & O_{11}    & O_{12}\\
Control   & O_{21}    & O_{22}
\end{matrix}

::: notes

:::

## Chi-square test of independence, 2 of 2

\begin{matrix}
          & No\ event                 & Event \\
Treatment & E_{11} = n_1 (1-\hat p_.) & E_{12}=n_1 \hat p_.\\
Control   & E_{21} = n_2 (1-\hat p_.) & E_{22}=n_2 \hat p_.
\end{matrix}

-   $X^2=\Sigma \frac{(O_{ij}-E_{ij})^2}{E_{ij}}$

::: notes

:::

## Expected counts for Titanic

Observed counts

```{r}
#| label: 06-01-observed

sv$table1
```


Expected counts
```{r}
#| label: 06-01-expected

chisq.test(sv$table1) |>
  extract2("expected")
```

## Chisquare test for Titanic

```{r}
#| label: 06-01-chisq

sv$chisq
```

## Odds ratio calculation

```{}
       No event  Event  Odds
Group1    a         b
Group2    c         d
```

-   Odds for group 1 = $b/a$
-   Odds for group 2 = $d/c$
-   Odds for group 1 = $\frac{d/c}{b/a} = \frac{ad}{bc}$
-   s.e.(log or) = $\sqrt{\frac{1}{a} + \frac{1}{b} + \frac{1}{c} + \frac{1}{d}}$


## Titanic data

```{}
       Survived   Died  Total
Female   308      154     462
Male     142      709     851
Total    450      863   1,313
```

::: notes

:::

## Titanic data, odds of death

```{}
       Survived   Died  Total  Odds
Female   308      154     462  2     to 1 against
Male     142      709     851  4.993 to 1 in favor
Total    450      863   1,313

Odds ratio = 4.993 / 0.5 = 9.986
```

::: notes
Clearly, a male passenger on the Titanic was more likely to die than a female passenger. But how much more likely? You can compute the odds ratio or the relative risk to answer this question.

The odds ratio compares the relative odds of death in each group. For females, the odds were exactly 2 to 1 against dying (154/308=0.5). For males, the odds were almost 5 to 1 in favor of death (709/142=4.993). The odds ratio is 9.986 (4.993/0.5). There is a ten fold greater odds of death for males than for females.
:::


## Odds ratio for survival by sex

```{r}
#| label: 06-01-or

sv$or
```
