---
title: "Splines, part 03"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

```{r}
#| label: 03-setup
#| message: false
#| warning: false

library(splines)
library(tidyverse)
```

## 06-03 B-splines

-   B-splines = Basis splines
    -   Basis is a mathematical property of a set and a subset
-   Mathematically equivalent to formulation shown earlier
-   Much less collinearity

::: notes

B-splines provide a solution with less issues of multi-collinearity.

:::

```{r}
#| label: 03-b-splines

b_spline <- bs(xm$c2, knots=c(5, 10, 15), degree=3, intercept=TRUE)
```

```{r}
#| label: 03-bplot

n <- ncol(b_spline)
bplot <- list(rep(NULL, n))
for (j in 1:n) {
  dat <- cbind(xm$c2, b_spline[ , j])
  display_curve(dat, "Cubic B-spline terms") |>
    display_dots(dat[1:21, ]) -> bplot[[j]]
}
```

## 06-03 B-spline, 1

```{r}
#| label: 03-bplot-1

bplot[[1]]
```

::: notes


The individual columns represent piecwise cubic polynomials. As I cycle through these graphs, notice how they are concentrated in certain intervals and there is only a partial overlap between these intervals. Also notice how they transition smoothly to zero outside those intervals.


:::

## 06-03 B-spline, 2

```{r}
#| label: 03-bplot-2

bplot[[2]]
```

## 06-03 B-spline, 3

```{r}
#| label: 03-bplot-3

bplot[[3]]
```

## 06-03 B-spline, 4

```{r}
#| label: 03-bplot-4

bplot[[4]]
```

## 06-03 B-spline, 5

```{r}
#| label: 03-bplot-5

bplot[[5]]
```

## 06-03 B-spline, 6

```{r}
#| label: 03-bplot-6

bplot[[6]]
```

## 06-03 B-spline, 7

```{r}
#| label: 03-bplot-7

bplot[[7]]
```

## 06-03 Correlation of B-spline terms

```{r}
#| label: 03-cor

b_spline |>
  data.frame() |>
  set_names(paste0("S", 1:7)) |>
  data.frame() |>
  cor() |>
  round(2)
```

::: notes

Although there is some correlation, this is not nearly as bad as the piecewise approach.

:::

## 06-03 B-spline fit to the artificial data

```{r}
#| label: 03-b-spline-fit

lm(xm$y~b_spline) %>%
  augment(newdata=xm) %>%
  select(c2, .fitted) %>%
    display_curve("Cubic spline fit to the data") %>%
    display_dots(simulated_example)
```

