---
title: "simon-5502-06-extra-data"
format: 
  html:
    embed-resources: true
editor: source
execute: 
  error: false
---

# --- Part 1

## File details

This program was written by Steve Simon on 2026-02-22 and is placed in the public domain. You can use this program any way you please.

This program reads in some files from DASL (Data and Story Library) and produces a few preliminary plots. Some of these files might be good for student projects.

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)

R.version.string
Sys.Date()
```

## Read Mazda data

```{r}
#| label: read-mazda

prices0 <- read_tsv(file="https://gksmyth.github.io/ozdasl/oz/mazdas.txt")
```

```{r}
#| label: plot-price

prices0 |>
  mutate(age=91-Year) |>
  ggplot() +
  aes(x=age, y=Price) +
  geom_point()
```

```{r}
#| label: plot-log-price

prices0 |>
  mutate(age=91-Year) |>
  mutate(log_price=log2(Price)) |>
  ggplot() +
  aes(x=age, y=log_price) +
  geom_point()
```

## Read dexterity data

```{r}
#| label: read-dexterity

dexterity <- read_tsv(file="https://dasl.datadescription.com/download/data/3253")
glimpse()

```{r}
#| label: plot-dexterity

dexterity |>
  ggplot() +
  aes(`Age(yr)`, Speed) +
  geom_point()
```

## Movie

```{r}
#| label: read-movie

movie <- read_tsv(file="https://dasl.datadescription.com/download/data/3347")
glimpse(movie)
```

```{r}
#| label: plot-movie

movie |>
  ggplot() +
  aes(x=Budget, y=USGross....) +
  geom_point()
```

## Marriage

```{r}
#| label: read-marriage

marriage <- read_tsv(file="https://dasl.datadescription.com/download/data/3329")
glimpse(marriage)

marriage |>
  mutate(gap=Men-Women) |>
  ggplot() +
  aes(x=Year, y=gap) +
  geom_point()
```

