---
title: "Module04 homework answers"
author: "Steve Simon"
date: "Created 2020-03-28"
output: html_document
---

1. Create factors for AgeGroup, Race, Completed. You do not have to create factors for the remaining categorical variables. Show frequency counts for all of these variables.

First, read in the data. I downloaded the tab separated value file from the web and stored it in the data folder with the name gardasil.tsv.

```{r read-gardasil}
raw_data <- read.delim("../data/gardasil.tsv", header=TRUE, sep="\t")
head(raw_data)
```

The last two columns are empty, so store the first ten in a new data frame.

```{r first-ten}
gar <- raw_data[ , 1:10]
```

Create factors with the factor function.

```{r create-factors}
gar$age_factor <- 
  factor(gar$AgeGroup, 
    labels=c(
      "child", 
      "adult"))
table(gar$age_factor)

gar$race_factor <- 
  factor(gar$Race, 
    labels=c(
      "white", 
      "black", 
      "Hispanic", 
      "other/unknown"))
table(gar$race_factor)

gar$completed_factor <-
  factor(gar$Completed,
    labels=c(
      "no", 
      "yes"))
table(gar$completed_factor)

```

2. What percentage of patients completed all three shots?

```{r completed}
prop.table(table(gar$completed_factor))
```

About 33% of the patients completed all three shots.

3. Use a crosstabulation to compare Age to AgeGroup. Are the recodings into AgeGroup done properly?

```{r age-by-agegroup}
table(gar$Age, gar$AgeGroup)
```

No, there is an inconsistency in how 18 year olds were classified.

4. Does the likelihood of completing all four shots vary by AgeGroup or Race? Calculate the appropriate percentages.

```{r likelihood-by-agegroup}
prop.table(table(gar$age_factor, gar$completed_factor), margin=1)
```

Children are slightly more likely to complete all shots.

```{r likelihood-by-race}
prop.table(table(gar$race_factor, gar$completed_factor), margin=1)
```

Blacks are slightly less likely to complete all shots compared to the other three groups.

5. Calculate frequency tables for both categorical variables in the dataset. Are there any missing values?

First, read in the data.

```{r read-diet}
raw_data <- read.delim("../data/diet.txt", header=TRUE, sep="\t")
head(raw_data)
```

The data look good as is. Store in a more descriptive variable name and calculated the frequencies of the two variables.

```{r diet-frequencies}
diet <- raw_data
table(diet$Bloat)
table(diet$Cracker)
```

6. Calculate a crosstabulation with nicely rounded percentages.  Note: do not add the "%" symbol, as it causes problems with R. Try to arrange the rows and columns and percentages so that your table best addresses the question, what type of cracker had the most and the least bloating.

```{r nice-percentages}
diet_counts <- table(diet$Cracker, diet$Bloat)
diet_proportions <- prop.table(diet_counts, margin=1)
diet_percentages <- round(100*diet_proportions)
diet_percentages
```

Gum had the most bloating and Bran had the least bloating.

Some students will use margin=2 instead of margin=1. It is not wrong per se, but the results are a bit trickier to interpret. Here is what you get.

```{r nice-percentages}
diet_counts <- table(diet$Cracker, diet$Bloat)
diet_proportions <- prop.table(diet_counts, margin=2)
diet_percentages <- round(100*diet_proportions)
diet_percentages
```

Notice that the percentages add up to 100 with each column rather than within each row. What this table tells you is that among patients reporting bloat, 10% were eating bran crackers, 29% combo, 19% control, and 43% gum. Since there were roughly the same number of patients in each group, this works out okay. But if you had more gum crackers, you wouldn't know if the differential percentages could be accounted for by this excess in crackers. It's a subtle but important issue.



Note that if you try to add a percent sign to the table, you lose the row and column names.

```{r lost-names}
paste0(diet_percentages, "%")
```

There is a way to get these back, but it is beyond the scope of this class. Here's the code, if anyone is curious.

```{r include-percent-sign}
cracker_names <- dimnames(diet_percentages)[[1]]
percentage_moderate_severe <- diet_percentages[ , 2]
count_moderate_severe <- diet_counts[ , 2]
count_crackers <- table(diet$Cracker)
colon <- ": "
percent <- "%"
comma <- ", "
slash <- "/"
fancy_percentages <- paste(
  cracker_names,
  colon,
  percentage_moderate_severe,
  percent,
  comma,
  count_moderate_severe,
  slash,
  count_crackers,
  sep="")
fancy_percentages
```