---
title: "Comments for MEDB 5501, Week 13"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: visual
---

## Covariance

-   $Cov(X,Y)=\frac{1}{n-1}\Sigma(X_i-\bar{X})(Y_i-\bar{Y})$
    -   $(X_i-\bar{X})(Y_i-\bar{Y})$ is positive if
        -   $X_i$ and $Y_i$ both above average
        -   $X_i$ and $Y_i$ both below average
    -   $(X_i-\bar{X})(Y_i-\bar{Y})$ is negative if
        -   $X_i$ above average and $Y_i$ below average
        -   $X_i$ below average and $Y_i$ above average

::: notes
Add note.
:::

## 

```{r}
library(tidyverse)
y <- 6 + 4*(1:6)
e <- c(4, -2, -6, 6, -6, 4)
x <- rev(y) + 2*e + 20
x <- x/2
y <- y/2
print(data.frame(x, y), row.names=FALSE)
xbar <- mean(x)
ybar <- mean(y)
sx <- round(sd(x), 1)
sy <- round(sd(y), 1)
```

$\ $
 $\bar{X}=`r xbar`$; $S_X=`r sx`$
 $\bar{Y}=`r ybar`$; $S_Y=`r sy`$

##

```{r}
data.frame(x, y) %>%
  ggplot(aes(x, y)) + 
    scale_y_continuous(
      expand=expansion(add=0.75),
      breaks=2*(0:10)) +
    scale_x_continuous(
      expand=expansion(add=2),
      breaks=10+2*(0:10)) -> base_plot
base_plot +
    geom_point(size=3)
```

## 

```{r}
zx <- paste0(ifelse(x-xbar>0, "+", ""), x-xbar)
zy <- paste0(ifelse(y-ybar>0, "+", ""), y-ybar)
lab <- paste0("(",zx, ",", zy, ")")
base_plot +
    geom_segment(
      x=mean(x), 
      y=min(y)-0.5,
      xend=mean(x), 
      yend=max(y)+0.5,
      linetype=2) +
    geom_segment(
      x=min(x)-1.5, 
      y=mean(y),
      xend=max(x)+1.5, 
      yend=mean(y),
      linetype=2) +
    geom_text(label=lab)
```

## Calculation of covariance

```{r}
d <- data.frame(
  x_centered=x-xbar, 
  y_centered=y-ybar, 
  product=(x-xbar)*(y-ybar))
print(d, row.names=FALSE)
total <- sum(d$product)
```

$\ $

-   $Cov(X,Y)=\frac{1}{5}(`r total`)=`r var(x,y)`$

## Correlation

-   $Corr(X,Y)=\frac{Cov(X,Y)}{S_XS_Y}$
    -   Also use $r_{XY}$
    -   Population correlation is $\rho_{XY}$

## Calculation of correlation

-   $r_{XY}=\frac{`r var(x,y)`}{`r sx` \times `r sy`}=`r cor(x,y)`$
    -   Always round!
        -   $r_{XY}=`r round(cor(x,y), 2)`$ or $`r round(cor(x,y), 1)`$

## Interpretation of correlation

-   r is always between -1 and +1
    -   Positive values imply positive association
    -   Negative values imply negative association
    -   Strongest associations closest to -1 or +1

## r between -1 and -0.7, strong negative association

```{r}
gen <- function(r, n=100) {
  x <- rnorm(n)
  x <- (x - mean(x))/sd(x)
  y <- rnorm(n)
  e <- resid(lm(y~x))
  e <- (e - mean(e))/sd(e)
  y <- sign(r)*(x+e*sqrt((1/r^2-1)))
  y <- (y- mean(y))/sd(y)
  data.frame(x, y) %>%
    ggplot(aes(x, y)) + 
      geom_point(shape=1, size=3) +
      expand_limits(x=c(-3,3), y=c(-3,3)) +
    ggtitle(paste("Example of data with r =", cor(x,y)))
}
```

:::: {.columns}
::: {.column width="50%"}
```{r}
#| fig-width: 4
#| fig-height: 4
gen(-0.9)
```
:::
::: {.column width="50%"}
```{r}
#| fig-width: 4
#| fig-height: 4
gen(-0.8)
```
:::
::::
  
## r between -0.7 and -0.3, weak negative association   

:::: {.columns}
::: {.column width="50%"}
```{r}
#| fig-width: 4
#| fig-height: 4
gen(-0.6)
```
:::
::: {.column width="50%"}
```{r}
#| fig-width: 4
#| fig-height: 4
gen(-0.4)
```
:::
::::

## r between -0.3 and +0.3, little or no association

:::: {.columns}
::: {.column width="50%"}
```{r}
#| fig-width: 4
#| fig-height: 4
gen(-0.1)
```
:::
::: {.column width="50%"}
```{r}
#| fig-width: 4
#| fig-height: 4
gen(0.1)
```
:::
::::

## r between +0.3 and +0.7, weak positive association   

:::: {.columns}
::: {.column width="50%"}
```{r}
#| fig-width: 4
#| fig-height: 4
gen(0.4)
```
:::
::: {.column width="50%"}
```{r}
#| fig-width: 4
#| fig-height: 4
gen(0.6)
```
:::
::::

## r between +0.7 and +1, strong positive association   

:::: {.columns}
::: {.column width="50%"}
```{r}
#| fig-width: 4
#| fig-height: 4
gen(0.8)
```
:::
::: {.column width="50%"}
```{r}
#| fig-width: 4
#| fig-height: 4
gen(0.9)
```
:::
::::

## Extreme case, perfect association

:::: {.columns}
::: {.column width="50%"}
```{r}
#| fig-width: 4
#| fig-height: 4
gen(1)
```
:::
::: {.column width="50%"}
```{r}
#| fig-width: 4
#| fig-height: 4
gen(-1)
```
:::
::::

## Extreme case, perfect negative association

```{r}
#| fig-width: 4
#| fig-height: 4
gen(-1)
```

