---
title: "b1-12-02, Relationship to linear regression"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

## Create indicator variables

```{r b1-12-02-setup}
#| message: false
#| warning: false
library(broom)
library(tidyverse)
hearing <- read_csv(
  "../data/hearing.csv",
  col_types="ncccn",
  col_names=TRUE)
hearing |>
  mutate(i_young=as.numeric(age=="young")) |>
  mutate(i_male=as.numeric(gender=="male")) |>
  select(
    age,
    gender,
    code,
    i_young,
    i_male,
    db)
```


:::notes
You need k-1 indicators for a categorical predictor that has k levels. In this simple example, that just means one indicator for age and one indicator for gender.
:::
  
## Two factor analysis of variance using aov

```{r b1-12-02-aov}
#| echo: true
m1 <- aov(db ~ age + gender, data=hearing)
anova(m1)
```

:::notes
Here is a repeat of the analysis of variance table using aov.
:::

## Two factor analysis of variance using linear regression, 1

```{r b1-12-02-lm-1}
#| echo: true
m2 <- lm(db ~ age + gender, data=hearing)
anova(m2)
```

:::notes
The analysis of variance table is identical when you use lm in place of aov.
:::

## Two factor analysis of variance using linear regression, 2

```{r b1-12-02-lm-2}
#| echo: true
tidy(m2)
```

:::notes
The intercept is the estimated db when both indicator variables equal zero. The first indicator, i_age, is the estimated average change in db when moving from female to male. The second indicator, i_gender, is the estimated average change in db when moving from old to young.
:::
