---
title: "Analysis of fruitfly data"
format: 
  html:
    embed-resources: true
---

This program reads data on fruit fly longevity. Find more information in the [data dictionary][dd].

[dd]: https://github.com/pmean/data/blob/master/files/fruitfly.yaml

This code was written by Steve Simon on 2024-10-29 and is placed in the public
domain.

## Load the tidyverse library

```{r setup}
#| message: false
#| warning: false
library(broom)
library(tidyverse)
```

#### Comments on the code

For most of your programs, you should load the tidyverse library. The broom 
library converts your output to a nicely arranged dataframe. The messages and
warnings are suppressed.

## List the variable names

```{r variable-list}
vlist <- c(
  "id",
  "partners",
  "type",
  "longevity",
  "thorax",
  "sleep")
```

#### Comments on the code

When a dataset does not have variables on the first line, you need to specify 
them in the code.

## Read the data and view a brief summary

```{r read}
fly <- read_fwf(
  "../data/fruitfly.txt",
  col_types="nnnnnn",
  fwf_widths(
    widths=c(2, 2, 2, 3, 5, 3),
    col_names=vlist))
glimpse(fly)
```

#### Comments on the code

The fruitfly dataset has a fixed width format (fwf). You need to specify the
columns that each variable uses.

Notice that the two categorical variables, partners and type, are actually
numbers rather than strings. To avoid having R treat these variables as if
they were continuous, use the factor function in some of the code below.

## Create a subset of fruitfly dataset

```{r subset}
fly |>
  filter(type != 9) -> fly_subset
```

#### Comments on the code

If you exclude the pure control group (No females), you can analyze the two
factors, partners and type individually. Partners has two category levels, 1 for
when one female was included in the cage and 8 for when eight females were 
included in the cage. Type also has two category levels, 0 for pregnant female
fly/flies and 1 for virgin fly/flies. A male fly will not mate with a pregnant
females, so you can think of this as a second level of controls. The two factors
are crossed, meaning that every possible combination of partners and type has 
outcomes measured.

## Draw boxplot of longevity by partners and type

```{r boxplots-partners-type}
#| fig.width: 6
#| fig.height: 2.5
fly_subset |>
  ggplot(aes(factor(partners), longevity, fill=factor(type))) +
    geom_boxplot() +
    xlab("Number of partners") +
    ylab("Lifespan in days") +
    ggtitle("Graphs drawn by Steve Simon on 2024-10-28") +
    labs(fill="Female fly type") +
    scale_fill_discrete(labels=c("Pregnant", "Virgin")) +
    coord_flip()
```

#### Interpretation of the output

The lifespans tend to be quite a bit shorter in cages with 8 virgin females.
There are no problems with the homogeneity and normality assumptions.

## Calculate average longevity by partners and type

```{r by-partners-and-type}
#| message: false
fly_subset |>
  group_by(type, partners) |>
  summarize(
    longevity_mn=mean(longevity),
    longevity_sd=sd(longevity),
    n=n()) -> fly_means
fly_means
```

#### Interpretation of the output

The means show the same pattern as noted above. The standard deviations are all
roughly the same.

## Draw line graph

```{r line-graph}
fly_means |>
  ggplot(aes(
    factor(partners), 
    longevity_mn, 
    group=factor(type),
    color=factor(type))) +
    geom_line(linewidth=2) +
	expand_limits(y=range(fly_subset$longevity)) +
  xlab("Number of partners") +
  ylab("Lifespan in days") +
  labs(color="Female fly type") +
  scale_color_discrete(labels=c("Pregnant", "Virgin"))
```

#### Interpretation of the output

This plot shows strong evidence of an interaction.

Male fruitflies do appear to have shorter average lifespans when sharing a cage
with virgin females compared to pregnant females, but this effect is far 
stronger when the number of female fruitflies is eight rather than one.

The line for pregnant (sharing a cage with pregnant females) is close to flat,
indicating that there is little impact of the number of pregnant females on
average longevity.

## Analysis of variance including the interaction

```{r anova-with-interaction}
m1 <- aov(longevity ~ factor(partners)*factor(type), data=fly_subset)
anova(m1)
```

#### Interpretation of the output

There is a statistically significant interaction between the number of female
partners and the type of partners (pregnant or virgin). When you have an
interaction, you do not normally interpret the individual categorical 
predictors.

## Using lm, 1

```{r lm-with-interactions-1}
m2 <- lm(longevity ~ factor(partners)*factor(type), data=fly_subset)
anova(m2)
```

#### Interpretation of the output

The analysis of variance table is identical when using lm instead of aov.

## Using lm, 2

```{r lm-with-interactions-2}
tidy(m2)
```

#### Interpretation of the output

The lifespan of fruitflies in cages with 8 virgin females is 17 days shorter 
than what you would expect. If the effect of type and partners were independent,
you'd expect to see a 9.4 (= 8.0 + 1.4) day decline on average. The interaction
shows that this cage is actually 16.6 days worse than that on average.

## Using lm, 3

```{r add-coefficients}
b <- coef(m2) 
fly_means |>
  mutate(b0=b[1]) |>
  mutate(b1=b[2]) |>
  mutate(b2=b[3]) |>
  mutate(b3=b[4]) |>
	select(-longevity_sd, -n) -> fly_coefficients
fly_coefficients[1, 5:7] <- 0
fly_coefficients[2, 6:7] <- 0
fly_coefficients[3, c(5,7)] <- 0
fly_coefficients
```

#### Comments on the code

The coef function takes the intercept and slopes from a linear regression model
and stores them in a vector. The mutate command adds the individual intercept
and slopes to separate columns. Finally, place zeros in places where the various
slopes are not needed.

#### Interpretation of the output

This table is not needed for the data analysis, but is included to illustrate 
how the regression coefficients are computed in a two factor analysis of 
variance with an interaction.

The intercept is the estimated average with both categories each to the "zero"
or reference level. In this case, the average lifespan of flies in a cage with
one pregnant female.

The first slope coefficient (b1) represents the estimated average difference in
lifespan between eight pregnant females and one pregnant females.

The second slope coefficient (b2) represents the estimated average difference in
lifespan between one pregnant females and one virgin female.

The third slope coefficient (b3) represents the interaction. It is a measure of
how much shorter the average lifespan is with eight virgin females above and
beyond the individual impact of eight versus one and virgin versus pregnant.

## Save important files for later use

```{r save}
save(
  fly,
  fly_subset,
  file="../data/fruitfly.RData")
```

#### Comments on the code

It is always good form to save the important pieces of your data analysis that
might be re-used sometime later.
