---
title: "Analysis of full-moon-er-admissions data"
format: 
  html:
    embed-resources: true
---

This program reads data on admissions to an emergency room before, during, and after a full moon. Find more information in the [data dictionary][dd].

[dd]: https://github.com/pmean/data/blob/master/files/full-moon-er-admissions.yaml

This code was written by Steve Simon on 2024-11-02 and is placed in the public domain.

## Load the tidyverse library

```{r}
#| label: setup
#| message: false
#| warning: false

library(broom)
library(tidyverse)
```

#### Comments on the code

For most of your programs, you should load the tidyverse library. The broom library converts your output to a nicely arranged dataframe. The messages and warnings are suppressed.

## Read the data and view a brief summary

```{r}
#| label: read

er <- read_tsv(
  "../data/full-moon-er-admissions.txt",
  col_types="ccn",
  col_names=TRUE)
names(er) <- tolower(names(er))
glimpse(er)
```

#### Comments on the code

This is a tab delimited file. I changed all the variable names to all lowercase. That's one less thing opportunity for inconsistency errors.

## Reorder the category levels, 1

```{r}
#| label: reorder-moon

er$moon1 <- factor(
  er$moon,
  levels=c(
    "Before",
    "During",
    "After"))
```

#### Comments on the code

If you let R order the categories, it will do it alphabetically. The factor function will place them in a more logical order.

## Reorder the category levels, 2

```{r}
#| label: reorder-month

er$month1 <- factor(
  er$month,
  levels=c(
    "Aug", "Sep", "Oct",
    "Nov", "Dec", "Jan",
    "Feb", "Mar", "Apr",
    "May", "Jun", "Jul"))
```

#### Comments on the code

Since the study started in August, it makes sense to order the months starting then. An alphabetical ordering (April first then August?) makes no sense here.

## Draw boxplot of admission versus moon

```{r}
#| label: boxplots-admission-vs-moon
#| fig.width: 6
#| fig.height: 2.0

er |>
  ggplot(aes(moon1, admission)) +
    geom_boxplot() +
    xlab("Moon phase") +
    ylab("ER admissions per day") +
    ggtitle("Graphs drawn by Steve Simon on 2024-11-02") +
    coord_flip()
```

#### Interpretation of the output

Although there are a few outliers, I would not worry too much about the normality assumption. It also looks like heterogeneity would not be too much of a concern.

## Draw boxplot of admission versus month

```{r}
#| label: boxplots-admission-vs-month
#| fig.width: 6
#| fig.height: 6

er |>
  ggplot(aes(admission, month1)) +
    geom_boxplot() +
	  labs(
	  	x="ER admissions per day",
	  	y="Month",
	  	caption="Steve Simon, 2024-11-02, CC0")
```

#### Interpretation of the output

Although there are some unusual patterns here, keep in mind that there are only two observations per month.

## Calculate average ER admissions versus moon

```{r}
#| label: by-moon

er |>
  group_by(moon1) |>
  summarize(
    admission_mn=mean(admission),
    admission_sd=sd(admission),
    n=n())
```

#### Interpretation of the output

There is a slightly higher average admission rate and slightly more vaiation during a full moon.

## Calculate average ER admissions versus month

```{r}
#| label: by-month

er |>
  group_by(month1) |>
  summarize(
    admission_mn=mean(admission),
    admission_sd=sd(admission),
    n=n())
```

#### Interpretation of the output

Er admissions are trending upward over time and there is some evidence of heterogeneity. Again, treat this with caution as there are only three observations per month.

## Analysis of variance

```{r}
#| label: anova

m1 <- aov(admission ~ moon1 + month1, data=er)
anova(m1)
```

#### Interpretation of the output

There is a borderline, but statistically significant difference in average admission rates during different phases of the moon. There is a statistically significant difference in average admission rates between some of the months.

## Using lm, 1

```{r}
#| label: lm-1

m2 <- lm(admission ~ moon1 + month1, data=er)
anova(m2)
```

#### Interpretation of the output

The analysis of variance table using lm is the same as the analysis of variance table using aov.

## Using lm, 2

```{r}
#| label: lm-2

tidy(m2)
```

#### Interpretation of the output

The intercept represents the estimated average admission rate before a full moon and during the month of August. The admission rate is 2.5 patients per day higher during a full moon compared to a full moon. This is small but probably still important. The difference in average admission rates is 0.54 patients per day higher after a full moon compared to before a full moon. The remaining estimates compare each month back to the first month (August).

## Using lm, 3

```{r}
#| label: lm-3

er |>
  group_by(moon1) |>
  summarize(admission_mn=mean(admission)) |>
  mutate(admission_before=first(admission_mn)) |>
  mutate(difference=admission_mn-admission_before)
```

#### Comments on the code

This table shows how to calculate the regression coefficients associated with phases of the moon. It is not really needed for the analysis, but helps illustrate how these coefficients are computed.

## Using lm, 4

```{r}
#| label: lm-4

er |>
  group_by(month1) |>
  summarize(admission_mn=mean(admission)) |>
  mutate(admission_aug=first(admission_mn)) |>
  mutate(difference=admission_mn-admission_aug)
```

#### Comments on the code

This table shows how to calculate the regression coefficients associated with months of the year.

## Save important files for later use

```{r}
#| label: save

save(
  er,
  file="../data/full-moon-er-admissions.RData")
```