---
title: "Indicator variable coding"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

```{r}
#| label: 03-02-setup
#| message: false
#| warning: false

library(broom)
library(glue)
library(tidyverse)

load("../data/module03.RData")
```

## Traditional t-test, 1

```{r}
#| label: 03-02-boxplot
#| fig-width: 6
#| fig-height: 2

plot_1
```

## Traditional t-test, 2

```{r}
#| label: 03-02-table

table_1
```

## Traditional t-test, 3

```{r}
#| label: 03-02-t-test

m1
```

## T-test using general linear model

```{r}
#| label: 03-02-lm

tidy(m2)
confint(m2)
```

## Indicator variable coding

-   Convert binary categories
    -   1 for "first" category
    -   0 for "second" category
        -   Choice of "first" is arbitrary
-   Can also be used for three or more categories

## Why use the general linear model?

-   Works for a variety of models
    -   Independent samples t-test
    -   Simple linear regression
    -   One factor analysis of variance
    -   Multiple linear regression
    -   Analysis of covariance
    -   Multi-factor analysis of variance

## Why use indicator variables

-   Simple interpretation
    -   Average indicator is a probability
-   Mix categorical and continuous independent variables

