---
title: "Module05 homework answers"
author: "Steve Simon"
date: "Created 2021-04-08"
output: html_document
---

Questions 1-5 refer to the Albuquerque dataset.

1. Convert the number codes for the categorical variables to factors. Show frequency tables for these variables after conversion. Are any values missing for these categorical variables?

```{r factors}
fn <- "../data/albuquerque-housing.txt"
al <- read.table(fn, header=TRUE, sep=" ")
head(al)
al$ne_factor <- 
  factor(
    al$ne,
    levels=0:1,
    labels=c("No", "Yes"))
table(al$ne_factor, useNA="always")

al$custom_factor <- 
  factor(
    al$cust,
    levels=0:1,
    labels=c("No", "Yes"))
table(al$custom_factor, useNA="always")

al$corner_factor <- 
  factor(
    al$cor,
    levels=0:1,
    labels=c("No", "Yes"))
table(al$corner_factor, useNA="always")
```

There are no missing values for any of these three variables.

2. Compute means and standard deviations for all the continuous variables. How many missing values are there for age and for tax?

```{r means}
mean(al$price, na.rm=TRUE)
sd(al$price, na.rm=TRUE)
sum(is.na(al$price))
```

3. Find the largest house (biggest square footage) in the data set. Is the largest house also the most expensive house?

4. Evaluate the relationship between PRICE and SQFT using a scatterplot. Do larger houses tend to cost more?

5. Show how much more expensive custom built homes are compared to standard homes using a boxplot and through calculation of the means and standard deviations.

Questions 6-7 refer to the Diamond dataset.

6. Show the relationship between carat size and price using a scatterplot.

```{r read-diamond}
fn <- "http://jse.amstat.org/v9n2/4Cdata.txt"
di <- read.table(fn)
names(di) <- c("carat", "color", "clarity", "certification", "price")
head(di)
```

7. Show the relationship between color and price and the relatonship between clarity and price using boxplots.