---
title: "Homework 02b"
author: "Steve Simon"
date: "Created 2020-02-26"
output: html_document
---

This program provides the answers to homework2a, written by the instructor, Steve Simon. If you are having trouble getting your program to work, compare what you did to my code. If you are still having problems after viewing my code, please contact me.


```{r version-and-current-date}
R.version.string
Sys.Date()
```                                                                               

Q1. Draw a histogram for total sleep and life span. Would a normal distribution (bell shaped curve) be a fair representation of the distribution of either variable?

There are two ways to do this. The [hist function](https://stat.ethz.ch/R-manual/R-devel/library/graphics/html/hist.html) is faster, but I like the geom_histogram function within ggplot2 better.

```{r histogram}
library(ggplot2)
ggplot(sleep, aes(x=TotalSleep)) +
  geom_histogram()
ggplot(sleep, aes(x=LifeSpan)) +
  geom_histogram()
```

Q2. Do animals who have high degrees of Predation, Exposure, or Danger sleep less? Show your results using a correlation coefficient and a plot with a trendline.
