---
title: "Homework 02a"
author: "Steve Simon"
date: "Created 2020-02-26"
output: html_document
---

This program provides the answers to homework2a, written by the instructor, Steve Simon. If you are having trouble getting your program to work, compare what you did to my code. If you are still having problems after viewing my code, please contact me.

First things first. It is always a good idea to list the version of R and the date that the program ran.

```{r version-and-current-date}
R.version.string
Sys.Date()
```

Load the file module02-datasets.RData and list its contents.

```{r read}
load("../data/module02-datasets.RData")
ls()
```

Q1. Notice that there is a huge range in body weight. Display the rows in the data frame associated with the smallest and the largest mammals.

Note that there are at least three ways to do this. This way is slow, but it works.

First, display what the maximum and minimum values are.

```{r summary}
summary(sleep$BodyWt)
```

Then identify the row number with the maximum and minimum values.

```{r identify-maximum-row}
max_row <- which(sleep$Body==max(sleep$BodyWt))
max_row
```

```{r identify-minimum-row}
min_row <- which(sleep$Body==min(sleep$BodyWt))
min_row
```

Then display the rows with the maximum and minimum values.

```{r display-maximum-row}
sleep[max_row, ]
```

```{r display-minimum-row}
sleep[min_row, ]
```

It is important to add some commentary here along the lines of "The largest animal in the dataset is the African Elephant and the smallest animal is the Lesser Short-tailed shrew."

If you are curious, you can get the answer a bit faster using the [which.max and which.min](https://stat.ethz.ch/R-manual/R-devel/library/base/html/which.min.html) functions. You can also use the [arrange function in the dplyr library](https://dplyr.tidyverse.org/reference/arrange.html). We did not cover any of this in the class, but if you read the help file, you can probably figure the alternatives out.

Q2. Which variables have missing data?

Again, there is more than one way to do this. The summary function provides a fast answer.

```{r summary-all-variables}
summary(sleep)
```

Non-dreaming, dreaming, and total sleep, lifespan, and gestation have missing values.

Another way you can does this, is by using the is.na function on each variable. 

Q3. Calculate the mean and standard deviation for TotalSleep.

```{r mean-and-stdev}
mean(sleep$TotalSleep, na.rm=TRUE)
sd(sleep$TotalSleep, na.rm=TRUE)
```

Don't forget the na.rm argument. The mean time for TotalSleep is 10.5 hours, with a standard deviation of 4.6.
