---
title: "Objects in R"
author: "Steve Simon"
date: "2021-03-14"
output: powerpoint_presentation
---

### Introduction

This is a very brief introduction to the basic obejcts in R.

```{r version-and-current-date}
R.version.string
Sys.Date()
```

### Scalars

```{r scalars}
scalar_example_1 <- 3
scalar_example_1
scalar_example_2 <- "R"
scalar_example_2
scalar_example_3 <- "3"
scalar_example_3
```

<div class="notes">

Single values in R are called scalars. The most common scalars are numbers and strings. Strings are always surrounded by quote marks.

There are other scalars that are important, such as dates

</div>

### Vectors

```{r vectors}
vector_example_1 <- c(1, 2, 3)
vector_example_1
vector_example_2 <- c("a", "b", "c")
vector_example_2
vector_example_3 <- c("a", 2)
vector_example_3
```

<div class="notes">

A combination of number or string scalars is called a vector. Use the c function to create vectors.

If you try to mix different scalar types, R will convert them to a common format.

</div>

### Naming vectors

```{r naming}
my_degrees <- c(BA=1977, MS=1978, PhD=1982)
my_degrees
my_name <- c(first_name="Stephen", middle_initial="D", last_name="Simon")
my_name
```

<div class="notes">

You can name the individual elements of a vector. This is not done that often, but the same naming concept will be very important for more complex objects.

</div>

### Matrices

```{r matrices-1}
matrix_example_1 <- cbind(c(1, 2, 3), c(4, 5, 6))
matrix_example_1
matrix_example_2 <- rbind(c(1, 2, 3), c(4, 5, 6))
matrix_example_2
```

<div class="notes">

A matrix is a combination of vectors, with the important restriction that all the vectors have to be the same type. You can't mix strings and numbers for example.

You can create a matrix using the cbind or rbind functions.

</div>

### Matrices

```{r matrices-2}
matrix_example_3 <- matrix(c(1, 2, 3, 4, 5, 6), nrow=2, ncol=3, byrow=TRUE)
matrix_example_3
```

<div class="notes">

The matrix function will reshape a vector into a matrix. There are lots of options in the matrix function that control the shape of the matrix and the order in which it is filled.

</div>

### Lists

```{r lists}
list_example_1 <- list(scalar_example_1, vector_example_2, matrix_example_3)
list_example_1
```

<div class="notes">

A list is a combination of objects of different types. It can include scalars, vectors, matrices, or other types of R objects that we have not seen yet. You can also include one list inside another list.

</div>

### List names

```{r list-names}
list_example_2 <- list(name=my_name, degrees=my_degrees, age=64)
list_example_2
```

<div class="notes">

A list can contain virtually anything, so using names is almost always recommended.

</div>

### Data frames

```{r data-frames}
data_frame_example_1 <- data.frame(vector_example_1, vector_example_2)
data_frame_example_1
```

<div class="notes">

A data frame is a combination of vectors. The vectors have to have the same length, but they can be different types.

</div>

### Naming data frame columns

```{r data-frames-naming}
data_frame_example_2 <- data.frame(c(1, 2, 3), c("a", "b", "c"))
data_frame_example_2
data_frame_example_3 <- data.frame(small_numbers=c(1, 2, 3), early_letters =c("a", "b", "c"))
data_frame_example_3
```

<div class="notes">

R will always name the columns in a data frame using common sense rules. Sometimes this leads to poor choices (common sense isn't always common sense). So it is strongly recommended that you specify names explicitly.

</div>