---
title: "Video examples from module04"
author: "Steve Simon"
date: "Created 2019-04-20"
output: html_document
---

This is a small program written in Rmarkdown to show how to work with a dataset with mostly categorical variables.

This program is in the public domain. You can use it any way that you please.

```{r version-and-current-date}
R.version.string
Sys.Date()
suppressMessages(
  suppressWarnings(
    library(tidyverse)))
```

## Titanic data set

### Read in the data

```{r read-titanic}
fn <- "../data/titanic_v00.txt"
ti <- read_tsv(fn,
  col_types="ccncn")
glimpse(ti)
```

### View first few rows

```{r display-first-few-rows}
print(ti, n=20)
```

Print without the names

```{r display-without-names}
print(ti[ , -1], n=20)
```

### view last few rows

```{r last-rows}
ti[1310:1313, -1]
```

### Who are the youngest and oldest passengers?

```{r min}
ti[which.min(ti$Age), ]
```

```{r max}
ti[which.max(ti$Age), ]
```

### Frequency counts and percentages

```{r counts}
PClass_counts <- table(ti$PClass)
PClass_counts
```

```{r percentages}
PClass_percents <- round(100*prop.table(PClass_counts))
paste0(PClass_percents,"%")
```

### Any missing values

```{r check-missing}
table(ti$PClass, useNA="always")
```

### Factors for survived

```{r survived-counts}
table(ti$Survived)
```

```{r create-factor}
ti$surv_factor <- factor(
  ti$Survived,
  levels=0:1,
  labels=c("No", "Yes"))
```

This gives the counts of those who survived and those who did not.

```{r surived-counts-factor}
table(ti$surv_factor)
```

### Crosstabulation

```{r crosstabs-1}
pclass_by_gender <-
  table(ti$PClass, ti$Sex)
pclass_by_gender
```

```{r crosstabs-2}
row_proportions <-
  prop.table(
    pclass_by_gender,
    margin=1)
row_proportions
```

### Bar chart

This plot shows the number of people in each passenger class.

```{r draw-bar-chart}
simple_barplot <-
  ggplot(ti, aes(x=PClass)) +
    geom_bar()
simple_barplot
```

### New categorical variables

Child is defined as less than 18 years.

```{r new-age-categories}
ti$child <- case_when(
  is.na(ti$Age) ~ "Unknown",
  ti$Age <=  0 ~ "Invalid",
  ti$Age <  18 ~ "Yes",
  ti$Age >= 18 ~ "No")
table(ti$child, useNA="always")
```

Quality check. Does the new variable have sensible age limits.

```{r quality-check-age}
range(ti$Age[ti$child=="No"])
range(ti$Age[ti$child=="Yes"])
table(ti$Age[ti$child=="Unknown"], useNA="always")
```
