---
title: "r-05-03, Boxplots"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

```{r}
#| message: false
#| warning: false
#| label: 05-03-setup

library(gt)
library(tidyverse)

load("../data/demo-05.RData")
# load("../data/module-05.RData")
``` 

## How to show relationships

-   Two continuous variables
    -   Scatterplot
-   Two categorical variables
    -   Crosstabulation
-   One categorical and one contnuous variable
    -   Boxplots

## Boxplots, 1

```{r}
#| fig.height: 2
#| fig.width: 6

plot_02
```

::: notes

If you have a categorical variable and a continuous variable, you show the relationship between them using a boxplot.

:::

## Illustration of a boxplot

```{r}
#| fig.height: 1.5
#| fig.width: 6
#| label: illustration

hypo_5 <- c(1, 2, 5, 7, 8, 10, 14, 20, 22, 24, 36, 54, 60)
qx <- quantile(hypo_5)

tibble(u=rep(" ", 13), v=hypo_5) |>
	ggplot() +
	aes(x=u, y=v) +
	geom_boxplot() +
	xlab(" ") +
	ylab(" ") +
	coord_flip() + scale_y_continuous(
		breaks=c(qx[1], qx[2], qx[3], qx[4], 36, 54, 60),
		minor=NULL,
		labels=c(
			"Min\nw/o outliers", 
			"25%", 
			"50%", 
			"75%", 
			"Max\nw/o outliers",
			"Outlier",
			"Outlier"))
```

## Clustered boxplots, 1

```{r}
#| fig.height: 3
#| fig.width: 6

plot_03
```

::: notes

If you have a second categorical variable, you can create a clustered box plot. It might look like this...

::::

## Clustered boxplots, 2

```{r}
#| fig.height: 3
#| fig.width: 6

plot_04
```

::: notes

... or it might look like this.

When deciding which way to cluster, consider the proximity principle. Ask yourself what comparison is most important. Then arrange your cluster so that the important comparison requires less distance.

In this clustered boxplots male smokers are close to male non-smokers. Female smoklers are close to female non-smokers. This makes sense if the most important thing to you is the impact of smoking and you want to examine differences separately in males and in females.

In the previous clustered boxplot, male smokers are close to female smokers and male non-smokers are close to female non-smokers. This makes sense if the most important thing to you is gender differences and you want to examine those differences in smokers and separately in non-smokers.

:::
