---
title: "r-05-01, The FEV data set"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

```{r}
#| label: 05-01-setup
#| message: false
#| warning: false

library(glue)
library(tidyverse)
library(yaml)

load("../data/demo-05.RData")
load("../data/module-05.RData")
``` 

## How do you characterize relationships?

+ Between two continuous variables
  + Correlations and scatterplots
+ Between two categorical variables
  + Crosstabulations
+ Between a continuous variable and a categorical variable
  + Boxplots
  
::: notes

In an earlier module, you saw datasets that had mostly continuous variables. If you wanted to examine the relationship between two continuous variables, you would look at correlations and scatterplots.

Then in a different module, you saw datasets that had mostly categorical variables. If you wanted to examine the relationship between two categorical variables, you would look at crosstabulations.

In this module, you will see datasets that have a mix of continuous and categorical variables. If you want to examine the relationship between a continuous variable and a categorical variable, you would use a boxplot.

:::

## FEV data

```{r}
#| label: 05-01-yaml

f1 <- "https://raw.githubusercontent.com"
f2 <- "pmean/data/refs/heads/main/files"
f3 <- "fev.yaml"

dd <- read_yaml(glue("{f1}/{f2}/{f3}"))
```

`r dd$description`

::: notes

The fev data set looks at pulmonary function in a group of children. The acronym FEV stands for Forced Expiratory Volume and represents how air you can blow out of your lungs.

This dataset has 654 rows and 5 variables: age (in years), fev (in liters), height (in inches), sex, and smoking status. Both sex and smoking status are categorical and use number codes. 

:::


## read in the FEV data set, glimpse

```{r}
#| label: 05-01-glimpse

glimpse(pulmonary)
```

::: notes

Here is a glimpse of the data.

:::

## Verify that there are no missing values

`r code[["missing"]]`

## Create factors

`r code[["factors"]]`

## Summary for continuous variables: age

`r code[["age"]]`

## Summary for continuous variables: fev

`r code[["fev"]]`

## Summary for continuous variables: ht

`r code[["ht"]]`

## Summary for categorical variables: sex

`r code[["sex"]]`

## Summary for categorical variables: smoke

`r code[["smoke"]]`
