---
title: "Demonstation program for module 05"
format: 
  html:
    embed-resources: true
execute: 
  error: true
editor: source
---

## File details

This program was written by Steve Simon on 2025-01-17 and is placed in the public domain. You can use this program any way you please.

This program reads data on a study of pulmonary function in a pediatric population. Refer to the [data dictionary][ref01] for a more detailed description.

[ref01]: https://github.com/pmean/data/blob/main/files/fev.yaml

```{r}
#| label: setup
#| message: false
#| warning: false
R.version.string
Sys.Date()
library(glue)
library(tidyverse)
```

#### Documentation of the intermediate datasets

-   pulmonary: original dataset as read in from fev.txt
-   pulmonary_1: create factors for smoke and sex

## Read the fev dataset

Put in some comments about where you found the fev dataset.

```{r}
#| label: read
fn <- "../data/fev.txt"
pulmonary <- read_fwf(
  file=fn,
  col_types="nnnnn",
  col_positions=fwf_cols(
    age=3,
    fev=7,
    ht=5,
    sex=4,
    smoke=6))
glimpse(pulmonary)
```

## Create factors

```{r}
#| label: factor
pulmonary |>
  mutate(
    smoke =
      factor(
        smoke,
        levels=0:1,
        labels=c("Non-smoker", "Smoker"))) |>
  mutate(
    sex = 
     factor(
       sex,
       levels=0:1,
       labels=c("Female", "Male"))) -> pulmonary_1
```

## Descriptive statistics, 1

```{r}
#| label: univariate-1
pulmonary_1 |>
  count(smoke) |>
  mutate(total=sum(n)) |>
  mutate(pct=round(100*n/total)) |>
  mutate(pct=glue("{pct}% ({total}/{n})")) |>
  select(-n, -total)
```

#### Comments on the code

You will notice a lot of repetition or near repetition in this and the following chunks. As a general rule of programming, if you end up with three or more repeats/near repeats in your code, you should consider using a loop or a function to simplify your code. I do not do so here because I don't want to introduce loops or functions in this module.

#### Interpretation of the output

Almost all of the patients are non-smokers.

## Descriptive statistics, 2

```{r}
#| label: univariate-2
pulmonary_1 |>
  count(sex)
  mutate(total=sum(n)) |>
  mutate(pct=round(100*n/total)) |>
  mutate(pct=glue("{pct}% ({total}/{n})")) |>
  select(-n, -total)
```

#### Interpretation of the output

Slightly more than half of the patients are male.

## Descriptive statistics, 3

```{r}
#| label: univariate-3
pulmonary_1 |>
  summarize(
    age_mn=round(mean(age), 1),
    age_sd=round(sd(age), 1),
    age_min=min(age),
    age_max=max(age))  
```

#### Comments on the code

You might consider rounding a bit more aggressively in this and the next couple of code chunks. It is a subjective decision. You are fine with any reasonable choice in your homework. You will not lose any points unless you forget to round altogether or if you display an excessive number of significant digits (four or more significant digits is almost always bad).

#### Interpretation of the output

The average patient is 10 years old and the variation is narrow, which makes sense for a pediatric study. The oldest patient is 19 years.

## Descriptive statistics, 4

```{r}
#| label: univariate-4
pulmonary_1 |>
  summarize(
    fev_mn=round(mean(fev), 2),
    fev_sd=round(sd(fev), 2),
    fev_min=min(fev),
    fev_max=max(fev))  
```

#### Interpretation of the output

Note that I am not an expert in this area. I encourage you to provide interpretations even if you are not that familiar with the data in your homework. Any reasonable interpretation is fine. You will only lose points if you fail to provide any interpretation. Also, please include units of measurement in any interpretations that you provide. An average birth weight of 5 is low if the units are pounds, but quite high if the units are kilograms.

The average forced expiratory volume is 2.6 liters, which is reasonable for a pediatric population. There is quite a bit of variation in these values which might be expected given the variation from 3 to 19 years in the ages of the patients.

Note: Although I hate acronyms, the use of FEV in place of forced expiratory volume is well understood in the medical community, even among those who do not specialize in pulmonology. I will rely on the acronym in the rest of this program.

## Descriptive statistics, 5

```{r}
#| label: univariate-5
pulmonary_1 |>
  summarize(
    ht_mn=round(mean(ht), 1),
    ht_sd=round(sd(ht), 1),
    ht_min=min(ht),
    ht_max=max(ht))  
```

#### Interpretation of the output

The average height is a bit more than 5 feet. The range is wide (slightly under 4 feet to slightly above 6 feet), again a reflection of the age ranges.

## Histograms, 1

```{r}
#| label: histograms-1
pulmonary_1 |>
  ggplot() +
    aes(x=age) +
    ggtitle("Graph drawn by Steve Simon on 2025-01-18") +
    xlab("Age in years") +
    geom_histogram(
      color="black",
      fill="white",
      binwidth=1)
```

#### Interpretation of the output

The ages are very slightly skewed with no outliers.

## Histograms, 2

```{r}
#| label: histograms-2
pulmonary_1 |>
  ggplot() +
    aes(x=fev) +
    ggtitle("Graph drawn by Steve Simon on 2025-01-18") +
    xlab("FEV (liters)") +
    geom_histogram(
      color="black",
      fill="white",
      binwidth=0.5)
```

#### Interpretation of the output

FEVs are skewed right, but there are no serious outliers.

## Histograms, 3

```{r}
#| label: histograms-3
pulmonary_1 |>
  ggplot() +
    aes(x=ht) +
    ggtitle("Graph drawn by Steve Simon on 2025-01-18") +
    xlab("Height (inches)") +
    geom_histogram(
      color="black",
      fill="white",
      binwidth=2)
```

#### Interpretation of the output

Heights are very slightly skewed to the left, but I would still call this close enough to a bell-shaped curve. There are no outliers.

## Draw boxplots, 1

```{r}
#| label: boxplot-1
#| fig-width: 6
#| fig-height: 2.5
pulmonary_1 |>
  ggplot() +
    aes(x=smoke, y=fev) +
    ggtitle("Graph drawn by Steve Simon on 2025-01-18") +
    xlab("Smoking status") +
    ylab("FEV (liters)") +
    coord_flip() +
    geom_boxplot()
```      

#### Comments on the code

The [geom_boxplot][ref02] function (part of the ggplot2/tidyverse pacakges) produces separate boxplot of a continuous variable y for each level of a categorical variable x. The variable x must be either a string or a factor. The [boxplot][ref03] function (part of the graphics/base libraries) will also produce boxplots.

[ref02]: https://ggplot2.tidyverse.org/reference/geom_boxplot.html
[ref03]: https://stat.ethz.ch/R-manual/R-devel/library/graphics/html/boxplot.html

#### Interpretation of the output

The FEV measurements are generally much higher in smokers than in non-smokers. This is a surprising result.

## Draw boxplots, 2

```{r}
#| label: boxplot-2
#| fig-width: 6
#| fig-height: 2.5
pulmonary_1 |>
  ggplot() +
    aes(x=smoke, y=age) +
    ggtitle("Graph drawn by Steve Simon on 2025-01-18") +
    xlab("Smoking status") +
    ylab("Age (years)") +
    coord_flip() +
    geom_boxplot()
```      

#### Interpretation of the output

The ages of the smokers tend to be much higher than the ages of the non-smokers. There are a few outliers in ages for the non-smoker group, but these are not too extreme.

The disparity in ages is not so surprising and might serve as an explanation of the previous boxplot. If older patients have higher fev values, then the effect of smoking is only caused by the disparity in ages.

## Calculate group means and standard deviations for fev

```{r}
#| label: fev-by-smoke
pulmonary_1 |>
  group_by(smoke) |>
  summarize(
    fev_mn=round(mean(fev), 2),
    fev_sd=round(sd(fev), 2))
```

#### Comments on the code

THe [group_by][ref04] function was discussed in an earlier module. In this program chunk, it allows you to quantify the difference in FEV between non-smokers and smokers.

#### Interpretation of the output

The average FEV value is 0.7 liters higher in smokers. The standard deviations are roughly comparable.

## Calculate group means and standard deviations for age

```{r}
#| label: age-by-smoke
pulmonary_1 |>
  group_by(smoke) |>
  summarize(
    age_mn=round(mean(age), 1),
    age_sd=round(sd(age), 1))
```

#### Interpretation of the output

The average age is 4 years higher in smokers. This disparity might account for the 0.7 liter difference in average FEV values.