---
title: "Demonstation program for module 05"
format: 
  html:
    embed-resources: true
execute: 
  error: true
editor: source
---

## File details

This program was written by Steve Simon on 2025-01-17 and is placed in the public domain. You can use this program any way you please.

This program reads data on a study of pulmonary function in a pediatric population. Refer to the [data dictionary][ref-fev-nodate] for a more detailed description.

[ref-fev-nodate]: https://github.com/pmean/data/blob/main/files/fev.yaml

```{r}
#| message: false
#| warning: false
#| label: setup

R.version.string
Sys.Date()

library(gt)
library(glue)
library(tidyverse)
```

#### Comments on the code

I am including a new library, [gt][ref-gt-nodate], that provides more professional looking displays of tables. I encourage you to use this library, but it is entirely optional. You will see a few simple examples of how gt works in the upcoming code.

[ref-gt-nodate]: https://gt.rstudio.com/reference/index.html

## Documentation of the intermediate datasets

-   pulmonary: original dataset as read in from fev.txt
-   pulmonary_1: create factors for smoke and sex

# --- Begin part 1 ---

## Peek at the FEV data

```{r}
#| label: peek

fn <- "../data/fev.txt"
read_lines(file=fn, n_max=10)
```

#### Interpretation of the output

The read_lines function (part of the readr/tidyverse libraries) is useful for displaying the first few lines of a text file. It will read in each line of a file and store it as a string. As such, it does not need any information about format or delimiters. It helps you decide what function you need for importing the data and with which arguments.

In this particular file, it could be tab delimited, but is more likely to be a fixed width format. If you count carefully, you will see that the first three columns represent the first variable, you need seven more columns for the second variable, and so forth.



## read in the FEV data set

```{r}
#| label: read

pulmonary <- read_fwf(file=fn,
  col_types="nnnnn",
  col_positions=fwf_cols(
    age=3, 
    fev=7,
    ht=5,
    sex=4,
    smoke=6))

glimpse(pulmonary)
```

## Verify that there are no missing values

```{r}
#| label: tbl-missing
#| tbl-cap: "Testing B"

pulmonary |>
  mutate(across(everything(), is.na)) |>
  summarize(across(everything(), sum)) -> table_01

table_01 |>
  gt() |>
  tab_header(md("**Missing value count**")) |>
  tab_source_note("Simon, 2025-04-20")
```
#### Comments on the code

The [across][ref-across-nodate] and the [everything][ref-everything-nodate] functions allow you to make transformations and summaries for every variable in the tibble or data frame.

This particular piece of code first converts every value in the entire tibble or data frame into a true/false value with true representing missing values. Then it sums the values for each variable. A true/false value gets converted to 1/0, so this ends up counting the number of true (i.e., missing) values.

The [tab_header][ref-header-nodate] and [tab_source_note][ref-source-nodate] functions place text at the top and bottom of a table, respectively. I suggest that you provide a very brief explanatory title at the top and your name and date at the bottom.

[ref-across-nodate]: https://tidyselect.r-lib.org/reference/everything.html
[ref-everything-nodate]: https://tidyselect.r-lib.org/reference/everything.html
[ref-header-nodate]: https://gt.rstudio.com/reference/tab_header.html
[ref-source-nodate]: https://gt.rstudio.com/reference/tab_source_note.html

#### Comments on the code

I am introducing a new package, gt, that produces publication ready tables. I want to encourage you to use gt, but it is entirely optional. The first couple of things to add are tab_header and tab_source_note. It is a good idea to produce a simple descriptive title and to list your name and date as a note at the end of the table.

#### Interpretation of the output

This output confirms that there are no missing values for any of the variables.

## Create factors

```{r}
#| label: factors

pulmonary |>
  mutate(smoke=factor(
    smoke, 
    levels=0:1,
    labels=c("nonsmoker", "smoker"))) |>
  mutate(sex=factor(
    sex, 
    levels=0:1,
    labels=c("female", "male"))) -> pulmonary_1
```


## Summary for continuous variables: age

```{r}
#| label: tbl-age
#| tbl-cap: "Testing B"

pulmonary_1 |>
  summarize(
    age_mean=mean(age),
    age_sd=sd(age),
    age_min=min(age),
    age_max=max(age)) -> table_02

table_02 |>
  gt() |>
  tab_header(md("**Descriptive statistics**")) |>
  tab_source_note("Simon, 2025-04-20") |>
  fmt_number(columns=1:2, n_sigfig=2)
```

#### Comments on the code

The [fmt_number][ref-number-nodate] function allows you to specify how numbers are displayed. The n_sigfig argument controls how many significant digits your numbers are rounded to. The columns argument restricts rounding to specific columns. 

[ref-number-nodate]: https://gt.rstudio.com/reference/fmt_number.html

#### Interpretation of the output

This is clearly a pediatric population.

## Summary for continuous variables: fev

```{r}
#| label: fev

pulmonary_1 |> 
  summarize(
    fev_mean=mean(fev),
    fev_sd=sd(fev),
    fev_min=min(fev),
    fev_max=max(fev)) -> table_03

table_03 |>
  gt() |>
  tab_header(md("**Descriptive statistics**")) |>
  tab_source_note("Simon, 2025-04-20") |>
  fmt_number(columns=1:2, n_sigfig=2)
```

#### Interpretation of the output

I am not an expert on FEV, but these values seem reasonable.

## Summary for continuous variables: ht

```{r}
#| label: ht

pulmonary_1 |> 
  summarize(
    ht_mean=mean(ht),
    ht_sd=sd(ht),
    ht_min=min(ht),
    ht_max=max(ht)) -> table_04

table_04 |>
  gt() |>
  tab_header(md("**Descriptive statistics**")) |>
  tab_source_note("Simon, 2025-04-20") |>
  fmt_number(columns=1:2, n_sigfig=2)
```

#### Interpretation of the output

Again, these seem to be reasonable values.

## Summary for categorical variables: sex

```{r}
#| label: sex

pulmonary_1 |>
  count(sex) |>
  mutate(total=sum(n)) |>
  mutate(pct=round(100*n/total)) |>
  mutate(pct=glue("{n}/{total} ({pct}%)")) -> table_05

table_05 |>
  gt() |>
  tab_header(md("**Descriptive statistics**")) |>
  tab_source_note("Simon, 2025-04-20")
```

#### Interpretation of the output

The patients are split almost evenly with slightly more males than females.

## Summary for categorical variables: smoke

```{r}
#| label: smoke

pulmonary_1 |>
  count(smoke) |>
  mutate(total=sum(n)) |>
  mutate(pct=round(100*n/total)) |>
  mutate(pct=glue("{n}/{total} ({pct}%)")) -> table_06

table_06 |>
  gt() |>
  tab_header(md("**Descriptive statistics**")) |>
  tab_source_note("Simon, 2025-04-20")
```

#### Interpretation of the output

Only a small fraction (10%) of the patients are smokers.

# --- Begin part 2 ---

## Group means, fev-by-sex

```{r}
#| label: means-fev-by-sex

pulmonary_1 |>
  group_by(sex) |>
  summarize(
    fev_mean=mean(fev),
    fev_sd=sd(fev),
    n=n()) -> table_07

table_07 |>
  gt() |>
  tab_header(md("**Group statistics**")) |>
  tab_source_note("Simon, 2025-04-20") |>
  fmt_number(columns=1:2, n_sigfig=2)
```

#### Comments on the code

#### Interpretation of the output

## Group means, fev-by-smoke

```{r}
#| label: means-fev-by-smoke

pulmonary_1 |>
  group_by(smoke) |>
  summarize(
    fev_mean=mean(fev),
    fev_sd=sd(fev),
    n=n()) -> table_08

table_08 |>
  gt() |>
  tab_header(md("**Group statistics**")) |>
  tab_source_note("Simon, 2025-04-20") |>
	fmt_number(columns=1:2, n_sigfig=2)
```

#### Interpretation of the output

## Group means, age-by-smoke

```{r}
#| label: means-age-by-smoke

pulmonary_1 |>
  group_by(smoke) |>
  summarize(
    age_mean=mean(age),
    age_sd=sd(age),
    n=n())  -> table_09

table_09 |>
  gt() |>
  tab_header(md("**Group statistics**")) |>
  tab_source_note("Simon, 2025-04-20") |>
	fmt_number(columns=1:2, n_sigfig=2)
```

#### Interpretation of the output

# --- Begin part 3 ---

## Scatterplot

```{r}
#| label: plot-1

pulmonary_1 |>
  ggplot() +
  aes(x=age, y=fev) +
  geom_point() +
  labs(
    title="Scatterplot",
    x="Age in years",
    y="Forced expiratory volume (liters)",
    caption="Simon, 2025-04-19") -> plot_01

plot_01
```

#### Inpterpretation of the output

There is a strong positive relationship between age and fev.

Recall that you use a scatterplot to examine the relationship between two continuous variables.

## Crosstabs

```{r}
#| label: sex-by-smoke

pulmonary_1 |>
  count(sex, smoke) |>
  group_by(sex) |>
  mutate(total=sum(n)) |>
  mutate(pct=round(100*n/total)) |>
  mutate(pct=glue("{n}/{total} ({pct}%)")) |>
  select(-n, -total) -> table_10

table_10 |>
  pivot_wider(
    names_from=smoke, 
    values_from=pct) |>
  ungroup() |>
  gt(rowname_col="sex") |>
  tab_header(md("**Crosstabulation**")) |>
  tab_source_note("Simon, 2025-04-20")
```

#### Interpretation of the output

Smoking is uncommon but occurs slightly more often with female patients.

## Boxplots

```{r}
#| fig.width: 6
#| fig.height: 1.5
#| label: smoke-by-fev

pulmonary_1 |>
  ggplot() +
  aes(x=smoke, y=fev) +
  geom_boxplot() +
  coord_flip() +
  labs(
    title="Boxplots",
    x="Smoking status",
    y="Forced expiratory volume (liters)",
    caption="Simon, 2025-04-19") -> plot_02

plot_02
  
```


#### Comments on the code

The [boxplot][ref-boxplot-nodate] function draws boxplots. Most of the time, you want to put two or more boxplots side by side. The categorical variable that defines what goes into each boxplot is designated as x and the continuous variable summarized by the boxplot is designated as y.

Boxplots, especially for a small number of levels of your categorical variable, tend to look nicer flipped horizontally and with a fairly small height. 

[ref-boxplot-nodate]: https://ggplot2.tidyverse.org/reference/geom_boxplot.html

#### Interpretation of the output

In a bit of a surprise, fev values tend to be higher for smokers than for non-smokers.

## Clustered boxplots, 1

```{r}
#| fig.width: 6
#| fig.height: 2.5
#| label: fev-by-smoke-sex

pulmonary_1 |>
  ggplot() +
  aes(x=smoke, y=fev, fill=sex) +
  geom_boxplot() +
  coord_flip() +
  labs(
    title="Boxplots",
    x="Smoking status",
    y="Forced expiratory volume (liters)",
    caption="Simon, 2025-04-19") -> plot_03

plot_03
```

#### Comments on the code

Specifying the fill argument tells ggplot that you want to further break down your boxplots by a second categorical variable, sex.

## Clustered boxplots, 2

```{r}
#| fig.width: 6
#| fig.height: 2.5
#| label: smoke-by-sex-smoke

pulmonary_1 |>
  ggplot() +
  aes(x=sex, y=fev, fill=smoke) +
  geom_boxplot() +
  coord_flip() +
  labs(
    title="Boxplots",
    x="Smoking status",
    y="Forced expiratory volume (liters)",
    caption="Simon, 2025-04-19") -> plot_04

plot_04
```

## Save everything

```{r}
#| label: save

warnings()

save.image("../data/demo-05.RData")
```

