---
title: "Programming template for module05"
author: "Steve Simon"
date: "Created 2022-05-02"
format: 
  html:
    embed-resources: true
editor: source
execute:
  error: true
---

## Program details

This program is written in R with quarto formatting. It shows how to work with a dataset with a mix of categorical and continuous variables.

This program was written by Steve Simon on 2022-05-02 with a major modification on 2024-12-23. This program is in the public domain. You can use it any way that you please.

```{r}
#| label: setup
#| message: false
#| warning: false
R.version.string
Sys.Date()
library(tidyverse)
```

## Read the pulmonary dataset

Put in some comments about where you found the pulmonary dataset.

```{r read-pulmonary}
#| label: read-pulmonary
pulmonary <- read_csv(
  file="../data/fev.csv",
  col_types="nnnnn",
  col_names=TRUE)
glimpse(pulmonary)
```

## Modify names

```{r}
#| label: modify-names
pulmonary |>
  rename(
    age=V1,
    fev=V2,
    ht=V3,
    sex=V4,
    smoke=V5) -> pulmonary
```

#### Comments about the code

The [rename function][ref01] changes the names of variables in your dataset. I can never remember, but the rule is that you list the new name on the left side of the equal sign and the old name on the right.

[ref01]: https://dplyr.tidyverse.org/reference/rename.html

## Create factor for sex

```{r}
#| label: create-sex-factor
pulmonary |>
  mutate(
    sex_factor=
      factor(
        sex,
        levels=0:1,
        labels=c("Female", "Male"))) -> pulmonary
```

## Create factor for smoke

```{r}
#| label: create-smoke-factor
pulmonary |>
  mutate(
    smoke_factor=
      factor(
        smoke,
        levels=0:1,
        labels=c("Non-smoker", "Smoker"))) -> pulmonary
```

## Get counts for sex and smoke

```{r}
#| label: sex-and-smoke-counts
pulmonary |>
  count(sex_factor)
pulmonary |>
  count(smoke_factor)
```

#### Interpretation of the output

Slightly more than half of the patients are male. Almost all patients are non-smokers.

## Draw boxplot of fev by smoke_factor

```{r pulmonary-boxplot}
#| fig.height: 1.5
#| fig.width: 6
pulmonary |>
  ggplot() +
    aes(x=smoke_factor, y=fev) +
    geom_boxplot() +
    xlab(" ") +
    ylab("Forced Expiratory Volume (liters)") +
    coord_flip() +
    ggtitle("Plot created by Steve Simon on 2024-12-24")
```      

#### Comments on the code

The base R package has a nice function, [boxplot][ref02], that will draw 
boxplots. This function produces nice boxplots, but I prefer to use the 
[geom_boxplot][ref03] function in R.

Normally, I prefer to put descriptive labels on the plot axes, but the values (smoker, non-smoker) on the x-axis are self-explanatory.

The [coord_flip][ref04] function produces horizontal boxplots, which I generally  prefer.

[ref02]: https://stat.ethz.ch/R-manual/R-devel/library/graphics/html/boxplot.html
[ref03]: https://ggplot2.tidyverse.org/reference/geom_boxplot.html
[ref04]: https://ggplot2.tidyverse.org/reference/coord_flip.html

## Draw boxplot of fev by smoke_factor

```{r pulmonary-boxplot}
#| fig.height: 1.5
#| fig.width: 6
pulmonary |>
  ggplot() +
    aes(x=smoke_factor, y=fev) +
    geom_boxplot() +
    xlab(" ") +
    ylab("Forced Expiratory Volume (liters)") +
    coord_flip() +
    ggtitle("Plot created by Steve Simon on 2024-12-24")
```      

#### Interpretation of the output

The fev values tend to be larger for the smokers than the non-smokers.

```{r age-boxplot}
pulmonary |>
  ggplot() +
    aes(x=smoke_factor, y=age) +
    geom_boxplot() + 
    ggtitle("Plot created by Steve Simon on 2024-12-24")
```      
      
### Calculate group means and standard deviations

```{r age-by-smoke}
pulmonary |>
  group_by(smoke_factor) |>
  summarize(
    age_mean=mean(age),
    age_sd=sd(age),
    age_min=min(age),
    age_max=max(age)) |>
  round(digits=1)
```
