---
title: "08-02, The t-distribution"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

## The t distribution

-   $T=\frac{\bar{X}_1-\bar{X}_2}{se}$
    -   Variation in the numerator AND the denominator
    -   Use a t-distribution, not a normal distribution
        -   $df=n_1+n_2-2$
    
:::notes
The test statistic, T, is the ratio of the difference in sample means to the standard error. This statistic has variation both in the numerator and the denominator. This produces a statistic that is not normally distributed, but close to normal. It is the t-distribution.

The t-distribution has degrees of freedom. A large degrees of freedom means very little sampling error in the denominator. It is the total sample size ($n_1+n_2$) minus two degrees of freedom associated with the two estimated means used in the standard deviation calculation.
:::

## Comparing the t and normal distributions

```{r compare-t-and-z}
#| message: false
#| warning: false
library(tidyverse)

x <- seq(-4, 4, length=100)
y1 <- dnorm(x)
y2 <- dt(x, 10)
y3 <- dt(x, 4)

data.frame(x, y1, y2, y3) |>
  ggplot(aes(x, y1)) + 
    geom_line(
      color="black",
      size=1) +
    geom_line(
      aes(y=y2), 
      color="darkgreen",
      size=1,
      linetype="51") +
    geom_line(
      aes(y=y3), 
      color="darkred",
      size=1,
      linetype="11") +
    geom_text(
      x=1.5, y=0.38, hjust=0, 
      label=
        "Black solid line is the normal distribution") +
    geom_text(
      x=1.5, y=0.36, hjust=0,
      color="darkgreen",
      label=
        "Dark green dashed line is the t with df=10") +
    geom_text(
      x=1.5, y=0.34, hjust=0,  
      color="darkred",
      label=
        "Dark red dotted line is the t with df=4") +
    ggtitle("Graph drawn by Steve Simon on 2024-10-06")
```

:::notes
This graph compares the normal distribution to a t distribution with 10 degrees of freedom and a t distribution with 4 degrees of freedom. Both the normal and the t-distributions are symmetric. The t-distributions have a little bit less probability near zero and a bit more probability at the extremes.
:::

```{r setup}
#| message: false
#| warning: false
library(tidyverse)
```

```{r t-with-4-df}
x <- seq(-4, 4, length=100)
y <- y3
data.frame(x, y) |>
  ggplot(aes(x, y)) + 
    geom_polygon(fill="white", color="black") -> t_4_curve
```

## P[t(4) < 1.5]

Use geom_vline and geom_label to draw a vertical reference line and add text to the normal curve. The qt function computes the t probability.

```{r prob-1}
a <- 1.5
t_4_curve +
  geom_vline(xintercept=a) +
  geom_label(x=a, y=0.4, label=a) +
  geom_label(x=a-0.5, y=0, label="Area = ?")
pt(1.5, 4)
```

The probability that a standard normal variable is less than 1.5 is about 93%.

## P[Z < -0.5]

```{r prob-2}
a <- -0.5
t_4_curve +
  geom_vline(xintercept=a) +
  geom_label(x=a, y=0.4, label=a) +
  geom_label(x=a-0.5, y=0, label="Area = ?")
pt(-0.5, 4)
```

The probability that a standard normal variable is less than -0.5 is about 31%.

## P[Z > 1]

When you are calculating the probability on the right (probability greater than some number), use 1-pnorm.

```{r prob-3}
a <- 1
t_4_curve +
  geom_vline(xintercept=a) +
  geom_label(x=a, y=0.4, label=a) +
  geom_label(x=a+0.5, y=0, label="Area = ?")
1- pt(1, 4)
```

The probability that a standard normal variable is larger than 1 is about 16%.

## P[Z > -2]

```{r prob-4}
a <- -2
t_4_curve +
  geom_vline(xintercept=a) +
  geom_label(x=a, y=0.4, label=a) +
  geom_label(x=a+0.5, y=0, label="Area = ?")
1- pt(-2, 4)
```

The probability that a standard normal variable is larger than -2 is about 98%.

## P[-2.5 < Z < 2.5]

When you are calculating the probability between two values, compute pnorm of the larger value minus pnorm of the smaller value.

```{r prob-5}
a <- 2.5
t_4_curve +
  geom_vline(xintercept=-a) +
  geom_vline(xintercept= a) +
  geom_label(x=-a, y=0.4, label=-a) +
  geom_label(x= a, y=0.4, label= a) +
  geom_label(x=0, y=0, label="Area = ?")
pt(2.5, 4)- pt(-2.5, 4)
```

## P[-0.5 < Z < 0.5]

```{r prob-6}
a <- 0.5
t_4_curve +
  geom_vline(xintercept=-a) +
  geom_vline(xintercept= a) +
  geom_label(x=-a, y=0.4, label=-a) +
  geom_label(x= a, y=0.4, label= a) +
  geom_label(x=0, y=0, label="Area = ?")
pt(0.5, 4) - pt(-0.5, 4)
```

The probability that a standard normal variable is between -0.5 and 0.5 is about 38%.

## 25th percentile of a standard normal

Use qnorm to calculate quantiles of the standard normal distribution.

```{r quantile-1}
p <- 0.25
a <- qt(p, 4)
t_4_curve +
  geom_vline(xintercept=a) +
  geom_label(x=a, y=0.4, label="Quantile = ?") +
  geom_label(x=a-0.5, y=0, label=p)
qt(0.25, 4)
```

The 25th percentile of a standard normal distribution is -0.67.

## 90th percentile of a standard normal

```{r quantile-2}
p <- 0.9
a <- qnorm(p, 4)
t_4_curve +
  geom_vline(xintercept=a) +
  geom_label(x=a, y=0.4, label="Quantile = ?") +
  geom_label(x=a-0.5, y=0, label=p)
qt(0.9, 4)
```

The 90th percentile of a standard normal distribution is 1.28.
