---
title: "08-03, Critical values and p-values"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

## Type I and Type II errors

-   Type I error, Rejecting the null hypothesis when the null hypothesis is true
    -   $\alpha$ is the probability of a Type I error
-   Type II error, Accepting the null hypothesis when the null hypothesis is false
    -   $\beta$ is the probability of a Type II error
    -   Power = 1 - $\beta$

## Critical values, 1

-   $H_0:\ \mu_1 - \mu_2 = 0$
-   $H_1:\ \mu_1 - \mu_2 \ne 0$
    -   $T=\frac{\bar{X}_1-\bar{X}_2}{se}$
    -   Accept $H_0$ if $t(\alpha/2, df) < T < t(1-\alpha/2, df)$

:::notes
The formal test of hypothesis looks at whether the test statistic T is close to zero. Close means that it falls between the $\alpha/2$ and $1-\alpha/2$ percentiles of a t distribution with n~1~ + n~2~ -2 degrees of freedom.
:::
  
## Critical values, 2

```{r critical-values-2}
#| message: false
#| warning: false
library(tidyverse)
load("../data/module08.RData")
x <- seq(-6, 6, length=100)
y <- dt(x, 4)

data.frame(x, y) |>
  ggplot(aes(x, y)) + 
    geom_line(
      color="black",
      size=1) +
    expand_limits(y=0.4) +
    ggtitle("Graph drawn by Steve Simon on 2024-10-06") -> t_curve
a <- qt(0.98, 4)
eq0 <- expression(paste("Area = 1 - ", alpha))
eq1 <- expression(paste("t(", alpha, "/2, df)"))
eq2 <- expression(paste("t(1-", alpha, "/2, df)"))
eq3 <- expression(paste("Area = ", alpha, "/2"))
t_curve +
  geom_segment(x=-a, y=0, xend=-a, yend=0.38) +
  geom_segment(x= a, y=0, xend= a, yend=0.38) +
  geom_label(x=0, y=0, label=eq0) +
  geom_label(x=-a, y=0.4, label=eq1) +
  geom_label(x= a, y=0.4, label=eq2) +
  geom_label(x=-a-0.9, y=0, label=eq3) +
  geom_label(x= a+0.9, y=0, label=eq3)
```

## Critical values, 3

```{r critical-values-3}
a <- qt(0.95, 4)
eq0 <- "Area = 0.90"
eq1 <- "t(0.05, df)"
eq2 <- "t(0.95, df)"
eq3 <- "Area = 0.05"
t_curve +
  geom_segment(x=-a, y=0, xend=-a, yend=0.38) +
  geom_segment(x= a, y=0, xend= a, yend=0.38) +
  geom_label(x=0, y=0, label=eq0) +
  geom_label(x=-a, y=0.4, label=eq1) +
  geom_label(x= a, y=0.4, label=eq2) +
  geom_label(x=-a-0.9, y=0, label=eq3) +
  geom_label(x= a+0.9, y=0, label=eq3)
```

## Critical values, 4

```{r critical-values-4}
a <- qt(0.995, 4)
eq0 <- "Area = 0.99"
eq1 <- "t(0.005, df)"
eq2 <- "t(0.995, df)"
eq3 <- "Area = 0.005"
t_curve +
  geom_segment(x=-a, y=0, xend=-a, yend=0.38) +
  geom_segment(x= a, y=0, xend= a, yend=0.38) +
  geom_label(x=0, y=0, label=eq0) +
  geom_label(x=-a, y=0.4, label=eq1) +
  geom_label(x= a, y=0.4, label=eq2) +
  geom_label(x=-a-0.9, y=0, label=eq3) +
  geom_label(x= a+0.9, y=0, label=eq3)
```

## Critical values, 5

```{r critical-values-5}
a <- qt(0.975, 4)
eq0 <- "Area = 0.95"
eq1 <- "t(0.025, df)"
eq2 <- "t(0.975, df)"
eq3 <- "Area = 0.025"
t_curve +
  geom_segment(x=-a, y=0, xend=-a, yend=0.38) +
  geom_segment(x= a, y=0, xend= a, yend=0.38) +
  geom_label(x=0, y=0, label=eq0) +
  geom_label(x=-a, y=0.4, label=eq1) +
  geom_label(x= a, y=0.4, label=eq2) +
  geom_label(x=-a-0.9, y=0, label=eq3) +
  geom_label(x= a+0.9, y=0, label=eq3)
```

## P-value

-   p-value = $2 P[t(n_1+n_2-2) > |T|]$
    -   Probability of sample results or results more extreme
    -   Accept $H_0$ if p-value > $\alpha$
-   Why the 2?
    -   Measuring extremity in either direction
    
## Postural sway data, 1

```{r sway-summary}
sway |>
  group_by(age) |>
  summarize(
    fb_mn=mean(fbsway),
    fb_sd=sd(fbsway),
    n=n())
m1 <- t.test(
  fbsway ~ age, 
  data=sway,
  alternative="two.sided",
  var.equal=TRUE)
```

## Postural sway data, 2

```{r sp}
sp <- sqrt((9*9.77^2+8*4.09^2)/(9+8))
```

-   Calculate pooled standard deviation
    -   $S_p$ = $\sqrt{\frac{n_1\ S_1^2\ +\ n_2\ S_2^2}{n_1\ +\ n_2}}$
    -   $S_p$ =$\sqrt{\frac{9\ (9.77)^2\ +\ 8\ (4.09)^2}{9\ +\ 8}}$
    -   $S_p$ = `r round(sp, 2)`


## Postural sway data, 3

```{r se}
se <- sp*sqrt(1/9+1/8)
```

-   Calculate standard error
    -   se = $S_p \sqrt{\frac{1}{n_1}+\frac{1}{n_2}}$
    -   se = $7.64 \sqrt{\frac{1}{9}+\frac{1}{8}}$
    -   se = `r round(se, 2)`

## Postural sway data, 4

```{r t}
t <- (26.3-18.1)/3.72
```

-   Calculate T
    -   T = $\frac{\bar{X}_1-\bar{X}_2}{se}$
    -   T = $\frac{26.3-18.1}{3.72}$
    -   T = `r round(t, 2)`

## Postural sway data, 5

-   Calculate critical values
    -    t(0.025, 15) = qt(0.025, 15) = `r round(qt(0.025, 15), 2)`
    -    t(0.975, 15) = qt(0.975, 15) = `r round(qt(0.975, 15), 2)`    
    -    Since T is outside the two critical values, reject $H_0$
    
## Postural sway data, 6

```{r p-v}
p_value <- 2*(1-pt(2.2, 15))
```

-   Calculate p-value
    -    p.value = $P[t(n_1+n_2-2) > |T|]$
    -    p.value = $2\ P[t(15) > |2.2|]$ = 2*(1-pt(2.2, 15)) = `r round(p_value, 3)`
    
## Postural sway data, 7

```{r t-test-sway}
m1 <- t.test(
  fbsway ~ age, 
  data=sway,
  alternative="two.sided",
  var.equal=TRUE)
m1
```