---
title: "Sample size justification for postural sway study"
format: 
  html:
    embed-resources: true
editor: source
---

This program evaluates various sample size calculations for a proposed replication of the postural-sway study in a different population. It was written by Steve Simon on 2024-10-08 and is placed in the public domain.

## Scenario 1

-   Replicate postural sway study
    -   Different populations
    -   Same outcome measure
-   Research hypothesis, $H_0\ \mu_1 - \mu_2 = 0$
-   Standard deviations: 9.77, 4.09
-   MCID = 4

```{r power-1}
power.t.test(
  n=NULL,
  delta=4,
  sd=9.8,
  sig.level=0.05,
  power=0.9,
  type="two.sample",
  alternative="two.sided")
```

With a sample of 128 patients, we would have 90% power for detecting a 4 unit difference in postural sway, using a two-sided test at an alpha level of 0.05.

## Scenario 2, MCID = 2

```{r power-2}
power.t.test(
  n=NULL,
  delta=2,
  sd=9.8,
  sig.level=0.05,
  power=0.9,
  type="two.sample",
  alternative="two.sided")
```

If we wanted to be able to detect a 2 unit difference, we would need a sample size that is about four times as large.
    
## Scenario 3, MCID=8

## Scenario 4, sd=4.9

## Scenario 5, sd=19.6

## Scenario 6, alpha=0.01

## Scenario 7, alpha=0.10

## Scenario 8, power=0.8

## Scenario 9, power=0.95
