---
title: "Analysis of postural sway data"
author: "Steve Simon"
format: 
  html:
    embed-resources: true
date: 2024-10-07
---

This program reads data and runs a two-sample t-test. Consult the [data dictionary][dic] for information about the data itself.

[dic]: https://github.com/pmean/data/blob/main/files/postural-sway.yaml

## Libraries

```{r setup}
#| message: false
#| warning: false
library(broom)
library(tidyverse)
```

## Read data

```{r read-memory}
sway <- read_tsv(
  file="../data/postural-sway.txt",
  col_types="cnn")
names(sway) <- tolower(names(sway))
glimpse(sway)
```

## Boxplot of fbsway by age

```{r}
#| fig.height: 1.5
sway |>
  ggplot(aes(age, fbsway)) +
    geom_boxplot() +
    ggtitle("Graph drawn by Steve Simon on 2024-10-05") +
    xlab("Treatment group") +
    ylab("Front to back sway") +
    coord_flip()
```

The elderly patients have generally higher sway values. Both groups have similar variation except for an extreme outlier in the elderly patients.

## Descriptive statistics for memory by treatment

```{r group-means}
sway |>
  group_by(age) |>
  summarize(
    fb_mn=mean(fbsway),
    fb_sd=sd(fbsway),
    n=n())
```

The average front-to-back sway is higher in the elderly patients. There is more variation in the elderly group, possibly caused by the extreme outlier.

## Two-sample t-test

```{r t-test}
m1 <- t.test(
  fbsway ~ age, 
  data=sway,
  alternative="two.sided",
  var.equal=TRUE)
m1
```

## Equivalent analysis using linear regression

```{r lm}
m2 <- lm(fbsway ~ age, data=sway)
tidy(m2)
confint(m2)
```