---
title: "Interactions in multi-factor analysis of variance"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
---

## Hypothetical applications

```{r}
#| label: 05-02-setup
#| message: false
#| warning: false

library(broom)
library(glue)
library(tidyverse)
load("../data/module05.RData")
```

-   Study with two drugs and two placebos
-   Four treatment options
    -   Placebo plus placebo
    -   Drug A plus placebo
    -   Placebo plus Drug B
    -   Drug A plus Drug B
    
## Listing of hypothetical data

```{r}
#| label: 05-02-listing

hyp <- tibble(
  a=c("P", "P", "A", "A", "P", "P", "A", "A"),
  b=c("P", "P", "P", "P", "B", "B", "B", "B"),
  y=c(1      -0.1,
      1      +0.1, 
      1+2    -0.1,
      1+2    +0.1, 
      1  +3  -0.1,
      1  +3  +0.1,
      1+2+3+4-0.1,
      1+2+3+4+0.1))
hyp$a <- factor(hyp$a, levels=c("P", "A"))
hyp$b <- factor(hyp$b, levels=c("P", "B"))
hyp
```

## Line plot of hypothetical data, 1

```{r}
#| label: 05-02-line-plot-1

hyp |>
  group_by(a, b) |>
  summarize(mean_y=mean(y), .groups="drop") |>
  mutate(lab=glue("{a} {b}")) |>
  ggplot() +
  aes(x=a, y=mean_y) +
  expand_limits(y=c(0, 12)) +
  geom_line(aes(group=b)) +
  scale_y_continuous(breaks=(0:6)*2) +
  ggtitle("Graph drawn by Steve Simon on 2025-02-18") +
  xlab("Drug A") +
  ylab("Hypothetical mean") -> p1
p1
```

## Create indicator variables

```{r}
#| label: 05-02-indicators

hyp |>
  mutate(i_a=as.numeric(a=="A")) |>
  mutate(i_b=as.numeric(b=="B")) |>
  mutate(i_ab=i_a*i_b) |>
  select(a, b, i_a, i_b, i_ab, y) -> hyp_1
hyp_1
```

## Interaction model, 1

-   `lm(y=a+b+a:b, data=hyp)`
-   `lm(y=a*b, data=hyp)`
-   `lm(y=i_a+i_b+i_ab, data=hyp_1)`

```{r}
#| label: 05-02-model

m1 <- lm(y~i_a+i_b+i_ab, data=hyp_1)
m1 |>
  tidy() |>
  select(term, estimate)
```

## Interaction model, 2

-   (Intercept): estimated average effort of two placebos
-   i_a: estimated average change due to drug A alone
-   i_b: estimated average change due to drug B alone
-   i_ab: estimated synergistic effect of both drugs

::: notes

The intercept is the estimated average value of Y when all of the indicator variables are equal to zero. This is the average in the P P group. The slope associated with i_a is the estimated average change when moving from placebo to drug A, holding drug B constant. The slope associated with i_b is the estimated average change when moving from placebo to drug B, holding drug A constant. The slope associated with i_ab is the interaction, a measure of how much more the effect of the two drugs combined compared to the effect of drug A along plus drug B alone.

:::

## Interaction model, 3

-   `P P mean = 1`
-   `A P mean = 1 + 2`
-   `P B mean = 1     + 3`
-   `A B mean = 1 + 2 + 3 + 4`

## Line plot of hypothetical data, 2

```{r}
#| label: 05-02-line-plot-2

p1 +
  geom_segment(x="P", y=1, xend="A", yend=1, linetype=2) +
  geom_segment(x="P", y=4, xend="A", yend=6, linetype=2) +
  geom_segment(x="P", y=0, xend="P", yend=1) +
  geom_segment(x="A", y=1, xend="A", yend=3) +
  geom_segment(x="P", y=1, xend="P", yend=4) +
  geom_segment(x="A", y=3, xend="A", yend=10) +
  geom_label(x="P", y=0.5, label="b0=1") +
  geom_label(x="A", y=2,   label="b1=2") +
  geom_label(x="P", y=2.5, label="b2=3") +
  geom_label(x="A", y=4.5, label="b2=3") +
  geom_label(x="A", y=8,   label="b3=4")
```

## Line plot of hypothetical data, 3

```{r}
#| label: 05-02-line-plot-3

p1 +
  geom_segment(x="P", y=1, xend="A", yend=1, linetype=2) +
  geom_segment(x="P", y=4, xend="A", yend=6, linetype=2) +
  geom_segment(x="P", y=0, xend="P", yend=1) +
  geom_segment(x="A", y=1, xend="A", yend=3) +
  geom_segment(x="P", y=1, xend="P", yend=4) +
  geom_segment(x="A", y=3, xend="A", yend=10) +
  geom_label(x="P", y=1, label="b0=1") +
  geom_label(x="A", y=3, label="b0+b1=1+2") +
  geom_label(x="P", y=4, label="b0+b2=1+3") +
  geom_label(x="A", y=6, label="b0+b1+b2=1+2+3") +
  geom_label(x="A", y=10,label="b0+b1+b2+b3=1+2+3+4")
```

## Study of three exercise programs

-   Swimming
-   Jogging
-   Reading (control)

## Descriptive statistics for exercise program

```{r}
#| label: 05-02-prog

sv$prog_pct
```

## Descriptive statistics for hours

```{r}
#| label: 05-02-hours

sv$hours_mean
```

## Descriptive statistics for gender

```{r}
#| label: 05-02-gender

sv$gender_pct
```

## Descriptive statistics for effort 

-   0 = no effort, 50=maximum effort

```{r}
#| label: 05-02-effort

sv$effort_mean
```

## Descriptive statistics for loss

-   A negative value is a weight gain

```{r}
#| label: 05-02-loss

sv$loss_mean
```

## Line plot, weight loss by program and gender

```{r}
#| label: 05-02-line

sv$loss_lines_2
```

## Interaction model

-   Reference categories: male and reading

```{r}
#| label: 05-02-interaction

sv$m2_betas
```
