---
title: "Directions for 5502-01 programming assignment"
format: 
  html:
    embed-resources: true
execute:
  error: true
---

## File details

This solution to the programming assignment for module01 of MEDB 5502 was written by Steve Simon on 2025-01-27 and is placed in the public domain.

This assignment is for practice only and there is no grading.

The data, [samara-velocity.txt][ref01] file is available on my github site, along with a [data dictionary][ref02].

[ref01]: https://github.com/pmean/data/blob/main/files/samara-velocity.txt
[ref02]: https://github.com/pmean/data/blob/main/files/samara-velocity.yaml

```{r}
#| label: setup
#| message: false
#| warning: false

library(glue)
library(tidyverse)
R.version.string
Sys.Date()
```

## Read the data

```{r}
#| label: read

sam <- read_tsv(
  file="../data/samara-velocity.txt",
  col_names=TRUE,
  col_types="n")
glimpse(sam)
```
    

## Question 1

Calculate descriptive statistics for the three variables in the dataset. 
Interpret these statistics. You do not need to produce any graphs.

```{r}
#| label: descriptives-tree

sam |>
  count(tree) |>
  mutate(total=sum(n)) |>
  mutate(pct=round(100*n/total)) |>
  glue(pct="{pct}% ({n}/{total}") |>
  select(-n, -total)
```

#### Interpretation of the output

```{r}
#| label: descriptives-load

sam|>
  summarize(
    load_mean=mean(load),
    load_sd=sd(load),
    load_min=min(load),
    load_max=max(load))
```

#### Interpretation of the output

```{r}
#| label: descriptives-load

sam|>
  summarize(
    load_mean=mean(load),
    load_sd=sd(load),
    load_min=min(load),
    load_max=max(load))
```

#### Interpretation of the output

## Question 2

Calculate a linear regression model using Load as the independent variable and
Velocity as the dependent variable. Interpret these results.

## Question 3

Calculate indicator variables for Tree=2 and Tree=3. Include these as 
independent variables in a linear regression model predicting Velocity.
Interpret this model.

## Your submission

-   Save the output in html format
-   Convert it to pdf format.
-   Make sure that the pdf file includes
    -   Your last name
    -   The number of this course
    -   The number of this module
-   Upload the file
-   Please note the [policy on late submissions and rework][sim3].

[sim3]: https://github.com/pmean/classes/blob/master/general/policy-on-extensions-and-rework.md

## If it doesn't work

Please review the [suggestions if you encounter an error page][sim4].

[sim4]: https://github.com/pmean/classes/blob/master/general/suggestions-if-you-encounter-an-error.md
