---
title: "A simple example of survival data"
format: 
  revealjs:
    slide-number: true
    embed-resources: true
editor: source
execute: 
  echo: false
  eval: true
---

## Survival analysis

-   Time to event models
    -   Death
    -   Relapse
    -   Rehospitalization
    -   Failure of medical device
    -   Pregnancy
-   Not every patient experiences the event
    -   These are censored observations

::: notes
Survival analysis models are more properly called time to event models. You follow a group of patients from a certain time point and note the amount of time until they die. 


Or the amount of time until they relapse. Or the amount of time until they need to be rehospitalized. Or the amount time until a medical device that you implanted in them fails.

I should note that while almost all of the events in a time to event model are bad, there are a few exceptions. In a study of couples with fertility problems, you might use a time to event model to study the time to pregnancy, a very happy event for any couple with fertility problems.

Mortality is the context under which time to event models were derived, so the term survival analysis has been used even when the event is different.

A key feature of survival analysis is that not every patient experiences the event. You should be glad that not everyone that you recruit for a clinical trial dies, but this adds a layer of complexity to the analysis.

The reasons for not experiencing the event can vary. Everyone dies, but maybe the event is death from cancer and if you patient gets hit by a bus, that patient does not experience the event. If the event is death from any cause, you still have to end the study in some time frame, and not everyone will die in that time frame. Not every patient gets rehospitalized, not every patient relapses, not every medical device fails.

A patient may drop out of a study, and you no longer are able to tell from that point onward whether that patient would have experienced the event sometime during the rest of your study.

When your patient does not die during the study, this is not a missing value. You have partial information. You know that the patient was alive for a certain amount of time. When you end the study within a certain time frame, you know that your patient dies at a time beyond the end date of your study. If your patient drops out after six months, you know that the patient survived for more than six months.
:::

## First fruit fly experiment, 1

```{r}
#| message: false
#| warning: false
#| label: 08-01-setup

library(ggfortify)
library(glue)
library(readr)
library(survival)
library(tidyverse)
library(yaml)
```

```{r}
#| label: 08-01-yaml

f0 <- "https://raw.githubusercontent.com/pmean"
f1 <- "data/refs/heads/main/files"
dd <- read_yaml(glue("{f0}/{f1}/fly1.yaml"))
cat(as.yaml(dd["data_dictionary"]))
cat(as.yaml(dd["description"]))
cat(as.yaml(dd["vars"]))
```

::: notes

The following data represents survival time for a group of fruit flies and is a subset of a larger data set found at the Data and Story Library (DASL). The data set has been slightly modified to simplify some of these explanations.

There are 25 flies in the sample, with the first fly dying on day 37 and the last fly dying on day 96.
:::

## First fruit fly experiment, 2

```{r}
#| label: 08-01-fly1a

fly1 <- read_csv(
  "../data/fly1.txt",
  col_names=FALSE,
  col_types="n")
cat(paste(unlist(fly1), collapse=", "))
```

::: notes


If you wanted to estimate the survival probability for this data, you would draw a curve that decreases by 4% (1/25) every time a fly dies.

:::

## First fruit fly experiment, 3

:::: {.columns}

::: {.column width=33%}

```{r}
#| label: 08-01-fly1b

names(fly1) <- "day"
fly1$p <- paste0(100-4*(1:25), "%")
fly1 <- data.frame(fly1)
fly1[1:9, ]
```

:::

::: {.column width=33%}

```{r}
#| label: 08-01-fly1c

fly1[10:18, ]
```

:::

::: {.column width=34%}

```{r}
#| label: 08-01-fly1d

fly1[19:25, ]
```

:::

::::

::: notes
The probability of survival drops by 4% (1/25) at each day of death.
:::

## First fruit fly experiment, 4


```{r}
#| label: fly-plot-1

f1 <- survfit(
  Surv(fly1$day, rep(1, 25)) ~ 1,
  data=fly1)

autoplot(f1, conf.int=FALSE) +
  ggtitle("Graph drawn by Steve Simon on 2025-03-10") +
  xlab("Survival time (days)") +
  ylab("Percentage survived")
```

::: notes
Here's a graph of these probabilities over time.

By tradition and for some rather technical reasons, you should use a stair step pattern rather than a diagonal line to connect adjacent survival probabilities.
:::

## Second fruit fly experiment, 1

```{r}
#| label: 08-01-fly2a

day2 <- fly1$day
day2[16:25] <- "??"
cat(paste(day2, collapse=", "))
```

::: notes
Now let's alter the experiment. Suppose that totally by accident, a technician leaves the screen cover open on day 70 and all the flies escape. This includes the fly who was going to die on the afternoon of the 70th day anyway. Oh the sadness of it all; the poor fly has the briefest of tastes of freedom then ends up shriveled up on a window sill.

You're probably worried that the whole experiment has been ruined. But don't be so pessimistic. You still have complete information on survival of the fruit flies up to their 70th day of life. 

:::

## Second fruit fly experiment, 2

:::: {.columns}

::: {.column width=33%}

```{r}
#| label: 08-01-fly2b

fly2 <- data.frame(day=fly1$day)
fly2$day[16:25] <- 70
fly2$event <- rep(1:0, c(15, 10))
fly2[1:9, ]
```

:::

::: {.column width=33%}

```{r}
#| label: 08-01-fly2c

fly2[10:18, ]
```

:::

::: {.column width=34%}

```{r}
#| label: 08-01-fly2d

fly2[19:25, ]
```

:::

::::

::: notes
Here's how you would code the data for importing into SPSS or any other software.
:::

## Second fruit fly experiment, 3

:::: {.columns}

::: {.column width=33%}

```{r}
#| label: 08-01-fly2e

fly2$p <- fly1$p
fly2$p[16:25] <- " "
fly2[1:9, ]
```

:::

::: {.column width=33%}

```{r}
#| label: 08-01-fly2f

fly2[10:18, ]
```

:::

::: {.column width=34%}

```{r}
#| label: 08-01-fly2g

fly2[19:25, ]
```

:::

::::

::: notes
We clearly have enough data to make several important statements about survival probability. For example, the median survival time is 61 days because roughly half of the flies had died before this day.
:::

## Second fruit fly experiment, 4

```{r}
#| label: fly-plot-2

f2 <- survfit(
  Surv(fly2$day, event) ~ 1,
  data=fly2)

autoplot(f2, conf.int=FALSE) +
  expand_limits(x=96, y=0) +
  ggtitle("Graph drawn by Steve Simon on 2025-03-10") +
  xlab("Survival time (days)") +
  ylab("Percentage survived")
```

::: notes

Here is a graph of the survival probabilities of the second experiment. This graph is identical to the graph in the first experiment up to day 70 after which you can no longer estimate survival probabilities.

By the way, you might be tempted to ignore the ten flies who escaped. But that would seriously bias your results. All of these flies were tough and hardy flies who lived well beyond the median day of death. If you pretended that they didn't exist, you would seriously underestimate the survival probabilities. The median survival time, for example, of the 15 flies who did not escape, for example, is only 54 days which is much smaller than the actual median.

:::

## Third fruit fly experiment, 1

```{r}
#| label: 08-01-fly3a

fly3 <- read_delim(
	"../data/fly3.txt",
	delim=" ",
	col_names=c("day", "event"),
	col_types="n")
fly3 <- data.frame(fly3)
day3 <- fly3$day
day3[fly3$day==70] <- "??"
cat(paste(day3, collapse=", "))

```

::: notes

Let's look at a third experiment, where the screen cover is left open and all but four of the remaining flies escape. It turns out that those four remaining flies who didn't bug out will allow us to still get reasonable estimates of survival probabilities beyond 70 days. 

:::

## Third fruit fly experiment, 2

:::: {.columns}

::: {.column width=33%}

```{r}
#| label: 08-01-fly3b

fly3[1:9, ]
```

:::

::: {.column width=33%}

```{r}
#| label: 08-01-fly3c

fly3[10:18, ]
```

:::

::: {.column width=34%}

```{r}
#| label: 08-01-fly3d

fly3[19:25, ]
```

:::

::::

::: notes
Here is how you would code the data for importing into SPSS.
:::

## Third fruit fly experiment, 3

:::: {.columns}

::: {.column width=33%}

```{r}
#| label: 08-01-fly3e

fly3$p <- fly1$p
fly3$p[fly3$day==70] <- " "
fly3$p[fly3$day > 70] <- paste0(10*c(3:0), "%")
fly3[1:9, ]
```

:::

::: {.column width=33%}

```{r}
#| label: 08-01-fly3f

fly3[10:18, ]
```

:::

::: {.column width=34%}

```{r}
#| label: 08-01-fly3g

fly3[19:25, ]
```

:::

::::

::: notes

What you do with the six escaped flies is to allocate their survival probabilities equally among the four flies who didn't bug out. This places a great responsibility among each of those four remaining flies since each one is now responsible for 10% of the remaining survival probability, their original 4% plus 6% more which represents a fourth of the 24% survival probability that was lost with the six escaping flies.

Another way of looking at this is that the six flies who escaped influence the denominator of the survival probabilities up to day 70 and then totally drop out of the calculations for any further survival probabilities. Because the denominator has been reduced, the jumps at each remaining death are much larger.

:::

## Third fruit fly experiment, 4

```{r}
#| label: fly-plot-3

f3 <- survfit(
  Surv(fly3$day, event) ~ 1,
  data=fly3)

autoplot(f3, conf.int=FALSE) +
  expand_limits(x=96, y=0) +
  ggtitle("Graph drawn by Steve Simon on 2025-03-10") +
  xlab("Survival time (days)") +
  ylab("Percentage survived") -> fly_plot_3

fly_plot_3
```


::: notes
Here is a graph of the survival probability estimates from the third experiment.

These survival probabilities differ only slightly from the survival probabilities in the original experiment. This works out because the mechanism that caused us to lose information on six of the fruit flies was independent of their ultimate survival.

If the censoring mechanism were somehow related to survival prognosis, then you would have the possibility of serious bias in your estimates. Suppose for example, that only the toughest of flies (those with the most days left in their short lives) would have been able to escape. The flies destined to die on days 70, 71, 72, and 73, were already on their deathbeds and unable to fly at all, much les make a difficult escape. Then these censored values would not be randomly interspersed among the remaining survival times, but would constitute some of the larger values. But since these larger values would remain unobserved, you would underestimate survival probabilities beyond the 70th day.

This is known as informative censoring, and it happens more often that you might expect. Suppose someone drops out of a cancer mortality study because they are abandoning the drugs being studied in favor of laetrile treatments down in Mexico. Usually, this is a sign that the current drugs are not working well, so a censored observation here might represent a patient with a poorer prognosis. Excluding these patients would lead to an overestimate of survival probabilities.
:::

## Interpreting Kaplan-Meier plots, 1

```{r}
#| label: fly-plot-4

fly4 <- fly3

fly4$event[seq(1, 23, 3)] <- 0
fly4$event[seq(2, 24, 3)] <- 0

f4 <- survfit(
  Surv(day, event) ~ 1,
  data=fly4)

autoplot(f4, conf.int=FALSE) +
  expand_limits(x=96, y=0) + 
  geom_label(x=95, y=0.95, label="Happy\ncorner") +
  ggtitle("Graph drawn by Steve Simon on 2025-03-10") +
  xlab("Survival time (days)") +
  ylab("Percentage survived")
```

## Interpreting Kaplan-Meier plots, 2

```{r}
#| label: fly-plot-5

fly3 |>
  mutate(day=(day %%21)) -> fly5 

fly5$day[c(5, 15, 25)] <- fly3$day[c(5, 15, 25)]

f5 <- survfit(
  Surv(day, event) ~ 1,
  data=fly5)

autoplot(f5, conf.int=FALSE) +
  expand_limits(x=96, y=0) + 
  geom_label(x=5, y=0.05, label="Sad\ncorner") +
  ggtitle("Graph drawn by Steve Simon on 2025-03-10") +
  xlab("Survival time (days)") +
  ylab("Percentage survived")
```

::: notes

When you see a survival curve in a research paper, there are three ways to interpret it.

First, presuming that the event in question is a sad event (such as death, relapse), then the upper right hand corner is the happy corner. Most of you patients go for a very long time with only a small proportion suffering the negative event.

In contrast, the lower left corner is the sad corner. Most of your patients experience the bad event, and they experience it very quickly.

:::

## Interpreting Kaplan-Meier plots, 3

```{r}
#| label: fly-plot-6

fly_plot_3 +
  geom_segment(
    x=0, 
    y=0.5, 
    xend=61, 
    yend=0.5, 
    color="red",
    linewidth=1.5) +
  geom_segment(
    x=61, 
    y=0.5, 
    xend=61, 
    yend=0, 
    color="red",
    linewidth=1.5) +
  ggtitle("Graph drawn by Steve Simon on 2025-03-10") +
  xlab("Survival time (days)") +
  ylab("Percentage survived")
```

::: notes

Next, you can get an estimate of the median (or other percentiles) by projecting horizontally until you intersect with the survival curve and then head down to get your estimate. In the survival curve we have just looked at, you would estimate the median survival as slightly more than 60 days.

:::

## Interpreting Kaplan-Meier plots, 4

```{r}
#| label: fly-plot-7

fly_plot_3 +
  geom_segment(
    x=80, 
    y=0, 
    xend=80, 
    yend=0.2,
    color="red",
    linewidth=1.5) +
  geom_segment(
    x=80, 
    y=0.2, 
    xend=0, 
    yend=0.2,
    color="red",
    linewidth=1.5) +
  ggtitle("Graph drawn by Steve Simon on 2025-03-10") +
  xlab("Survival time (days)") +
  ylab("Percentage survived")  
```

::: notes

You can also estimate probabilities for survival at any given time by projecting up from the time and then moving to the left to estimate the probability. In the example below, you can see that the 80 day survival probability is a little bit less than 25%.

:::

