---
title: "simon-5502-08-demo"
format: 
  html:
    embed-resources: true
editor: source
execute: 
  error: false
---

## Start part 1

## File details

This program was written by Steve Simon on 2025-02-23 and is placed in the public domain. You can use this program any way you please.

There is one data file used in this program

-   whas500, [Tab delimited file][ref01], [dictionary][ref02]

[ref01]: https://github.com/pmean/data/blob/main/files/whas500.dat
[ref02]: https://github.com/pmean/data/blob/main/files/whas500.yaml

```{r}
#| label: setup
#| message: false
#| warning: false

library(broom)
library(ggfortify)
library(glue)
library(lubridate)
library(survival)
library(tidyverse)

R.version.string
Sys.Date()
```

## Intermediate files

-   whas: Original data from whas500.dat
-   whas_1: Convert dates from strings, drop last column
-   whas_2: Check date calculations
-   whas_3: Remove patients who died in the hospital
-   whas_4: Convert days to years

## Read Worcester Heart Attack Study

```{r}
#| label: whas-read

whas_names <- c(
  "id",
  "age", 
  "gender",
  "hr",
  "sysbp",
  "diasbp",
  "bmi",
  "cvd",
  "afb",
  "sho",
  "chf",
  "av3",
  "miord",
  "mitype",
  "year",
  "admitdate",
  "disdate",
  "fdate",
  "los",
  "dstat",
  "lenfol",
  "fstat",
  "not_needed")

whas <- read_table(
  file="../data/whas500.dat",
  col_types="nnnnnnnnnnnnnnncccnnnnn",
  col_names=whas_names)

glimpse(whas) 
```

## Convert dates

```{r}
#| label: dates

whas |>
  select(-23) |>
  mutate(admitdate=mdy(admitdate)) |>
  mutate(disdate=mdy(disdate)) |>
  mutate(fdate=mdy(fdate)) -> whas_1
```

## Count events

```{r}
#| label: count

whas_1 |>
  count(fstat)
```

#### Interpretation of the output

There are enough events to support a reasonably complex model (215/15 = 14.3).

## Check lenfol calculations

```{r}
#| label: check-1

whas_1 |>
  mutate(f_minus_admit = fdate - admitdate) |>
  mutate(f_minus_dis = fdate - disdate) |>
  mutate(dis_minus_admit = disdate - admitdate) -> whas_2

whas_2 |>
  select(lenfol, f_minus_admit, f_minus_dis, dis_minus_admit)
```

## Check dstat=1

```{r}
#| label: check-2

whas_2 |>
  filter(dstat == 1) |>
  select(lenfol, f_minus_admit, f_minus_dis, dis_minus_admit)
```

## Remove patients who died in the hospital

```{r}
#| label: remove

whas_1 |>
  filter(dstat != 1) -> whas_3
```

## Convert days to years

```{r}
#| label: convert

whas_3 |>
  mutate(lenfol=lenfol/365.25) -> whas_4
```

## Overall Kaplan-Meier curve

```{r}
#| label: overall

m0 <- survfit(
  Surv(lenfol, fstat) ~ 1,
  data=whas_4)

autoplot(m0, conf.int=FALSE) +
  ggtitle("Graph drawn by Steve Simon on 2025-03-10") +
  xlab("Survival time (days)") +
  ylab("Percentage survived")
```

## Comparison of two or more survival curves

```{r}
#| label: gender

m1 <- survfit(
  Surv(lenfol, fstat) ~ gender,
  data=whas_4)

autoplot(m1, conf.int=FALSE) +
  ggtitle("Graph drawn by Steve Simon on 2025-03-10") +
  xlab("Survival time (days)") +
  ylab("Percentage survived")
```

## Log rank test

```{r}
survdiff(
  Surv(lenfol, fstat) ~ gender,
  data=whas_4)
```

## Cox proportional hazards test

```{r}
coxph(
  Surv(lenfol, fstat) ~ gender,
  data=whas_4)
```

## Save everything

```{r}
#| label: save

save.image(file="../data/module08.RData")
```
