---
title: "simon-5502-08-demo"
format: 
  html:
    embed-resources: true
editor: source
execute: 
  error: false
---

# Start part 1 ----------------------------------------------------------------

## File details

This program was written by Steve Simon on 2025-02-23 and is placed in the public domain. You can use this program any way you please.

This is the correct program for the Spring 2026 semester.

There is one data file used in this program

-   whas500, [Tab delimited file][ref-whas-dat], [dictionary][ref-whas-yaml]

[ref-whas-dat]: https://github.com/pmean/data/blob/main/files/whas500.dat
[ref-whas-yaml]: https://github.com/pmean/data/blob/main/files/whas500.yaml

```{r}
#| label: setup
#| message: false
#| warning: false

library(broom)
library(ggfortify)
library(glue)
library(lubridate)
library(survival)
library(tidyverse)

R.version.string
Sys.Date()
```

## Intermediate files

-   whas0: Original data from whas500.dat
-   whas1: Convert dates, add factors, drop last column
-   whas2: Convert days to years
-   whas1: Create age groups
-   whas4: Compute age in decades rather than years

## Read Worcester Heart Attack Study

```{r}
#| label: whas-read

whas_names <- c(
  "id",
  "age", 
  "gender",
  "hr",
  "sysbp",
  "diasbp",
  "bmi",
  "cvd",
  "afb",
  "sho",
  "chf",
  "av3",
  "miord",
  "mitype",
  "year",
  "admitdate",
  "disdate",
  "fdate",
  "los",
  "dstat",
  "lenfol",
  "fstat",
  "not_needed")

whas0 <- read_table(
  file="../data/whas500.dat",
  col_types="nnnnnnnnnnnnnnncccnnnnn",
  col_names=whas_names)

glimpse(whas0) 
```

## Convert dates, add factors

```{r}
#| label: dates

whas0 |>
  select(-not_needed) |>
  mutate(lenfol=lenfol/365.25) -> whas1
```

#### Comments on the code

For long term studies, you will find it easier to interpret results if the follow-up time is measured in years rather than days. The 365.25 in the divisor accounts for leap years. It is not a perfect adjustment, especially for follow-up times less than four years, but it comes close enough.

## Count events

```{r}
#| label: count

whas1 |>
  count(fstat) |>
  mutate(total=sum(n)) |>
  mutate(pct=round(100*n/total)) |>
  mutate(pct=glue("{pct}% ({n}/{total})")) |>
  select(-n, -total) -> whas_fstat_count_table

whas_fstat_count_table
```

#### Interpretation of the output

There are enough events to support a reasonably complex model (215/15 = 14.3).

## Overall Kaplan-Meier curve

```{r}
#| label: overall

whas_null_km_model <- survfit(
  formula=Surv(lenfol, fstat) ~ 1,
  data=whas1)

autoplot(whas_null_km_model, conf.int=TRUE) +
  labs(
    x="Survival time (years)",
    y="Percentage survived",
    caption="Steve Simon, 2025-03-10, CC0") -> whas_null_km_plot

whas_null_km_plot
```

#### Comments on the code

The [Surv function][ref-Surv], part of the survival package, creates a survival object that combines information about the follow-up time combined with a censoring variable. The [survfit function][ref-survfit] creates a Kaplan-Meier curve.

The Kaplan-Meier curve can be easily plotted using the [plot function][ref-plot] in base R, but I prefer to use the ggplot2 functions. They require that you install the [ggfortify library][ref-ggfortify] before using the [autoplot function][ref-autoplot]. The autoplot function peeks inside the object being fed into it and draws a graph appropriate for that object.

The conf.int argument tells R whether to place confidence limits on the survival curve. I generally don't do this because it can clutter up the graph, but other researchers like it. I will use confidence limits if there is only one survival curve.

[ref-autoplot]: https://ggplot2.tidyverse.org/reference/autoplot.html
[ref-ggfortify]: https://cran.r-project.org/web/packages/ggfortify/refman/ggfortify.html
[ref-Surv]: https://stat.ethz.ch/R-manual/R-devel/library/survival/html/Surv.html
[ref-survfit]: https://stat.ethz.ch/R-manual/R-devel/library/survival/html/survfit.html

#### Interpretation of the output

The survival probabilities decline over time, a bit more sharply in the first year than in later years. The censored observations (the plus signs on the graphs) are clustered. An investigation into this would yield a reasonable explanation, but I will leave this unsaid for now.

## Start part 2 ----------------------------------------------------------------

## Descriptive statistics on gender

```{r}
#| label: count-gender

whas1 |>
	count(gender, fstat) -> whas_gender_count_table

whas_gender_count_table
```

#### Comments on the code

Although I normally compute percentages after using the count function, it is not needed here. I just want to make sure that the number of events is reasonably large in both groups.

#### Interpretation of the output

There are a reasonable number of events (fstat=1) among both males (111) and females (104).

## Comparison of two or more survival curves

```{r}
#| label: km-gender-

whas_gender_km_model <- survfit(
  formula=Surv(lenfol, fstat) ~ gender,
  data=whas1)

autoplot(whas_gender_km_model, conf.int=FALSE) +
  labs(
    x="Survival time (years)",
    y="Percentage survived",
    caption="Steve Simon, 2025-03-10, CC0") -> whas_gender_km_plot

whas_gender_km_plot
```

#### Comments on the code

The survift function, described above, allows you to draw seperate curves for any categorical variable. Place that categorical variable on the right hand side of the tilde. With two curves, I decided to suppress the confidence limits with the argument conf.int=FALSE.

#### Interpretation of the output

The females (gender=1) have a slightly worse outcome than males because their curve is closer to the sad lower left corner of the graph. At two years, the survival probability is about 10% lower.

## Log rank test

```{r}
#| label: log-rank

survdiff(
  formula=Surv(lenfol, fstat) ~ gender,
  data=whas1) -> whas_gender_logrank_model

whas_gender_logrank_model
```

#### Comments on the code

The [survdiff function][ref-survdiff] takes a formula for a survival model and runs a logrank test. Other tests are also available for this function.

[ref-survdiff]: https://stat.ethz.ch/R-manual/R-devel/library/survival/html/survdiff.html

#### Interpretation of the output

Males (gender=0) have fewer deaths than you would expect under the null hypothesis while females (gender=1) have more deaths than expected. This difference is statistically significant because the chi-squared statistic is much larger than its degrees of freedom and the p-value is small.

## Cox model for gender

```{r}
coxph(
  formula=Surv(lenfol, fstat) ~ gender,
  data=whas1) -> whas_gender_cox_model

whas_gender_cox_model
```



## Start part 3 ----------------------------------------------------------------

## Histogram for age

```{r}
#| label: age-histogram

whas1 |>
  ggplot() +
  aes(x=age) |>
  geom_histogram(
    color="black",
    fill="white",
    binwidth=1) -> whas_age_histogram_plot

whas_age_histogram_plot
```

#### Interpretation of the output

The data follows a roughly bell-shaped curve. This plot also helps use decide how to combine the data into decades. There are very few patients under the age of 40 or over the age of 90. So it makes sense to combine these with the 50 to 59 year and 80 to 89 year groups, respectively.

## Create age_groups

```{r}
#| label: age-groups

whas1 |>
  mutate(age_group=case_when(
    age < 50 ~ "50 years and under",
    age < 60 ~ "60 to 69 years",
    age < 70 ~ "70 to 79 years",
    age >= 80 ~ "80 years and above")) -> whas2
```

## Draw Kaplan-Meier curves for age groups

```{r}
#| label: km-age

whas_age_group_km_model <- survfit(
  formula=Surv(lenfol, fstat) ~ age_group,
  data=whas2)

autoplot(whas_age_group_km_model, conf.int=FALSE) +
  labs(
    x="Survival time (years)",
    y="Percentage survived",
    caption="Steve Simon, 2026-03-10, CC0") -> km_by_age_group

km_by_age_group
```

#### Comments on the code

You cannot do a good visualization of Kaplan-Meier curves for age. Age is continuous and there are too many values. You can get a rough approximation of the trends with age by creating age groups and seeing if they follow a dose-response pattern.

#### Interpretation of the output

Here the groups are ordered as you would expect, with the youngest patients closest to the happy upper right corner and the oldest patients closest to the sad lower left corner.

One concern is that the spacing between the age groups is about the same for 50 and under, 60 to 69, and 70 to 79. The last group, 80 and above seems to be quite a bit worse. This might be okay, but it might mean that you should consider a model that accelerates for the very old patients. This could be done with a quadratic term or with a spline.

## Cox model for age

```{r}
whas1 |>
  mutate(age_in_decades=age/10) -> whas2

coxph(
  formula=Surv(lenfol, fstat) ~ age_in_decades,
  data=whas2) -> whas_age_cox_model

whas_age_cox_model
```

#### Comments on the code

Many times in these survival models, you get more interesting and interpretable results by changing the units of measure. In this example, I convert age in years to age in decades. A one unit change for the latter variable means a change in risk for a one unit (one decade) change in age.

#### Interpretation of the output

Each additional decade causes an almost doubling of the hazard function.

## Comparison of ages by gender

```{r}
#| label: comparison
#| fig.width: 6
#| fig.height: 1.5

whas2 |>
	group_by(gender) |>
	summarize(age_mean=mean(age)) -> table_mean_age_by_gender

table_mean_age_by_gender

whas2 |>
  ggplot() +
  aes(x=age, y=factor(gender)) +
  geom_boxplot()
```

#### Comments on the code

Perhaps an imbalance in ages between males and females could account for the worse survival results among females. Let's look at some descriptive statistics and a boxplot to see.

#### Interpretation of the output

Both the means and the boxplot indicate that males (gender=0) tend to be younger than females.

## Test for gender, adjusting for age

```{r}
coxph(
  formula=Surv(lenfol, fstat) ~ gender + age_in_decades,
  data=whas2) -> whas_adjusted_cox_model

whas_adjusted_cox_model
```

#### Interpretation of the output

Although the unadjusted model showed a strong effect of gender, when you control for the age gap, the hazard ratio for gender becomes statistically non-significant. This indicates, perhaps, that the better survival probabilities for men can be accounted for by the earlier age that they entered into the study.

## Save everything

```{r}
#| label: save

save.image(file="../data/module08.RData")
```
